---
output: 
    thesisdown::thesis_pdf: default
bibliography: bib/thesis.bib
csl: csl/nature.csl
space_betwee_paragraphs: true
fig_caption: true
always_allow_html: yes
link-citations: true
toc-depth: 2
lot: true
lof: true
indent: true
header-includes: # include other LaTeX packages here
    \usepackage{booktabs}
    \usepackage{longtable}
    \usepackage{siunitx}
    \usepackage[left]{lineno}
    \usepackage{float}
    \floatplacement{figure}{H}
    \linenumbers
    \pagestyle{plain}
    \raggedbottom 
    \usepackage{indentfirst}
editor_options: 
  chunk_output_type: console
---

```{r include=FALSE}
# word count = 
doc.type <- knitr::opts_knit$get('rmarkdown.pandoc.to') # this will tell the .Rmd file what output you are knitting to (word, pdf, html) so that you cna use if else statements when making tables - html/pdf output tables do not work well in word. For word we need to use the 'flextable' packaged to make tables.
## packages
library(dplyr)
library(kableExtra)
library(flextable)
```

# EpiViz: a tool to visualise large association analyses {#visualisation}

<style>
body {
text-align: justify}
</style>

## Chapter summary {-}

The exploration, interpretation, display, and communication of large association analyses is complicated. Previous studies have used Circos plots, however these are cumbersome and through personal experience[@Taylor2019] have been time consuming to produce. This chapter details `EpiViz`, a web application and `R` package that efficiently produces Circos plots. `EpiViz` is used in the following chapters to gain global overview and identify patterns of association in observational and mendelian randomization analyses. `EpiViz` considerably improves the speed and efficiency of producing Circos plots and was used in Bos et al. (2020)[@Bos2021] because of this. \par

#### Contribution statement {-}
Work in this chapter was conducted in collaboration with Osama Mahmoud and Luke A McGuinness. The original project was started by Osama Mahmoud under the supervision of Nicholas Timpson. I worked with Osama to develop the code that would form the original draft of the visualisation tool. I took over the project and remade the base code for the `R` package and `Shiny` application. Luke McGuinness helped with aspects of the `Shiny` application. `EpiViz` is used in Bos et al. (2020)[@Bos2021] for which I am second author, some of the work in this chapter is discussed briefly in that paper. \par

`EpiViz` was greatly aided by the `R` packages [`Circlize`](https://jokergoo.github.io/Circlize_book/book/index.html)[@Gu2014] and [`ComplexHeatmap`](https://jokergoo.github.io/ComplexHeatmap-reference/book/)[@Gu2016], whose code is freely available and was used as a reference point. The following colleagues provided feedback in the development of `EpiViz`: Caroline Bull, Charlie Hatcher, Kurt Taylor, Nancy Mcbride, Neil Goulding, Steph Suddell. \par

\newpage

## Introduction
Large epidemiological analyses, involving potentially hundreds of exposure-outcome associations, pose a significant challenge for the digestion and interpretation of results. This is further complicated by the nature of the analysed data; selecting one of many hundreds of results to follow-up or discuss is difficult if the analyses involved highly correlated variables. This is a particularly important question when considering analyses such as mendelian randomization, where selecting pleiotropic instruments can bias results. \par

The metabolome is one such complex system. The metabolome is made up of many feedback and feed-forward loops, meaning metabolic products of one pathway will likely influence those of another. The intercorrelated nature of metabolites[@Rosato2018] is naturally complicated by the wide degree of genetic correlation[@Gallois2019] and shared genetic architecture[@Shin2014; @Kettunen2016; @Long2017; @Gallois2019]. As such, a perturbation in a single metabolite rarely occurs in isolation. Capturing the overall profile effect of these perturbations may therefore allow for more meaningful interpretation of results. \par

There is evidence that adiposity has a strong effect on the systemic metabolite profile[@Wurtz2014; @Moore2014; @SantosFerreira2017; @Cirulli2019; @Lau2020; @Brachem2020; @Stevens2020; @OKeeffe2020; @Wulaningsih2019; @Frigerio2021; @Neeland2019; @Bachlechner2016; @Rangel-Huerta2019]. These studies show that this effect is broadly to increase the systemic profile. However, breaking the systemic profile down into metabolite classes and sub classes reveals a different picture. For example evidence points to a decreasing effect of adiposity on amino acids but an increasing effect on branched chain and aromatic amino acids[@Wurtz2014]. \par

Metabolites can be profiled in numerous ways, not least through pathways and the use of statistical tools such as principal components. Presenting results of analyses with these profiles is an important means to gain a better understanding of results and of further-work. However, the use of tables with many hundreds to thousands of rows to achieve this is inefficient and makes interpretation more difficult, as patterns of association are not-easily identified. \par

Data visualization aims to improve the exploration, interpretation, communication, and display of data[@Tufte1983]. Forest plots are widely used in epidemiology to achieve this and have been used successfully for metabolomic analyses[@May-Wilson2017; @Taylor2019; @Soares2019; @SantosFerreira2019; @Hartley2020; @Borges2017]. They enable many point estimates and associated elements such as confidence intervals, and p-values to be plotted. Importantly, data can easily be grouped to highlight metabolite profiles. Forestplots can be easily made with a number of `R` packages such as [`ggforestplot`](https://github.com/nightingalehealth/ggforestplot) and [`TwoSampleMR`](https://mrcieu.github.io/TwoSampleMR/#to-many-forest-plot)[@Hemani2018]. However, when dealing with hundreds of variables forest plots can become cumbersome, requiring many separate plots to be created in order to present all of the analyses. This makes comparison and the ability to gain global overview of large analyses difficult. \par

An alternative approach is to compress information into a circular rather than vertical or horizontal form. Circos plots[@Krzywinski2009], as implemented in many genetics studies to condense large genomic information into informative visuals[@Saben2014], provide an efficient visualisation tool for incorporating hundreds to thousands of data points. Circos plots have been used effectively to communicate results from large metabolomic analyses[@Kettunen2012; @Kettunen2016; @Shin2014; @Bos2021; @Taylor2019]. However, Circos software is designed for genomic data and written in programming languages unfamiliar to many epidemiologists.  The `R` package `Circlize`[@Gu2014] provides most of the functionality of the Circos software and is publicly available. However, as a port of the Circos software it is designed for genomic data. As such, and as used in Taylor et al. (2019)[@Taylor2019], it is not easily adapted for epidemiological data. This makes it inaccessible for many epidemiologists. \par

In order to efficiently and reproducibly create Circos plots that enables the exploration, interpretation, communication, and display of large epidemiological analyses as conducted in this thesis, `EpiViz` was developed. `EpiViz` is a web application and `R` package, that builds on the `Circlize`[@Gu2014] and `ComplexHeatmap`[@Gu2016] `R` packages. It enables data to be presented in a compact form that allows for the efficient identification of profile changes in metabolomic data. \par

## Methods
### Circos plots for association analyses
`EpiViz` adapts and builds on the `Circlize` foundation and uses elements of the `ComplexHeatmap` `R` package to do this. Circos plots are composed of six elements: template, plotting space, tracks, sections, data, legend (Figure \@ref(fig:epiviz-annotated-circos-plot)). The template element is a square of defined proportions within which information is plotted. Each additional element is layered onto the template one after the other. The plotting space element is an empty circle which is layered and centred on top of the template. Data is plotted on to the plotting space. An optional extra of the Circos plot, the legend element takes the dimensions of the template and creates a separate plotting space that can be layered on to the bottom of the template element. \par

The plotting space is separated into tracks and sections. Tracks are laid down as rings within the plotting space. Each track represents a single element of information such as an exposure or method. Tracks are numbered from the outside to the centre of the circle and can be coloured separately. Sections divide the plotting space into distinct areas, much like a pie chart. Sections are defined by a grouping variable such as a metabolite profile (e.g. class or sub class). A section track is placed at the outside of the tracks to give a header for each section. The header is referenced in the legend element. \par

Once the template, plotting space, tracks and sections are laid down, coordinates for each section and track location can be called to plot the data element. Each track and section coordinate, e.g. track 2 section 3, is treated as an individual plotting space. As such, data can be plotted based on the following coordinates: track, section, $X$, $Y$. The $X$ axis of each track is defined by the number of rows in the data frame, i.e a data frame with 100 rows will have an $X$ axis of length 100 with each row given an $X$ axis coordinate from 1:100. The $Y$ axis is defined by the minimum and maximum of the data for that track. As such, each track and section coordinate, e.g. track 2 section 3, can be considered an individual plot with a $Y$ axis that is shared by all of the sections in that track. For each position on the $X$ axis the label element of each row is plotted outside of the section header. \par

(ref:epiviz-annotated-circos-plot-cap) **Circos plot highlighting the six elements used to plot data**.

(ref:epiviz-annotated-circos-plot-scap) Circos plot highlighting the six elements used to plot data

```{r epiviz-annotated-circos-plot, echo=FALSE, out.width='100%', fig.align = 'center', fig.cap='(ref:epiviz-annotated-circos-plot-cap)', fig.scap='(ref:epiviz-annotated-circos-plot-scap)'}
knitr::include_graphics("data/visualisation/figures/annotated_circos_plot.pdf")
```

### Implementation
`EpiViz` is a `Shiny` web application and `R` package. `R`[@r2019] version 3.6.2 and `Shiny`[@Chang2019] version 1.4.0 were used to develop the web application; `R` version 3.6.2 was used to develop the `R` package. `Shiny` is an `R` package that enables the development and deployment of web applications written in the `R` programming language. Development of `EpiViz` was progressive and feedback from colleagues was vital in this process. \par

### Operation
The [web application](https://mattlee.shinyapps.io/EpiViz/) is publicly accessible and held under an [MIT license](https://github.com/mattlee821/EpiViz/blob/master/LICENSE.txt). The web application has been tested on computers running macOS (version 10.14) and Windows (version 10) using: Internet Explorer (version 11; Windows), Google Chrome (version 79; macOS and Windows), Safari (version 13; macOS). \par

The [`R` package](https://github.com/mattlee821/EpiViz/tree/master/R_package) is publicly accessible through GitHub and held under an[MIT license](https://github.com/mattlee821/EpiViz/blob/master/LICENSE.txt). The `R` package is accessible on all computers with `R` version 3.3.0 or higher and has been tested on macOS (version 10.14) and Windows (version 10) running `R` version 3.3.0 or higher. \par

A legend function is available for both the web application and `R` package and is implemented using functions from the `ComplexHeatmap`[@Gu2016] `R` package. By default, the colours used for the Circos plot in both the web-application and `R` package are accessible colours identified using [i want hue](https://medialab.github.io/iwanthue/). Example data is provided on the web application *Home* tab and with the `R` package using the function `EpiViz::EpiViz_data*()` where `*` is 1-3. Example data can be produced for use with the web-application and `R` package using code on [GitHub](https://github.com/mattlee821/EpiViz/blob/master/R_package/data-raw/example_data_simulation.R). \par

#### `Shiny` web application
In order to use the [web application](https://mattlee.shinyapps.io/EpiViz/) a web-browser and an internet connection of at least 1Mbps is required. No other system requirements are needed. Upon opening the web application, users are shown a number of example Circos plots created with the application and are directed towards the *Home* tab. The *Home* tab provides users with a short summary of the application, a link to the `R` package, and example data for use with the app. \par

The *How to* tab provides instructions for using the application and is split into three: Step 1 deals with the preparation of data to be used with the application, step 2 deals with how to use the application, and step 3 provides information on potential next steps. Users are instructed to upload one data frame per track of the Circos plot. Each data frame should be a tab delimited text file and `R` code is provided for users to achieve this. The user is guided through an example utilizing the example data. \par

Having prepared their data as instructed in Step 1, users select the *Analysis* tab and upload one data frame for each track of the Circos plot. If formatted as described in the *How to* tab, descriptive information including a volcano plot will be produced automatically. The *Plot* tab holds a series of drop down lists which auto-populate with the column names from the first uploaded data frame. The following column information is required for the Circos plot: label, group, estimate, p-value, lower confidence interval, upper confidence interval. \par

A p-value adjustment ($X$) can be provided to indicate a significance threshold in the format $\displaystyle \frac{0.05} {X}$. On the Circos plot, a solid point is indicated as reaching the significance threshold. An optional legend function is provided and users can choose the labels for the legend points. The legend is auto-populated using the levels in the `group` column of the up-loaded data frame. Finally, users can select to use *Accessible colours* (default) or *Not accessible colours* for the colours of the points on the Circos plot. \par

#### `R` package
The `EpiViz` `R` package is accessed using `R` version 3.3.0 or above. Documentation for using the package is available as a `README` on [GitHub](https://github.com/mattlee821/EpiViz/tree/master/R_package). The `README` includes use cases and troubleshooting. The `R` package can be installed and loaded into the current `R` session directly from GitHub into with the following code: 

```{r eval=FALSE}
# Install devtools
install.packages("devtools")
library(devtools)

# Install directly from GitHub
devtools::install_github("mattlee821/EpiViz/R_package")
library(EpiViz)
```

Once installed, users can use the example data provided with the package to produce Figure \@ref(fig:epiviz-figure-example-circos1), which illustrates the three different types of track available (point, line, bar), using the following `R` code:

```{r eval=FALSE}
circos_plot(track_number = 3,
            track1_data = EpiViz::EpiViz_data1,
            track2_data = EpiViz::EpiViz_data2,
            track3_data = EpiViz::EpiViz_data3,
            track1_type = "points",
            track2_type = "lines",
            track3_type = "bar",
            label_column = 1,
            section_column = 9,
            estimate_column = 2,
            pvalue_column = 3,
            pvalue_adjustment = 1,
            lower_ci = 4,
            upper_ci = 5,
            lines_column = 2,
            lines_type = "o",
            bar_column = 2)
```

(ref:epiviz-figure-example-circos1-cap) **Example Circos plot using `EpiViz` `R` package and example data**.

(ref:epiviz-figure-example-circos1-scap) Example Circos plot using `EpiViz` `R` package and example data

```{r epiviz-figure-example-circos1, echo=FALSE, out.width='100%', fig.align = 'center', fig.cap='(ref:epiviz-figure-example-circos1-cap)', fig.scap='(ref:epiviz-figure-example-circos1-scap)'}
knitr::include_graphics("data/visualisation/figures/r_package_3_track_type_plot.pdf")
```

In producing plots using the `R` package, users are advised to save as `PDF` using the below code. This is because the Circos plot is designed to be larger than normal plots, as such viewers like the `R` Studio Viewer pane display the plot in a compressed form, squashing the plot. Viewers like this should be used only as a guide when making the plot. `PDF` files can be converted to other image formats without compression. \par

Users are advised to iterate the sizing of the plot adjusting the width and height arguments to get the desired plot size and then adjust the point size argument to increase and decrease the size of the information in the plot (labels, points, lines etc.). The numbers provided work with most plots and were used for all examples. In addition, users can adjust the size of the Circos plot directly using the argument `circle_size`. The default for `circle_size` is 25, smaller numbers increase the size of the circle and larger numbers decrease the size of the circle:

```{r eval=FALSE}
pdf("my_circos_plot.pdf",
    width = 30, height = 30, pointsize = 35)
circos_plot(...
            circle_size = 25)
dev.off()
```

Finally, users can adjust the height of tracks individually using `track*_height` where `*` refers to the track number. The default for each track is 20 percent of the total size of the circle (the section track is fixed at 5 percent):

```{r eval=FALSE}
circos_plot(...
            track1_height = 0.2,
            track2_height = 0.2,
            track3_height = 0.2)
```

In order to minimise the time required to maintain the `R` package further customisation is achieved by the users themselves. The function is written to aid this customisation with *Default parameters* and *Customisable parameters* located at the top of the `circos_plot()` function script. Guidance and code is provided on [GitHub](https://github.com/mattlee821/EpiViz/tree/master/R_package) to aid user customisation, for example changing track colours requires users to navigate to the *Customisable parameters* section and then to the *Colours* section and replace the following lines with their desired colours:

```{r eval=FALSE}
discrete_palette <- c("#00378f", # track 1 colour
                      "#ffc067", # track 2 colour
                      "#894300") # track 3 colour
```

## Discussion
Circos plots have been used to visualise and provide overview of large metabolomic association analyses[@Shin2014; @Kettunen2012; @Kettunen2016; @Bos2021; @Taylor2019]. The creation of Circos plots for Taylor et al. (2019)[@Taylor2019] was cumbersome, revisions were time consuming and inefficient with each Circos plot requiring bespoke code. `EpiViz` simplifies and streamlines the process of making Circos plots. In Bos et al. (2020)[@Bos2021], we used `EpiViz` to summarise and visually compare observational and Mendelian randomization analyses of multiple sleep traits on over 100 metabolites. `EpiViz` made producing and refining Circos plots for this analysis faster and more efficient. \par

The web application is intended for researchers with limited to no experience of `R`, while those with some experience can gain added functionality from the `R` package. Though both are intended to be low maintenance, there is scope for further development. Of broad appeal to both is the inclusion of new plotting methods such as the chord diagram which shows relatedness between two nodes. In the case of metabolites a chord diagram would provide greater understanding about the relationship between different sections. In addition, the ability to filter and choose specific sections to display would simplify the presentation of key results -- currently this is achieved with supplementary forestplots. \par

The web application is less flexible than the `R` package, however there are specific extensions of interest. Firstly interactive plots would provide users the ability to expand sections of interest and display additional information. Rather than a final presentation of the data this would primarily focus on exploration of data. Additionally, introducing features from the `R` package such as different plot types will broaden the appeal of the web application. These additions are likely to increase the costs of maintenance however. \par

The `R` package has extensive scope for enhancement and presents a more efficient investment of time for extensions over the web application due to the flexibility of `R`. Currently, only a single data point can be plotted on each track at each $X$ axis coordinate. This is to prevent over-crowding. However, if a data set has 50 points for example, adding a second point to each $X$ axis coordinate of the track would not overcrowd the plot. \par

The current limited customisation afforded with the web application and the `R` package is designed to reduce the maintenance costs. As a result, functions that users may wish to use, such as multiple point estimates on a single track, must be coded by the user in `R`. Guidance is provided on GitHub for such customisations. Similarly, although designed to be efficient, users must provide data in a specific format in order to use the web application and `R` package. `R` code is provided to aid this effort but the requirement for users with no `R` experience to create a specific data frame will be off-putting for some. \par

In order to achieve the desired goal of providing global overview of large association analyses Circos plots must be much larger than traditional plots such as forest plots. As such, Circos plots are best utilised as an online tool rather than in print. Circos plots are best viewed using a screen with the ability to zoom in to sections where more detail is required. This fact will reduce the utility of this type of Circos plot in publications as fixed margins will limit the size and resolution of the plot. Resolution can be improved by saving the Circos plots as a `PDF` file. Many journals do not accept `PDF` images and will convert these into lossless formats such as `PNG` or `JPEG` reducing the resolution provided for readers. \par

`EpiViz` is simple and efficient to use. The web app provides a platform for quick and simple plots to be generated while the `R` applicaton provides customisation. `EpiViz` has been successfully used to interrogate large metabolomic association analyses[@Bos2021] and will be used throughout this thesis to achieve the same goals. \par

\newpage
