---
output: 
    thesisdown::thesis_pdf: default
bibliography: bib/thesis.bib
csl: csl/nature.csl
space_betwee_paragraphs: true
fig_caption: true
always_allow_html: yes
link-citations: true
toc-depth: 2
lot: true
lof: true
indent: true
header-includes: # include other LaTeX packages here
    \usepackage{booktabs}
    \usepackage{longtable}
    \usepackage{siunitx}
    \usepackage[left]{lineno}
    \usepackage{float}
    \floatplacement{figure}{H}
    \linenumbers
    \pagestyle{plain}
    \raggedbottom 
    \usepackage{indentfirst}
editor_options: 
  chunk_output_type: console
---

```{r include=FALSE}
# word count = 
doc.type <- knitr::opts_knit$get('rmarkdown.pandoc.to') # this will tell the .Rmd file what output you are knitting to (word, pdf, html) so that you cna use if else statements when making tables - html/pdf output tables do not work well in word. For word we need to use the 'flextable' packaged to make tables.
## packages
library(dplyr)
library(kableExtra)
library(flextable)
options(knitr.kable.NA = '')
```

# Associations between multiple measures of adiposity and metabolites: observational analysis {#observational} 
```{r include=FALSE}
ALSPAC_N <- read.table("data/observational/data/metabolomics/data_prep/other/ALSPAC_N.txt", header = T, sep = "\t")
ALSPAC_QC_N <- read.table("data/observational/data/metabolomics/data_prep/final/qc/ALSPAC_QC_N.txt", header = T, sep = "\t")
```


<style>
body {
text-align: justify}
</style>

## Chapter Summary {-}
In Chapters \@ref(introduction) and \@ref(systematic-review), the link between adiposity and disease from existing literature was presented in observational and causal analysis frameworks. This work highlighted that adiposity is associated, likely causally, with many outcomes. A key takeaway from Chapter \@ref(introduction) was that the underlying mechanisms of disease development are not well understood, but that some relationships between adiposity and disease may be explained by changes in metabolic pathways. In this chapter, observational epidemiological methods are used to explore associations between adiposity and metabolites. The aim of this chapter is to provide an observational grounding for subsequent causal analysis work in Chapters \@ref(MR) and \@ref(mediation), and is the first of two analyses that will assess the same question with different study designs (i.e., triangulation) with the nest study design presented in Chapter \@ref(MR). \par

\newpage

## Introduction {#observational-intro} 
Adiposity is associated with an increased risk of numerous diseases, as well as mortality[@Flegal2013; @Lee2015; @Elagizi2018; @Jenkins2018; @Pischon2008; @Rost2018; @Mørkedal2011; @Dong2018; @Lee2018; @Bigaard2004] (see Chapters \@ref(introduction) and \@ref(systematic-review)). There is a need to understand the mechanisms underlying these associations so that intervention strategies, which are challenging to implement effectively, can be targeted and more efficacious. \par

Adiposity has been linked with many downstream measures that could serve as intermediates of disease and thus be useful targets for reducing the burden of adiposity. Notably, as adiposity has been associated with many metabolites[@Wurtz2014; @Moore2014; @SantosFerreira2017; @Cirulli2019; @Lau2020; @Brachem2020; @Stevens2020; @Wulaningsih2019; @Frigerio2021; @Neeland2019; @Bachlechner2016; @Rangel-Huerta2019], previous work has highlighted metabolites[@Bray2004; @Haslam2005; @Collaboration2009] as a possible link between adiposity and diseases such as cancer[@Bhaskaran2014], coronary heart disease (CHD)[@Paul2006; @Koliaki2019], and type 2 diabetes[@Al-Goblan2014]. In the largest study to date by Wurtz et al., (2014)[@Wurtz2014], adiposity was found to influences whole classes of metabolites including amino acids, fatty acids, hormones, inflammatory markers, and lipids. Metabolites are intermediate or end products of cellular processes with multiple functions including those with energy, signalling, transportation, and structural components. Metabolic effects can be far reaching[@Johnson2016; @Wishart2019] and during homeostasis are tightly controlled. The many functions metabolites have mean that imbalances can be detrimental[@Griffin2006; @Johnson2016; @Wishart2019]. \par

Measurement of metabolites has become increasingly common among cohort studies with mass spectrometry (MS) and nuclear magnetic resonance (NMR) platforms able to perform targeted, semi-targeted, and untargeted assays. Although there is evidence to show the relationship between adiposity and metabolites varies between sexes and over time[@Wurtz2014], the focus of many analyses has been on body mass index (BMI) as a measure for overall adiposity, has included small sample sizes, and has looked at a single time-point. While BMI correlates well with, and is a predictor of, many health outcomes[@Donato1998; @NIH2013], it is a crude measure of adiposity with several issues, not least the inability to differentiate lean and fat mass[@Flegal2009]. Evidence highlights the importance and utility of combining complimentary measurements of adiposity[@WorldHealthOrganisation2008; @Collaboration2009]. As many studies have focussed on the effect of adiposity on metabolites in adulthood and given the effects of adiposity are shown to present early in life[@Bibbins-Domingo2007; @Hannon2005; @Baer2005; @Swerdlow2002; @Biro2010], there is a need to understand how and importantly when, adiposity-related metabolites may be of most use in reducing the burden of adverse health outcomes. Investigation of the effects of adiposity over time may therefore provide additional information on the relationship between adiposity and metabolites. \par

The Avon Longitudinal Study of Parents and Children (ALSPAC), a longitudinal birth cohort study, with repeated metabolomic and anthropometric measures, provides an opportunity to expand on the current literature. In this chapter, observational analysis assessing the association between multiple measures of adiposity and NMR derived metabolites provide a basis from which to investigate causality with a triangulation approach. \par

## Methods {#observational-methods} 
Data were available from ALSPAC for exposures (measures of adiposity), outcomes (metabolites), and potential confounders. Exposures included BMI, waist hip ratio (WHR) and body fat percentage (BF). Metabolomic data were available for up to `r ALSPAC_N$combined_metabolites[[1]]` NMR derived metabolites. These, predominantly lipid-based, metabolites include directly measured metabolites such as the amino acids tyrosine and phenylalanine, as well as derived (not-directly measured) metabolite measures such as the ratio of saturated fatty acids to total fatty acid. Throughout, "metabolites" is used to refer to both directly measured and derived (not-directly measured) measures, otherwise they are refered to as "directly measured metabolites" and "derived/not-directly measured metabolites" respectively. \par

Covariables included age, sex, mother's or own education, smoking history, alcohol history, diet, and physical activity. Covariables were chosen as evidence has shown an association between them and adiposity and metabolites. All data were pre-exisiting, that is I did not collect this data. All analyses and data manipulation were performed using `R` (version 3.6.2)[@r2019]. Specific `R` packages are described, where appropriate. All code for this work is available on [GitHub](https://github.com/mattlee821/000_thesis/tree/master/index/data/observational). \par

### Data overview 
ALSPAC[@Fraser2013; @Boyd2013; @Northstone2019] is a prospective cohort study that invited women resident in Avon, UK with expected dates of delivery between 1st April 1991 and 31st December 1992 to participate. The initial number of pregnancies enrolled was 14,541 (for these at least one questionnaire was returned or a “Children in Focus” clinic has been attended by 19/07/99). Of these initial pregnancies, a total of 14,676 foetuses, resulted in 14,062 live births and 13,988 children alive at one year of age. The mothers and fathers associated with each pregnancy are referred to as generation 0 (G0) while the children of each eligible pregnancy (including individuals from subsequent recruitment drives) are referred to as generation 1 (G1). \par

When the oldest G1 individuals were approximately seven years of age, an attempt was made to bolster the initial sample with eligible cases who had failed to join the study originally. As a result, when considering variables collected from the age of seven onwards (and potentially abstracted from obstetric notes), there are data available for more than the 14,541 pregnancies mentioned above. The number of new pregnancies not in the initial sample (known as Phase I enrolment) that are currently represented on the built files and reflecting enrolment status at the age of 24 is 913 (456, 262, and 195 recruited during Phases II, III, and IV respectively), resulting in an additional 913 G1 individuals being enrolled. The phases of enrolment are described in more detail in the cohort profile paper and its update[@Fraser2013; @Boyd2013; @Northstone2019]. The total sample size for analyses using any data collected after the age of seven is therefore 15,454 pregnancies, resulting in 15,589 foetuses, of which 14,901 were alive at one year of age. \par

The study [website](http://www.bristol.ac.uk/alspac/) contains details of all the data that is available through a fully searchable data dictionary and [variable search tool](http://www.bristol.ac.uk/alspac/researchers/our-data/). Ethical approval for the study was obtained from the ALSPAC Ethics and Law Committee and the [Local Research Ethics Committees](http://www.bristol.ac.uk/alspac/researchers/research-ethics/). Informed consent for the use of data collected via questionnaire and clinics was obtained from participants following recommendations of the ALSPAC Ethics and Law Committee at the time. Full details of the ALSPAC consent procedures are available on the [study website](http://www.bristol.ac.uk/alspac/researchers/research-ethics/). \par

Data in ALSPAC are split by clinic visits. For this work, metabolomic data were available for G1 individuals from the following clinics: Focus at 7 (~8 years old), Focus at 8 (~9 years old), Before Breakfast Study (~8 years old), Teen Focus 3 (~18 years old), Teen Focus 4 (~17 years old), and Focus at 24 (~24 years old). Metabolomic data for G0 individuals were available from: Focus on Mothers 1 (~48 years old), Focus on Mothers 2 (~51 years old), and Focus on Fathers 1 (~53 years old). In order to maximize the sample size at each metabolomic clinic, data were combined where clinics were within a similar age range. Data were combined in the following groups for G1 individuals: Focus at 7 and Before Breakfast Study (children), and Teen Focus 3 and Teen Focus 4 (adolescents). The Focus at 24 clinic is referred to as young adults. For G0 individuals data from Focus on Mothers 1, Focus on Mothers 2, and Focus on Fathers 1 were combined into an adults group. For these combined data sets, duplicate individuals (i.e., those attending both clinics) were identified, and the measurement from the most recent clinic was dropped. \par

All data on exposures and covarialbes were obtained from the same clinic from which metabolomic data were collected. Where data on exposures and covarialbes were not available at the metabolomic clinic visit, they were obtained from the most recent clinic with available data. The Before Breakfast Study, unlike the other clinics, only collected metabolomic data, as such data on exposures and covarialbes were extracted for these individuals from the Focus at 8 clinic. metabolomic data for each time point were extracted first and subsequent data on exposures and covarialbes were extracted for individuals with metabolomic data. The metabolomic data were provided with standard exclusions for identifiable individuals and those with withdrawn consent already excluded. \par

G0 and G1 individuals were independent, however there is familial overlap between them. Importantly, G1 individuals were not independent, that is, the same individuals will have attended multiple clinics. As such, any effects specific to an individual will be propagated through all G1 clinics that individual attends. \par

### Exposures: adiposity {#observational-methods-exposures}
Measures of adiposity (BMI, WHR, and BF) were obtained for all individuals with available metabolomic data. Data on WHR was not available for the Teen Focus 3 and 4 clinics. BMI was calculated as $\displaystyle \frac{weight(kg)}{height (m^2)}$ and WHR as $\displaystyle \frac{waist\ circumference\ (cm)} {hip\ circumference\ (cm)}$. Height was measured once to the last complete mm using the Harpenden Stadiometer. Individuals were positioned with their feet flat and heels together, standing straight so that their heels, calves, buttocks and shoulders came into contact with the vertical backboard of the stadiometer. The headboard was lowered until it touched the individual's head and a 1kg weight was placed on the headboard to ensure head contact and to minimise hair thickness. Weight was measured once using the Tanita Body Fat Analyser (Models TBF 305 and 401A) or electronic bathroom scales, if the individual had a pacemaker. Individuals were encouraged to pass urine and undress to their underclothes. Individuals stepped onto the measuring platform which had been wiped with disinfecting alcohol and positioned so that both feet were located in parallel with the toe and heel in contact with their respective electrodes. Measurement was completed when the weight and fat ratio readings were fixed and the buzzer beeped. Weight was measured to the nearest 50g for G1 individuals and to the nearest 0.1kg for G0 individuals. For all G1 individuals, ‘Female Standard’ was entered into the Tanita Body Fat Analyser as the sex variable. Circumferences were measured using the Seca 200 or 201 body tension tape and were repeated twice for accuracy. \par

BF was measured using dual-energy x-ray absorptiometry (DXA) in all individuals except for individuals from Focus at 7. Briefly, measurement required individuals to be prone and stationary while a Lunar prodigy narrow fan beam densitometer performed a whole body DXA scan. Data were processed using Lunar Prodigy software. Individuals did not have measurements taken if they: were pregnant, had a radiological investigation using contrast media within the week before the DXA scan, had a recent nuclear medicine investigation with persistent radioactivity or weighed greater than 159kg. BF was calculated as $\displaystyle \frac{fat\ mass} {fat\ mass + fat\ free\ mass}\ *\ 100$. \par

For Focus at 7, BF was not measured, instead bioeletrical impedance data, which were converted to estimates of BF, were available. Briefly, individuals were encouraged to pass urine and undress to their underclothes. A Tanita Body Fat Analyser (Model TBF 305) was used to measure weight and impedance. Height was entered to the nearest cm and 'Female Standard' was used as the sex variable for all individuals. The Tanita Body Fat Analyser TBF 305 is a single frequency (50kHz) leg-to-leg device. In single frequency devices, impedance is a representation of resistance which is related to the volume of water (which one assumes makes up the majority of fat free mass (FFM)), as such, the higher the resistance (impedance) the greater the amount of FFM. Calculation of BF from the impedance measure is only possible at the time of measurement, however these derived BF measures were not stored and the equation to calculate them was not available from the manufacturer. Previous work[@Chouinard2007] has shown that comparison of BF derived from the manufacturer's equation and an alternative[@Jebb2000] showed little difference in resulting BF estimates. The alternative equation was derived in a study involving 205 (101 women) healthy adults with a mean age of 43.8 (SD = 16) for men and 40.4 (SD = 13.6) for women. The alternative equation, where $Z$ is the impedance measure from the device in ohms, height is in metres, weight is in kilograms, age is in years, and female-specific components are given as $19.6\ +\ ln(height)$, is given as: \par

\begin{equation}
\begin{split}
  BF = -156.1 - 89.1\ ln(height) \\
  \ +\ 45.6\ ln(weight) \\
  \ +\ 0.120\ age\ \\
  \ +\ 0.0494\ Z \\
  \ +\ (19.6\ ln(height))
\end{split}
  (\#eq:BF)
\end{equation}

Given that the equation was derived from adult data, its application to child data in ALSPAC was explored in this chapter. A raw impedance measure, from a similar model (Tanita Body Fat Analyser (Model TBF 401A)), was obtained for individuals from Teen Focus 3 and 4 (i.e., where both DXA and impedance measures were available) and the equation was used to compare BF derived from the impedance device and BF measured with DXA in adolescents. Exploration involved visual inspection of distribution and Spearman's correlation with BMI, height, weight and other BF measures from Teen Focus 3 and 4. The same observations were carried out for raw impedance. The calculated BF estimates were positively correlated with height and weight, however there were negative values of calculated BF. As the estimates derived in linear models are in reference to the per unit increase in an exposure (rather than the range), the absolute value of the exposure does not need to positive. Thus, the negative estimates of BF will not impact the inference of a linear regression between BF and any metabolite. As such, BF calculated using equation \@ref(eq:BF) was used in subsequent analyses as a measure of BF in children. \par

<!-- The raw impedance values correlated well with height and weight, and in adolescents were highly correlated with impedance derived and DXA derived BF estimates (See Appendix \@ref(appendix-observational-BF-validation)). -->

### Outcomes: metabolites
```{r QC-data, echo=FALSE}
load("data/observational/data/metabolomics/data_prep/final/qc/children/MetaboQC_release/ReportData.Rdata")
children_samples_removed <- nrow(raw_data$sample_data) - nrow(qc_data$sample_data)
children_features_removed <- ncol(raw_data$metabolite_data) - ncol(qc_data$metabolite_data)
children_independent <- nrow(qc_data$varexp)

load("data/observational/data/metabolomics/data_prep/final/qc/adolescents/MetaboQC_release/ReportData.Rdata")
adolescents_samples_removed <- nrow(raw_data$sample_data) - nrow(qc_data$sample_data)
adolescents_features_removed <- ncol(raw_data$metabolite_data) - ncol(qc_data$metabolite_data)
adolescents_independent <- nrow(qc_data$varexp)

load("data/observational/data/metabolomics/data_prep/final/qc/young_adults/MetaboQC_release/ReportData.Rdata")
young_adults_samples_removed <- nrow(raw_data$sample_data) - nrow(qc_data$sample_data)
young_adults_features_removed <- ncol(raw_data$metabolite_data) - ncol(qc_data$metabolite_data)
young_adults_independent <- nrow(qc_data$varexp)

load("data/observational/data/metabolomics/data_prep/final/qc/adults/MetaboQC_release/ReportData.Rdata")
adults_samples_removed <- nrow(raw_data$sample_data) - nrow(qc_data$sample_data)
adults_features_removed <- ncol(raw_data$metabolite_data) - ncol(qc_data$metabolite_data)
adults_independent <- nrow(qc_data$varexp)
```
Metabolomic data were measured using the same NMR platform for all individuals. Briefly, high-throughput proton (^1^H) NMR assays were performed on ethylenediaminetetraacetic acid (EDTA) plasma/serum samples. Samples were fasted. Measurements were taken at three molecular windows (lipoprotein lipids, low molecular-weight metabolites, and lipid extracts) enabling broad quantification of metabolomic measures. These measures also included derived measures, lipoprotein particle sizes, and fatty acid ratios, inclusion of which has shown to increase overall power in statistical analyses[@Altmaier2008; @Gieger2008; @Suhre2010]. Metabolite values for each individual were provided by the NMR platform in the originally measured units (e.g., mmol/l). Derived metabolite values are as a %. Full details on the NMR methodology has previously been described[@Soininen2009; @Inouye2010; @Kettunen2012; @Soininen2015] and is available from the [ALSPAC data dictionary](http://www.bristol.ac.uk/alspac/researchers/access/) (data dictionary identifiers: children = D5704, mothers = D5705, fathers = D5700). The Before Breakfast study does not have a documentation file and is described elsewhere[@Ong2004]; fasted metabolomic data, not the post glucose challenge metabolomic data, were taken from The Before Breakfast Study. Descriptions of metabolites are available on [GitHub](https://github.com/mattlee821/000_thesis/blob/master/index/data/observational/tables/metabolites.txt). \par

The spectral NMR data were processed by Nightingale Health and provided by the ALSPAC team as a processed file with identifiable individuals (triplets/quadruplets) and individuals who had withdrawn consent removed. Some mothers and fathers were duplicated in this data due to the way in which mothers were originally enrolled into the study and assigned IDs. If a mother enrolled with two different pregnancies (both having an expected delivery date within the recruitment period (April 1991-December 1992)), she will have two separate IDs. A father associated with both of these pregnancies will also be duplicated. Duplicate measurements for mothers and fathers were removed. No metabolites were excluded at this stage, however the number of metabolites available for each clinic visit differed as metabolites were added and removed over time by Nightingale as validations change. \par

Pre-analysis processing of metabolomic data is important as there can be quality issues with sample and metabolite data[@Hughes2021]. However, there is no standardised method for performing pre-analysis processing and for deciding what thresholds to use to exclude metabolites and individuals from downstream analyses. The `R` package [`metaboprep`](https://github.com/MRCIEU/metaboprep)[@Hughes2021] can be used to process data before analyses using a transparent and reproducible workflow. Here, after combining clinic visit data, where appropriate, `metaboprep` (version 0.0.1) was used to identify and exclude individuals and metabolites that did not meet certain requirements. This process was performed twice, firstly including and secondly excluding the derived metabolomic measures from missingness and clustering analyses to see if these derived measures were unduly influencing exclusions. Briefly, individuals and then metabolites, with high missingness ($\geq$ 80%) were removed. Missingness was then re-calculated for individuals and metabolites, with removal based on 20% missingness. Individuals were then removed based on total sum abundance, considering outliers as $\geq$ 5 standard deviations (SD) away from the mean. Using this metabolite dataset, a dendrogram using complete-linkage and a Spearman's rho distance matrix was produced, and a set of clusters identified based on a Spearman's rho of 0.5. For each cluster, the metabolite with the least missingness was tagged as the representative feature. Finally, principal component (PC) analysis was conducted using the representative features to evaluate structure among individuals. Outliers were identified as being $\geq$ 5 SDs away from the mean of PC 1 and 2, and were excluded. Additionally, in order to gain an overview of the variability of metabolite concentrations, the mean, SD of the mean, median, and range of metabolite concentrations were visually compared across age groups. \par

### Covariables
```{r ALSPAC-confounders, echo=FALSE, warning=FALSE, error=FALSE}
children_qc <- read.table("data/observational/data/analysis/children_qc_phenofile.txt", header = T, sep = "\t")
adolescents_qc <- read.table("data/observational/data/analysis/adolescents_qc_phenofile.txt", header = T, sep = "\t")
young_adults_qc <- read.table("data/observational/data/analysis/young_adults_qc_phenofile.txt", header = T, sep = "\t")
adults_qc <- read.table("data/observational/data/analysis/adult_qc_phenofile.txt", header = T, sep = "\t")

# table ====
data <- data.frame(group = c("Metabolomic N",
                             "age", "age", "age",
                             "sex","sex",
                             "education", "education", "education", "education", "education",
                             "smoking",
                             "alcohol", "alcohol", "alcohol", "alcohol", "alcohol",
                             "diet", "diet", "diet",
                             "physical activity", "physical activity","physical activity"),
                   subgroup = c("N",
                                "N","mean","sd",
                                "N","female",
                                "1","2","3","4","5",
                                "N",
                                "1","2","3","4","5",
                                "N","mean","sd",
                                "N/1","mean/2","sd/3"),
                   children = c(ALSPAC_QC_N$N[1],
                                nrow(children_qc[!is.na(children_qc$age),]),round(mean(children_qc$age, na.rm  = T), 2),round(sd(children_qc$age, na.rm  = T), 2),
                                nrow(children_qc[!is.na(children_qc$sex),]),table(children_qc$sex)[[2]],
                                (table(children_qc$maternal_education)[[1]]/ALSPAC_QC_N$N[1] * 100),(table(children_qc$maternal_education)[[2]]/ALSPAC_QC_N$N[1] * 100),(table(children_qc$maternal_education)[[3]]/ALSPAC_QC_N$N[1] * 100),(table(children_qc$maternal_education)[[4]]/ALSPAC_QC_N$N[1] * 100),(table(children_qc$maternal_education)[[5]]/ALSPAC_QC_N$N[1] * 100),
                                "-",
                                "-","-","-","-","-",
                                nrow(children_qc[!is.na(children_qc$diet),]),round(mean(children_qc$diet, na.rm  = T), 2),round(sd(children_qc$diet, na.rm  = T), 2),
                                "-","-","-"),
                   adolescents = c(ALSPAC_QC_N$N[2],
                                   nrow(adolescents_qc[!is.na(adolescents_qc$age),]),round(mean(adolescents_qc$age, na.rm  = T), 2),round(sd(adolescents_qc$age, na.rm  = T), 2),
                                   nrow(adolescents_qc[!is.na(adolescents_qc$sex),]),table(adolescents_qc$sex)[[2]],
                                   (table(adolescents_qc$maternal_education)[[1]]/ALSPAC_QC_N$N[2] * 100),(table(adolescents_qc$maternal_education)[[2]]/ALSPAC_QC_N$N[2] * 100),(table(adolescents_qc$maternal_education)[[3]]/ALSPAC_QC_N$N[2] * 100),(table(adolescents_qc$maternal_education)[[4]]/ALSPAC_QC_N$N[2] * 100),(table(adolescents_qc$maternal_education)[[5]]/ALSPAC_QC_N$N[2] * 100),
                                   nrow(adolescents_qc[!is.na(adolescents_qc$ever_smoked),]),
                                   (table(adolescents_qc$frequency_alcohol)[[1]]/ALSPAC_QC_N$N[2] * 100),(table(adolescents_qc$frequency_alcohol)[[2]]/ALSPAC_QC_N$N[2] * 100),(table(adolescents_qc$frequency_alcohol)[[3]]/ALSPAC_QC_N$N[2] * 100),(table(adolescents_qc$frequency_alcohol)[[4]]/ALSPAC_QC_N$N[2] * 100),(table(adolescents_qc$frequency_alcohol)[[5]]/ALSPAC_QC_N$N[2] * 100),
                                   nrow(adolescents_qc[!is.na(adolescents_qc$diet),]),round(mean(adolescents_qc$diet, na.rm  = T), 2),round(sd(adolescents_qc$diet, na.rm  = T), 2),
                                   nrow(adolescents_qc[!is.na(adolescents_qc$physical_activity),]),round(mean(adolescents_qc$physical_activity, na.rm  = T), 2),round(sd(adolescents_qc$physical_activity, na.rm  = T), 2)),
                   young_adults = c(ALSPAC_QC_N$N[3],
                                    nrow(young_adults_qc[!is.na(young_adults_qc$age),]),round(mean(young_adults_qc$age, na.rm  = T), 2),round(sd(young_adults_qc$age, na.rm  = T), 2),
                                    nrow(young_adults_qc[!is.na(young_adults_qc$sex),]),table(young_adults_qc$sex)[[2]],
                                    (table(young_adults_qc$maternal_education)[[1]]/ALSPAC_QC_N$N[3] * 100),(table(young_adults_qc$maternal_education)[[2]]/ALSPAC_QC_N$N[3] * 100),(table(young_adults_qc$maternal_education)[[3]]/ALSPAC_QC_N$N[3] * 100),(table(young_adults_qc$maternal_education)[[4]]/ALSPAC_QC_N$N[3] * 100),(table(young_adults_qc$maternal_education)[[5]]/ALSPAC_QC_N$N[3] * 100),
                                    nrow(young_adults_qc[!is.na(young_adults_qc$ever_smoked),]),
                                    (table(young_adults_qc$frequency_alcohol)[[1]]/ALSPAC_QC_N$N[3] * 100),(table(young_adults_qc$frequency_alcohol)[[2]]/ALSPAC_QC_N$N[3] * 100),(table(young_adults_qc$frequency_alcohol)[[3]]/ALSPAC_QC_N$N[3] * 100),(table(young_adults_qc$frequency_alcohol)[[4]]/ALSPAC_QC_N$N[3] * 100),(table(young_adults_qc$frequency_alcohol)[[5]]/ALSPAC_QC_N$N[3] * 100),
                                    "-","-","-",
                                    nrow(young_adults_qc[!is.na(young_adults_qc$physical_activity),]),round(mean(young_adults_qc$physical_activity, na.rm  = T), 2),round(sd(young_adults_qc$physical_activity, na.rm  = T), 2)),
                                    
                   adults = c(ALSPAC_QC_N$N[4],
                              nrow(adults_qc[!is.na(adults_qc$age),]),round(mean(adults_qc$age, na.rm  = T), 2),round(sd(adults_qc$age, na.rm  = T), 2),
                              nrow(adults_qc[!is.na(adults_qc$sex),]),table(adults_qc$sex)[[2]],
                              (table(adults_qc$education)[[1]]/ALSPAC_QC_N$N[4] * 100),(table(adults_qc$education)[[2]]/ALSPAC_QC_N$N[4] * 100),(table(adults_qc$education)[[3]]/ALSPAC_QC_N$N[4] * 100),(table(adults_qc$education)[[4]]/ALSPAC_QC_N$N[4] * 100),(table(adults_qc$education)[[5]]/ALSPAC_QC_N$N[4] * 100),
                              nrow(adults_qc[!is.na(adults_qc$ever_smoked),]),
                              (table(adults_qc$frequency_alcohol)[[1]]/ALSPAC_QC_N$N[4] * 100),(table(adults_qc$frequency_alcohol)[[2]]/ALSPAC_QC_N$N[4] * 100),(table(adults_qc$frequency_alcohol)[[3]]/ALSPAC_QC_N$N[4] * 100),(table(adults_qc$frequency_alcohol)[[4]]/ALSPAC_QC_N$N[4] * 100),(table(adults_qc$frequency_alcohol)[[5]]/ALSPAC_QC_N$N[4] * 100),
                              "-","-","-",
                              (table(adults_qc$physical_activity)[[1]]/ALSPAC_QC_N$N[4] * 100),(table(adults_qc$physical_activity)[[2]]/ALSPAC_QC_N$N[4] * 100),(table(adults_qc$physical_activity)[[3]]/ALSPAC_QC_N$N[4] * 100))
                   )
```

Evidence has shown that age[@Yu2012; @Brennan2020], sex[@Brennan2020], education[@Zajacova2018], smoking[@Bonevski2014], alcohol[@Bonevski2014], diet[@Afshin2019], and physical activity[@ODonoghue2018] all influence the metabolomic profile and adiposity. Data for each of these covarialbes were obtained for all individuals with metabolomic data. Age was taken from the metabolomic clinic visit as months since birth. Sex was taken from the initial assessment questionnaire for G1 individuals completed by their parents. For G0, an individual's sex was as self-reported at the metabolomic clinic visit. \par

Maternal education was used as an adjustment for children, adolescents and young adults. Own education was used for mothers and mother-reported partner education was used for fathers. Specifically, mothers were asked, during their pregnancy, 'What educational qualifications do you, your partner, your mother, and your father have?' with possible answers: CSE or GCSE (D, E, F or G); O-level or GCSE (A, B, or C); A-level; qualifications in shorthand/ typing/or other skills e.g. hairdressing;, apprenticeship; state enrolled nurse; state registered nurse; City and Guilds intermediate technical; City & Guilds final technical; City & Guilds full technical; teaching qualification; university degree; no qualification; qualifications not known; not applicable; or other (please describe). This data was available as a recoded variable of 5 categories of lowest (1) to highest (5) level of education. \par

The variable capturing smoking was binary; adolescents (at the metabolomic clinic), young adults (at the metabolomic clinic), and adults (mothers were asked during pregnancy; fathers were asked at a clinic prior to the metabolomic clinic) were asked whether they had ever smoked a cigarette before. \par

<!-- fathers asked in 2013 -->

Alcohol was assessed differently for Teen Focus 3 and Teen Focus 4, Focus at 24, and G0 individuals. Individuals from Teen Focus 3 were asked what their alcohol drinking pattern was with possible answers: only ever tried drinking once/twice, used to drink sometimes [but] never drink now, sometimes drink but less than once a week, usually drink on 1 or 2 days a week, usually drink on >2 days a week but not every day, and usually drink every day. Individuals from Teen Focus 4, Focus at 24, and G0 individuals (mothers were asked at a prior clinic visit to the metabolomic clinic and fathers at the metabolomic clinic) were asked the frequency they had drinks containing alcohol with possible answers: never, monthly or less, two to four times a month, two to three times a week, four or more times a week. \par

<!-- mothers asked in 2013 -->

Diet data derived by Anderson et al., (2013)[@Anderson2013] were available for G1 individuals aged 7 and 13. Data from age 7 was matched with metabolomic data for Focus at 7 while data from age 13 was matched with metabolomic data for Teen Focus 3 and 4. Diet data were not available for young adults or adults. Data is given as predicted kilo-calories consumed per day. \par

Data on physical activity were collected differently for different clinic visits. For G1 individuals from Teen Focus 3 and Focus at 24, accelerometry data were collected at the same clinic at which metabolomic data were collected. Briefly, individuals wore an accelerometry device for the 7 days following their clinic visit whilst keeping a diary of the times they wore and took off the device. Individuals were advised to wear the accelerometer device if the following days were part of a ‘normal week’ with regards to their activity. For Focus at 24 individuals, physical activity data were the average number of minutes per day spent doing moderate to vigorous physical activity. For Teen Focus 3 individuals, physical activity data were the mean counts per minute spent doing moderate to vigorous physical activity for the whole week. For G0, individuals were asked 'Do you take part in physical activity (e.g., running, swimming, dancing, golf, tennis, squash, jogging, and bowls)?' with possible answers: no, occasionally (less than monthly), and frequently (once a month or more). Data for mothers were available in a clinic prior to the metabolomic clinic, fathers data were available at the metabolomic clinic. Physical activity data were not available for Focus at 7. \par

<!-- mothers asked in 2010 -->

### Statistical analysis
Metabolomic data were available for: Focus at 7 (n = `r ALSPAC_N$total_N[1]`; metabolites = `r ALSPAC_N$total_metabolites[1]`), Before Breakfast Study (n = `r ALSPAC_N$total_N[2]`; metabolites = `r ALSPAC_N$total_metabolites[2]`), Teen Focus 3 (n = `r ALSPAC_N$total_N[3]`; metabolites = `r ALSPAC_N$total_metabolites[3]`), Teen Focus 4 (n = `r ALSPAC_N$total_N[4]`; metabolites = `r ALSPAC_N$total_metabolites[4]`), Focus at 24 (n = `r ALSPAC_N$total_N[5]`; metabolites = `r ALSPAC_N$total_metabolites[5]`), Focus on Mothers 1 (n = `r ALSPAC_N$total_N[6]`; metabolites = `r ALSPAC_N$total_metabolites[6]`), Focus on Mothers 2 (n = `r ALSPAC_N$total_N[7]`; metabolites = `r ALSPAC_N$total_metabolites[7]`), Focus on Fathers (n = `r ALSPAC_N$total_N[8]`; metabolites = `r ALSPAC_N$total_metabolites[8]`). \par

After combining the G1 clinics Focus at 7 and Before Breakfast Study (children), and Teen Focus 3 and Teen Focus 4 (adolescents), metabolomic data were available for `r ALSPAC_N$combined_N[1]` children (mean age (SD) = 7.56 (0.36); metabolites = `r ALSPAC_N$combined_metabolites[1]`), `r ALSPAC_N$combined_N[3]` adolescents (mean age (SD) = 16.06 (1.11); metabolites = `r ALSPAC_N$combined_metabolites[3]`), and `r ALSPAC_N$total_N[5]` young adults from the Focus at 24 clinic (mean age (SD) = 24.03 (0.85); metabolites = `r ALSPAC_N$total_metabolites[5]`; Table \@ref(tab:observational-table-ALSPAC-N)). After combining the G0 clinics Focus on Mothers 1, Focus on Mothers 2, and Focus on Fathers 1, metabolomic data were available for `r ALSPAC_N$combined_N[9]` adults (mean age (SD) = 49.53 (5.32); metabolites = `r ALSPAC_N$combined_metabolites[9]`). \par

To estimate the association between measures of adiposity and metabolites, all exposures were $Z$-scored and linear regression was performed. Metabolites were not transformed as the majority did not appear to have skewed distributions after pre-analysis processing using `metaboprep`. Variables known to influence the metabolomic profile and adiposity (age, sex, education, smoking, alcohol, diet, and physical activity), were included as covarialbes. Three linear models were used to investigate potential effects of these covarialbes. Model 1 included age and sex. Model 2 included variables in model 1 and mother's/own level of education, whether respondent had ever smoked, frequency respondent had a drink containing alcohol, and predicted kilo-calories consumed per day. Model 3 comprised all variables included in model 2 and physical activity. To maximise sample size for analyses, model 1 and 2 comprised individuals with data on all covarialbes except physical activity. Model 3 comprised all individuals with data on all covarialbes. Model 2, as the most adjusted and given the reduced sample size in model 3, is presented as the main analysis for this work. \par

For all analyses, units represent the unit change in each metabolite per standard deviation change in the exposure. Metabolite abbreviations are used in figures and tables for space. For complete labels, class, subclass, and units for each metabolite, see [GitHub](https://github.com/mattlee821/000_thesis/blob/master/index/data/observational/tables/metabolites.txt). 95% confidence intervals (CIs) were calculated and a multiple testing threshold specific for each age group was applied. Multiple testing thresholds were calculated as the number of independent metabolites within the raw metabolomic data given a Spearman’s rho of approximately 0.75 among the metabolites with data for at least 20% of samples -- this was calculated during metabolite pre-analysis processing using `metaboprep`. The number of independent metabolites in each age group were calculated as: children = `r children_independent`, adolescents = `r adolescents_independent`, young adults = `r young_adults_independent`, adults = `r adults_independent`. \par

Metabolites were grouped into subclasses (grouping data provided by the metabolomic platform) based on biological pathway. Consistency in the direction of effect estimates was investigated (i) across the three linear models for each exposure within each age group and (ii) across exposures within age groups for model 2. A consistent  positive or negative direction is reported when all estimates being compared give the same direction of effect. For example, when comparing directions of effect across the three linear models for the effect of BMI on a metabolite in children, a consistent direction of effect is reported if all three models effect estimates are positive or negative. If one of the models reported a direction of effect that was positive and the other two models reported a direction of effect that was negative, then an "inconsistent direction" would be reported for the effect of BMI on that metabolite in children. In addition, the number of tests reaching the multiple testing thresholds were reported. \par

A Spearmans rho correlation analysis was used to investigate the correlation between effect estimates across exposures and age groups. Circos plots (via the `EpiViz` `R` package; see Chapter \@ref(visualisation)) were used to visualize and compare global metabolic profiles within age groups across exposures. Forest plots, created using the `ggforestplot` `R` package, were used to examine specific subclasses. Results for derived measures and lipoprotein particle size and fatty acid ratios are presented in the Appendix. Directions of effect estimates were compared to that of previous work by Wurtz et al. (2014)[@Wurtz2014] where appropriate, i.e., where metabolites were present both in the ALSPAC analyses and the Wurtz paper. \par

## Results {#observational-results} 
```{r echo=FALSE, warning=FALSE, error=FALSE}
# data ====
data <- read.table("data/observational/data/analysis/results/combined/combined.txt", header = T, sep = "\t")
data <- subset(data, subclass != "NA")
# shared metabolites ====
a <- subset(data, model == "model2")
a <- subset(a, exposure == "bmi")
b <- subset(a, group == "children")
c <- subset(a, group == "adolescents")
d <- subset(a, group == "young_adults")
e <- subset(a, group == "adults")
b <- as.data.frame(b[,1])
colnames(b) <- "metabolite"
b$group <- "children"
c <- as.data.frame(c[,1])
colnames(c) <- "metabolite"
c$group <- "adolescents"
d <- as.data.frame(d[,1])
colnames(d) <- "metabolite"
d$group <- "young_adults"
e <- as.data.frame(e[,1])
colnames(e) <- "metabolite"
e$group <- "adults"
f <- full_join(b,c, by = "metabolite")
f <- full_join(f,d, by = "metabolite")
f <- full_join(f,e, by = "metabolite")
metabolites <- f[complete.cases(f), ]
metabolites <- metabolites[,1]
metabolites <- droplevels(as.factor(metabolites))

# n in each plot
plot_data <- subset(data, exposure == "bmi")
plot_data <- subset(plot_data, model == "model2")
plot_data <- plot_data[plot_data$metabolite %in% metabolites, ]
plot_data <- subset(plot_data, group == "children")
ratios <- subset(plot_data, subclass == "Ratios")
fa_ratios <- subset(plot_data, subclass == "Fatty acids ratios")
lp_size <- subset(plot_data, subclass == "Lipoprotein particle size")
```

### Data overview
Prior to statistical analysis, the predominantly lipid-based metabolites underwent pre-analysis processing using the `metaboprep` `R` package. In pre-analysis processing of metabolomic data, derived metabolite measures (e.g.,) increased the number of representative metabolites in the dataset, suggesting these derived measures contain novel information. As such, metabolomic data that underwent pre-analysis processing with the inclusion of derived features were used throughout. The total number of individuals with metabolomic data, as well as the number of metabolites available prior to pre-processing is given in Table \@ref(tab:observational-table-ALSPAC-N). \par

```{r observational-table-ALSPAC-N, echo=FALSE}
table <- read.table("data/observational/tables/metabolomics_data_pre_processing.txt", header = T, sep = "\t")
knitr::kable(table, longtable = F, booktabs = T, format = "latex", escape = F,
    caption = 'Metabolomic data available prior to pre-analysis processing', 
    row.names = F,
    col.names = linebreak(c("Age group", "N", "Metabolites", "Clinic", "N", "Unique\n N", "M", "Unique\n M"))) %>%
  collapse_rows(columns = 1:3) %>%
  kable_styling(latex_options = c("striped", "HOLD_position"), position = "center") %>%
  kableExtra::footnote(general = "Data were available from multiple time points. Generation 1 individuals were were measured at 5 time points and combined into three groups (Children, Adolescents, Young adults). Generation 0 individuals were measured at three time points and combined (Adults). N = number of individuals in each combined age group; Metabolites = the number of metabolites measured for each combined age group; Clinic = clinic identifier; Unique N = the number of individuals from that clinic who do not appear in the other clinic for that combined age group; M and Unique M = Number of metabolites measured at each clinic and how many of these were unique to that clinic within the combined age group.", threeparttable = T, general_title = "", footnote_as_chunk = T)
```

Pre-analysis processing of metabolomic data resulted in the following exclusions: `r children_samples_removed` individuals and `r children_features_removed` metabolites were removed from the children's data (4 samples excluded for $\geq$ 80% missingness, 1 sample excluded for total sum abundance $\geq$ 5 SD from the mean, 1 sample excluded as a result of being $\geq$ 5 SD from the mean of PC1 and 2, and 4 metabolites removed for $\geq$ 20% missingness); `r adolescents_samples_removed` individuals and `r adolescents_features_removed` metabolites were removed from the adolescents data (1 sample excluded for total sum abundance $\geq$ 5 SD from the mean and 4 samples excluded as a result of being $\geq$ 5 SD from the mean of PC1 and 2); `r young_adults_samples_removed` individuals and `r young_adults_features_removed` metabolites were removed from the young adults data (1 sample excluded for total sum abundance $\geq$ 5 SD from the mean and 3 samples excluded as a result of being $\geq$ 5 SD from the mean of PC1 and 2); `r adults_samples_removed` individuals and `r adults_features_removed` metabolites were removed from the adults data (1 sample excluded for total sum abundance $\geq$ 5 SD from the mean, 6 samples excluded as a result of being $\geq$ 5 SD from the mean of PC1 and 2, and 4 metabolites removed for $\geq$ 20% missingness). Processed metabolomic data available for statistical analysis is given in Table \@ref(tab:observational-table-ALSPAC-QC-N). \par

```{r observational-table-ALSPAC-QC-N, echo=FALSE}
x <- ALSPAC_QC_N

x[3,1] <- "Young adults"
knitr::kable(x, longtable = F, booktabs = T, format = "latex", escape = F,
             row.names = F,
             col.names = c("Group", "N", "Metabolites"),
             caption = 'Metabolomic data available after pre-analysis processing') %>%
  kable_styling(latex_options = c("striped", "HOLD_position"), position = "center") %>%
  kableExtra::footnote(general = "Number of individuals (N) and metabolites available after pre-analysis processing of metabolomic data was performed using the metaboprep R package and including derived measures.", threeparttable = T, general_title = "", footnote_as_chunk = T)
```

Of individuals with metabolomic data, a majority also had data on measures of adiposity (Table \@ref(tab:observational-table-ALSPAC-adiposity)), except for adolescents where data on WHR was not available. Adiposity data were normally distributed (Appendix Figure \@ref(appendix-observational-figure-adiposity-distribution)). Data on covarialbes were also available in the majority of individuals, the exception being physical activity where fewer individuals had available data and no data was available for children (Table \@ref(tab:observational-table-ALSPAC-adiposity)). \par

<!-- Do you factor in difference in N for covarites in the analyses????-->
<!-- A total of `r nlevels(metabolites)` metabolites were available in all age groups. -->

```{r observational-table-ALSPAC-adiposity, echo=FALSE, warning=FALSE, error=FALSE}
adiposity_table <- read.table("data/observational/tables/adiposity_table.txt", header = T, sep = "\t")

knitr::kable(adiposity_table, longtable = F, booktabs = T, format = "latex", escape = F,
             caption = 'Measures of adiposity available for individuals with metabolomic data',
             row.names = F,
             col.names = c("Age group", "N", "N","mean","SD","N","mean","SD","N","mean","SD")) %>%
  add_header_above(c(" " = 2, "BMI" = 3, "WHR" = 3, "BF" = 3)) %>%
    kable_styling(latex_options = c("striped", "HOLD_position"), position = "center") %>%
  kableExtra::footnote(general = "Measures were obtained for all individuals with processed metabolomic data. Waist hip ratio (WHR) was not available for adolescents. Body fat percentage (BF) was not available for children but a raw impedance measure was available (see methods). Age group = the age group into which clinic visits were combined; BMI = body mass index (kg/m^2^); N = the number of individuals with available data; mean = the mean of the anthropometric measure; SD = standard deviation of the mean; - = data not available.", threeparttable = T, general_title = "", footnote_as_chunk = T)
```

\newpage
\thispagestyle{empty}
\vspace*{-3cm}

```{r observational-table-ALSPAC-covariates, echo=FALSE, warning=FALSE, error=FALSE}
covariate_table <- read.table("data/observational/tables/covariate_table_manual.txt", header = T, sep = "\t", colClasses = "character")

knitr::kable(covariate_table, longtable = T, booktabs = T, format = "latex", escape = F,
             caption = 'Covariales available for individuals with metabolomic data', row.names = F, 
             col.names = c("","","Children","Adolescents","Young adults","Adults")) %>%
  collapse_rows(columns = 1) %>%
  kable_styling(latex_options = c("striped", "HOLD_position"), position = "center") %>%
  kableExtra::footnote(general = "Education is highest level of education (1-5); for children, adolescents, and young adults education is maternal education; for adults, education is own education. Smoking is binary. Alcohol is frequency respondent consumes an alcoholic drink, with 1 being low and 5 high. Diet is predicted kilo-calories consumed per day. Physical activity in adolescents is mean counts per minute of activity across 7 days. For young adults physical activity is the average number of valid minutes per day spent doing moderate to vigorous activity. For adults, physical activity is: no physical activity (1), occasionally (2), frequently (3). N = the number of individuals with available data; mean = the mean of the measure; SD = standard deviation of the mean; - = data not available.", threeparttable = T, general_title = "", footnote_as_chunk = T)
```

When looking at the variability of metabolites across age groups, the mean and SD of the metabolite value, as well as the median and range between the minimum and maximum values for each metabolite were generally similar across all age groups. Variability in metabolite values tended to increase (larger SD of mean metabolite value and range of metabolite value for each metabolite) with age however, with the largest variability predominantly seen in adults. Variability appeared much larger when looking exclusively at derived metabolite measures as opposed to directly measured metabolites. All metabolite distribution plots are available on [GitHub](https://github.com/mattlee821/000_thesis/tree/master/index/data/observational/figures/metabolite_concentrations). \par

### Statistical analysis
```{r echo=FALSE, warning=FALSE, error=FALSE}
library(tidyr)
# how many consistent significnat effects
# main analysis, consistent direction, significant ====
data <- read.table("data/observational/data/analysis/results/combined/combined.txt", header = T, sep ="\t")
data_all <- data
data <- subset(data, model == "model2")
children <- subset(data, group == "children")
adolescents <- subset(data, group == "adolescents")
young_adults <- subset(data, group == "young_adults")
adults <- subset(data, group == "adults")

n_children <- nrow(children)/3
n_adolescents <- nrow(adolescents)/2
n_young_adults <- nrow(young_adults)/3
n_adults <- nrow(adults)/3

## subset for multiple testing threshold ====
children <- subset(children, p <= 0.05/children_independent)
adolescents <- subset(adolescents, p <= 0.05/adolescents_independent)
young_adults <- subset(young_adults, p <= 0.05/young_adults_independent)
adults <- subset(adults, p <= 0.05/adults_independent)

# limit to interested rows ====
children <- children[,c(1,3,13,14,15)]
adolescents <- adolescents[,c(1,3,13,14,15)]
young_adults <- young_adults[,c(1,3,13,14,15)]
adults <- adults[,c(1,3,13,14,15)]

## convert to wide based on model ====
children <- spread(children, exposure, b)
adolescents <- spread(adolescents, exposure, b)
young_adults <- spread(young_adults, exposure, b)
adults <- spread(adults, exposure, b)

### children ====
data <- children[,c(4:ncol(children))]
data <- data[complete.cases(data), ]
data$direction <- sapply(1:nrow(data), function(x) ifelse(all(sign(data[x,]) == 1), 1, ifelse(all(sign(data[x,]) == -1), 2, 0)))
data <- data %>%
  mutate(direction_group = case_when(direction == 1 ~ "Positive effect",
                                     direction == 2 ~ "Negative effect",
                                     direction == 0 ~ "Opposite effect"))
data$direction_group <- factor(data$direction_group,
                               levels = c("Positive effect", "Negative effect", "Opposite effect"),ordered = TRUE)
children <- data
children$group <- "children"
children_consistent_significant <- nrow(children)

### adolescents ====
data <- adolescents[,c(4:ncol(adolescents))]
data <- data[complete.cases(data), ]
data$direction <- sapply(1:nrow(data), function(x) ifelse(all(sign(data[x,]) == 1), 1, ifelse(all(sign(data[x,]) == -1), 2, 0)))
data <- data %>%
  mutate(direction_group = case_when(direction == 1 ~ "Positive effect",
                                     direction == 2 ~ "Negative effect",
                                     direction == 0 ~ "Opposite effect"))
data$direction_group <- factor(data$direction_group,
                               levels = c("Positive effect", "Negative effect", "Opposite effect"),ordered = TRUE)
adolescents <- data
adolescents$group <- "adolescents"
adolescents_consistent_significant <- nrow(adolescents)

### young_adults ====
data <- young_adults[,c(4:ncol(young_adults))]
data <- data[complete.cases(data), ]
data$direction <- sapply(1:nrow(data), function(x) ifelse(all(sign(data[x,]) == 1), 1, ifelse(all(sign(data[x,]) == -1), 2, 0)))
data <- data %>%
  mutate(direction_group = case_when(direction == 1 ~ "Positive effect",
                                     direction == 2 ~ "Negative effect",
                                     direction == 0 ~ "Opposite effect"))
data$direction_group <- factor(data$direction_group,
                               levels = c("Positive effect", "Negative effect", "Opposite effect"),ordered = TRUE)
young_adults <- data
young_adults$group <- "young_adults"
young_adults_consistent_significant <- nrow(young_adults)

### adults ====
data <- adults[,c(4:ncol(adults))]
data <- data[complete.cases(data), ]
data$direction <- sapply(1:nrow(data), function(x) ifelse(all(sign(data[x,]) == 1), 1, ifelse(all(sign(data[x,]) == -1), 2, 0)))
data <- data %>%
  mutate(direction_group = case_when(direction == 1 ~ "Positive effect",
                                     direction == 2 ~ "Negative effect",
                                     direction == 0 ~ "Opposite effect"))
data$direction_group <- factor(data$direction_group,
                               levels = c("Positive effect", "Negative effect", "Opposite effect"),ordered = TRUE)
adults <- data
adults$group <- "adults"
adults_consistent_significant <- nrow(adults)

```

```{r echo=FALSE, warning=FALSE, error=FALSE}
data <- read.table("data/observational/data/analysis/results/combined/combined.txt", header = T, sep ="\t")

# children ====
children_m1_bmi <- subset(data, model == "model1" & group == "children" & exposure == "bmi")
children_m1_bmi_sig <- subset(children_m1_bmi, p < 0.05/children_independent)
children_m2_bmi <- subset(data, model == "model2" & group == "children" & exposure == "bmi")
children_m2_bmi_sig <- subset(children_m2_bmi, p < 0.05/children_independent)
children_m1_whr <- subset(data, model == "model1" & group == "children" & exposure == "whr")
children_m1_whr_sig <- subset(children_m1_whr, p < 0.05/children_independent)
children_m2_whr <- subset(data, model == "model2" & group == "children" & exposure == "whr")
children_m2_whr_sig <- subset(children_m2_whr, p < 0.05/children_independent)
children_m1_bf <- subset(data, model == "model1" & group == "children" & exposure == "bf")
children_m1_bf_sig <- subset(children_m1_bf, p < 0.05/children_independent)
children_m2_bf <- subset(data, model == "model2" & group == "children" & exposure == "bf")
children_m2_bf_sig <- subset(children_m2_bf, p < 0.05/children_independent)

# adolescents ====
adolescents_m1_bmi <- subset(data, model == "model1" & group == "adolescents" & exposure == "bmi")
adolescents_m1_bmi_sig <- subset(adolescents_m1_bmi, p < 0.05/adolescents_independent)
adolescents_m2_bmi <- subset(data, model == "model2" & group == "adolescents" & exposure == "bmi")
adolescents_m2_bmi_sig <- subset(adolescents_m2_bmi, p < 0.05/adolescents_independent)
adolescents_m3_bmi <- subset(data, model == "model3" & group == "adolescents" & exposure == "bmi")
adolescents_m3_bmi_sig <- subset(adolescents_m3_bmi, p < 0.05/adolescents_independent)
adolescents_m1_bf <- subset(data, model == "model1" & group == "adolescents" & exposure == "bf")
adolescents_m1_bf_sig <- subset(adolescents_m1_bf, p < 0.05/adolescents_independent)
adolescents_m2_bf <- subset(data, model == "model2" & group == "adolescents" & exposure == "bf")
adolescents_m2_bf_sig <- subset(adolescents_m2_bf, p < 0.05/adolescents_independent)
adolescents_m3_bf <- subset(data, model == "model3" & group == "adolescents" & exposure == "bf")
adolescents_m3_bf_sig <- subset(adolescents_m3_bf, p < 0.05/adolescents_independent)

# young_adults ====
young_adults_m1_bmi <- subset(data, model == "model1" & group == "young_adults" & exposure == "bmi")
young_adults_m1_bmi_sig <- subset(young_adults_m1_bmi, p < 0.05/young_adults_independent)
young_adults_m2_bmi <- subset(data, model == "model2" & group == "young_adults" & exposure == "bmi")
young_adults_m2_bmi_sig <- subset(young_adults_m2_bmi, p < 0.05/young_adults_independent)
young_adults_m3_bmi <- subset(data, model == "model3" & group == "young_adults" & exposure == "bmi")
young_adults_m3_bmi_sig <- subset(young_adults_m3_bmi, p < 0.05/young_adults_independent)
young_adults_m1_whr <- subset(data, model == "model1" & group == "young_adults" & exposure == "whr")
young_adults_m1_whr_sig <- subset(young_adults_m1_bmi, p < 0.05/young_adults_independent)
young_adults_m2_whr <- subset(data, model == "model2" & group == "young_adults" & exposure == "whr")
young_adults_m2_whr_sig <- subset(young_adults_m2_bmi, p < 0.05/young_adults_independent)
young_adults_m3_whr <- subset(data, model == "model3" & group == "young_adults" & exposure == "whr")
young_adults_m3_whr_sig <- subset(young_adults_m3_bmi, p < 0.05/young_adults_independent)
young_adults_m1_bf <- subset(data, model == "model1" & group == "young_adults" & exposure == "bf")
young_adults_m1_bf_sig <- subset(young_adults_m1_bf, p < 0.05/young_adults_independent)
young_adults_m2_bf <- subset(data, model == "model2" & group == "young_adults" & exposure == "bf")
young_adults_m2_bf_sig <- subset(young_adults_m2_bf, p < 0.05/young_adults_independent)
young_adults_m3_bf <- subset(data, model == "model3" & group == "young_adults" & exposure == "bf")
young_adults_m3_bf_sig <- subset(young_adults_m3_bf, p < 0.05/young_adults_independent)

# adults ====
adults_m1_bmi <- subset(data, model == "model1" & group == "adults" & exposure == "bmi")
adults_m1_bmi_sig <- subset(adults_m1_bmi, p < 0.05/adults_independent)
adults_m2_bmi <- subset(data, model == "model2" & group == "adults" & exposure == "bmi")
adults_m2_bmi_sig <- subset(adults_m2_bmi, p < 0.05/adults_independent)
adults_m3_bmi <- subset(data, model == "model3" & group == "adults" & exposure == "bmi")
adults_m3_bmi_sig <- subset(adults_m3_bmi, p < 0.05/adults_independent)
adults_m1_whr <- subset(data, model == "model1" & group == "adults" & exposure == "whr")
adults_m1_whr_sig <- subset(adults_m1_bmi, p < 0.05/adults_independent)
adults_m2_whr <- subset(data, model == "model2" & group == "adults" & exposure == "whr")
adults_m2_whr_sig <- subset(adults_m2_bmi, p < 0.05/adults_independent)
adults_m3_whr <- subset(data, model == "model3" & group == "adults" & exposure == "whr")
adults_m3_whr_sig <- subset(adults_m3_bmi, p < 0.05/adults_independent)
adults_m1_bf <- subset(data, model == "model1" & group == "adults" & exposure == "bf")
adults_m1_bf_sig <- subset(adults_m1_bf, p < 0.05/adults_independent)
adults_m2_bf <- subset(data, model == "model2" & group == "adults" & exposure == "bf")
adults_m2_bf_sig <- subset(adults_m2_bf, p < 0.05/adults_independent)
adults_m3_bf <- subset(data, model == "model3" & group == "adults" & exposure == "bf")
adults_m3_bf_sig <- subset(adults_m3_bf, p < 0.05/adults_independent)

# precent of total ====
children_m1_bmi_percent <- nrow(children_m1_bmi_sig) / nrow(children_m1_bmi) * 100
children_m2_bmi_percent <- nrow(children_m2_bmi_sig) / nrow(children_m2_bmi) * 100
children_m1_whr_percent <- nrow(children_m1_whr_sig) / nrow(children_m1_whr) * 100
children_m2_whr_percent <- nrow(children_m2_whr_sig) / nrow(children_m2_whr) * 100
children_m1_bf_percent <- nrow(children_m1_bf_sig) / nrow(children_m1_bf) * 100
children_m2_bf_percent <- nrow(children_m2_bf_sig) / nrow(children_m2_bf) * 100
adolescents_m1_bmi_percent <- nrow(adolescents_m1_bmi_sig) / nrow(adolescents_m1_bmi) * 100
adolescents_m2_bmi_percent <- nrow(adolescents_m2_bmi_sig) / nrow(adolescents_m2_bmi) * 100
adolescents_m3_bmi_percent <- nrow(adolescents_m3_bmi_sig) / nrow(adolescents_m3_bmi) * 100
adolescents_m1_bf_percent <- nrow(adolescents_m1_bf_sig) / nrow(adolescents_m1_bf) * 100
adolescents_m2_bf_percent <- nrow(adolescents_m2_bf_sig) / nrow(adolescents_m2_bf) * 100
adolescents_m3_bf_percent <- nrow(adolescents_m3_bf_sig) / nrow(adolescents_m3_bf) * 100
young_adults_m1_bmi_percent <- nrow(young_adults_m1_bmi_sig) / nrow(young_adults_m1_bmi) * 100
young_adults_m2_bmi_percent <- nrow(young_adults_m2_bmi_sig) / nrow(young_adults_m2_bmi) * 100
young_adults_m3_bmi_percent <- nrow(young_adults_m3_bmi_sig) / nrow(young_adults_m3_bmi) * 100
young_adults_m1_whr_percent <- nrow(young_adults_m1_whr_sig) / nrow(young_adults_m1_whr) * 100
young_adults_m2_whr_percent <- nrow(young_adults_m2_whr_sig) / nrow(young_adults_m2_whr) * 100
young_adults_m3_whr_percent <- nrow(young_adults_m3_whr_sig) / nrow(young_adults_m3_whr) * 100
young_adults_m1_bf_percent <- nrow(young_adults_m1_bf_sig) / nrow(young_adults_m1_bf) * 100
young_adults_m2_bf_percent <- nrow(young_adults_m2_bf_sig) / nrow(young_adults_m2_bf) * 100
young_adults_m3_bf_percent <- nrow(young_adults_m3_bf_sig) / nrow(young_adults_m3_bf) * 100
adults_m1_bmi_percent <- nrow(adults_m1_bmi_sig) / nrow(adults_m1_bmi) * 100
adults_m2_bmi_percent <- nrow(adults_m2_bmi_sig) / nrow(adults_m2_bmi) * 100
adults_m3_bmi_percent <- nrow(adults_m3_bmi_sig) / nrow(adults_m3_bmi) * 100
adults_m1_whr_percent <- nrow(adults_m1_whr_sig) / nrow(adults_m1_whr) * 100
adults_m2_whr_percent <- nrow(adults_m2_whr_sig) / nrow(adults_m2_whr) * 100
adults_m3_whr_percent <- nrow(adults_m3_whr_sig) / nrow(adults_m3_whr) * 100
adults_m1_bf_percent <- nrow(adults_m1_bf_sig) / nrow(adults_m1_bf) * 100
adults_m2_bf_percent <- nrow(adults_m2_bf_sig) / nrow(adults_m2_bf) * 100
adults_m3_bf_percent <- nrow(adults_m3_bf_sig) / nrow(adults_m3_bf) * 100

min_significant_percent <- round(min(children_m1_bmi_percent,children_m2_bmi_percent,children_m1_whr_percent,children_m2_whr_percent,children_m2_bf_percent,
    adolescents_m1_bmi_percent,adolescents_m2_bmi_percent,adolescents_m3_bmi_percent,adolescents_m1_bf_percent,adolescents_m2_bf_percent,adolescents_m3_bf_percent,
    young_adults_m1_bmi_percent,young_adults_m2_bmi_percent,young_adults_m3_bmi_percent,young_adults_m1_whr_percent,young_adults_m2_whr_percent,young_adults_m3_whr_percent,young_adults_m1_bf_percent,young_adults_m2_bf_percent,young_adults_m3_bf_percent,
    adults_m1_bmi_percent,adults_m2_bmi_percent,adults_m3_bmi_percent,adults_m1_whr_percent,adults_m2_whr_percent,adults_m3_whr_percent,adults_m1_bf_percent,adults_m2_bf_percent,adults_m3_bf_percent),2)

max_significant_percent <- round(max(children_m1_bmi_percent,children_m2_bmi_percent,children_m1_whr_percent,children_m2_whr_percent,children_m2_bf_percent,
    adolescents_m1_bmi_percent,adolescents_m2_bmi_percent,adolescents_m3_bmi_percent,adolescents_m1_bf_percent,adolescents_m2_bf_percent,adolescents_m3_bf_percent,
    young_adults_m1_bmi_percent,young_adults_m2_bmi_percent,young_adults_m3_bmi_percent,young_adults_m1_whr_percent,young_adults_m2_whr_percent,young_adults_m3_whr_percent,young_adults_m1_bf_percent,young_adults_m2_bf_percent,young_adults_m3_bf_percent,
    adults_m1_bmi_percent,adults_m2_bmi_percent,adults_m3_bmi_percent,adults_m1_whr_percent,adults_m2_whr_percent,adults_m3_whr_percent,adults_m1_bf_percent,adults_m2_bf_percent,adults_m3_bf_percent),2)

mean_significnat_percent <- round(mean(children_m1_bmi_percent,children_m2_bmi_percent,children_m1_whr_percent,children_m2_whr_percent,children_m2_bf_percent,
    adolescents_m1_bmi_percent,adolescents_m2_bmi_percent,adolescents_m3_bmi_percent,adolescents_m1_bf_percent,adolescents_m2_bf_percent,adolescents_m3_bf_percent,
    young_adults_m1_bmi_percent,young_adults_m2_bmi_percent,young_adults_m3_bmi_percent,young_adults_m1_whr_percent,young_adults_m2_whr_percent,young_adults_m3_whr_percent,young_adults_m1_bf_percent,young_adults_m2_bf_percent,young_adults_m3_bf_percent,
    adults_m1_bmi_percent,adults_m2_bmi_percent,adults_m3_bmi_percent,adults_m1_whr_percent,adults_m2_whr_percent,adults_m3_whr_percent,adults_m1_bf_percent,adults_m2_bf_percent,adults_m3_bf_percent),2)

median_significant_percent <- round(median(children_m1_bmi_percent,children_m2_bmi_percent,children_m1_whr_percent,children_m2_whr_percent,children_m2_bf_percent,
    adolescents_m1_bmi_percent,adolescents_m2_bmi_percent,adolescents_m3_bmi_percent,adolescents_m1_bf_percent,adolescents_m2_bf_percent,adolescents_m3_bf_percent,
    young_adults_m1_bmi_percent,young_adults_m2_bmi_percent,young_adults_m3_bmi_percent,young_adults_m1_whr_percent,young_adults_m2_whr_percent,young_adults_m3_whr_percent,young_adults_m1_bf_percent,young_adults_m2_bf_percent,young_adults_m3_bf_percent,
    adults_m1_bmi_percent,adults_m2_bmi_percent,adults_m3_bmi_percent,adults_m1_whr_percent,adults_m2_whr_percent,adults_m3_whr_percent,adults_m1_bf_percent,adults_m2_bf_percent,adults_m3_bf_percent),2)

# percent change ====
adolescents_m1_m3_bmi_percent_change <- (nrow(adolescents_m1_bmi_sig) - nrow(adolescents_m3_bmi_sig)) / nrow(adolescents_m1_bmi_sig) * 100


```

```{r echo=FALSE, warning=FALSE, error=FALSE}
data <- read.table("data/observational/data/analysis/results/combined/combined.txt", header = T, sep ="\t")
m1 <- subset(data, model == "model1")
m2 <- subset(data, model == "model2")
m3 <- subset(data, model == "model3")

children_m1 <- subset(m1, group == "children")
children_m2 <- subset(m2, group == "children")
children_m3 <- subset(m3, group == "children")

adolescents_m1 <- subset(m1, group == "adolescents")
adolescents_m2 <- subset(m2, group == "adolescents")
adolescents_m3 <- subset(m3, group == "adolescents")

young_adults_m1 <- subset(m1, group == "young_adults")
young_adults_m2 <- subset(m2, group == "young_adults")
young_adults_m3 <- subset(m3, group == "young_adults")

adults_m1 <- subset(m1, group == "adults")
adults_m2 <- subset(m2, group == "adults")
adults_m3 <- subset(m3, group == "adults")

table <- data.frame(group = c("Children", "Adolescents", "Young adults", "Adults"),
                    N = c(5650,4484,3265,6399),
                    m1_m2 = c(unique(children_m1$n), unique(adolescents_m1$n), unique(young_adults_m1$n), unique(adults_m1$n)),
                    m3 = c("--", unique(adolescents_m3$n), unique(young_adults_m3$n), unique(adults_m3$n)))
```

In total, metabolomic data were available for 5,650 children, 4,484 adolescents, 3,265 young adults, and 6,399 adults. After obtaining data on adiposity measured and covarialbes data were available for between `r min(data$n)`--`r max(data$n)` individuals (Table \@ref(tab:observational-table-N)). Model 1 and model 2 were restricted to all individuals with complete data for adiposity and covarialbe data for model 2. Model 3 was restricted to individuals with complete data for all adiposity measures and all covarialbes. \par

```{r observational-table-N, echo=FALSE, warning=FALSE, error=FALSE}
data <- read.table("data/observational/data/analysis/results/combined/combined.txt", header = T, sep ="\t")
m1 <- subset(data, model == "model1")
m2 <- subset(data, model == "model2")
m3 <- subset(data, model == "model3")

children_m1 <- subset(m1, group == "children")
children_m2 <- subset(m2, group == "children")
children_m3 <- subset(m3, group == "children")

adolescents_m1 <- subset(m1, group == "adolescents")
adolescents_m2 <- subset(m2, group == "adolescents")
adolescents_m3 <- subset(m3, group == "adolescents")

young_adults_m1 <- subset(m1, group == "young_adults")
young_adults_m2 <- subset(m2, group == "young_adults")
young_adults_m3 <- subset(m3, group == "young_adults")

adults_m1 <- subset(m1, group == "adults")
adults_m2 <- subset(m2, group == "adults")
adults_m3 <- subset(m3, group == "adults")

table <- data.frame(group = c("Children", "Adolescents", "Young adults", "Adults"),
                    N = c(5650,4484,3265,6399),
                    m1_m2 = c(unique(children_m1$n), unique(adolescents_m1$n), unique(young_adults_m1$n), unique(adults_m1$n)),
                    m3 = c("--", unique(adolescents_m3$n), unique(young_adults_m3$n), unique(adults_m3$n)))

knitr::kable(table, longtable = T, booktabs = T, format = "latex", escape = T,
             caption = 'Total number of individulas with metabolomic data and total number of individuals included in each linear model', row.names = F, 
             col.names = c("Age group","N","Model 1 & 2","Model 3")) %>%
  kable_styling(latex_options = c("striped", "HOLD_position"), position = "center") %>%
  kableExtra::footnote(general = "The total number of individuals with available metabolomic data (N) and the sample size for each linear model (Model 1 & 2 and Model 3).", threeparttable = T, general_title = "", footnote_as_chunk = T)
```

Across all models and exposures, between `r min_significant_percent`--`r max_significant_percent` % (median = `r round(median_significant_percent,1)` %) of metabolites reached a p-value multiple testing threshold (multiple testing threshold for children = `r round(0.05/children_independent,4)`, adolescents = `r round(0.05/adolescents_independent,4)`, young adults = `r round(0.05/young_adults_independent,4)`, adults = `r round(0.05/adults_independent,4)`). Multiple testing thresholds, calculated during pre-analysis processing using `metaboprep`, were calculated as the number of independent metabolites within the raw metabolomic data given a Spearman’s rho of approximately 0.75. In summary, between 83-193, 139-193, and 75-197 metabolites reached a multiple testing threshold across all age groups for BMI, WHR, and BF respectively (Table \@ref(tab:observational-table-significant-results)).  \par

```{r observational-table-significant-results, echo=FALSE, warning=FALSE, error=FALSE}
data <- read.table("data/observational/tables/multiple_testing_table.txt", header = T, sep ="\t")

knitr::kable(data, longtable = T, booktabs = T, format = "latex", escape = F,
             caption = 'Number of tests reaching a multiple testing threshold', row.names = F, 
             col.names = c("","N","1","2","3","1","2","3","1","2","3")) %>%
  collapse_rows(columns = 1) %>%
  kable_styling(latex_options = c("striped", "HOLD_position"), position = "center") %>%
  add_header_above(c(" " = 2, "BMI" = 3, "WHR" = 3, "BF" = 3)) %>%
  kableExtra::footnote(general = "Number of metabolites (out of total N) reaching a multiple testing threshold for each model (1-3) within each age group for each exposure. Multiple testing thresholds: children = 0.0012, adolescents = 0.0012, young adults = 0.0013, adults = 0.0011. BMI = body mass index; WHR = waist hip ratio; BF = body fat percentage; 1 = model 1, adjustment for age and sex; 2 = model 2, adjustment for model 1 plus maternal/own education, smoking status, alcohol frequency, diet (where available); 3 = model 3, adjustment for model 2 plus physical activity (where available).", threeparttable = T, general_title = "", footnote_as_chunk = T)
```

#### Directional consistency between effect estimates across three linear models within exposures and age groups
Across models, within each exposure and age group, the majority of tests resulted in directionally consistent effect estimates (Figure \@ref(fig:observational-figure-directional-consistency)). Of these directionally consistent effect estimates, the majority of effect estimates were positive. That is, adiposity was associated with an increase in a majority of metabolites. For children, effect sizes and CIs were broadly consistent across models 1 and 2. For adolescents, young adults, and adults, effect sizes and CIS were broadly consistent across across all three models, though there was some attenuation of effect size and wider CIs for model 3 (See forest plots on GitHub for [children](https://github.com/mattlee821/000_thesis/blob/master/index/data/observational/appendix/forestplot_main_children.pdf), [adolescents](https://github.com/mattlee821/000_thesis/blob/master/index/data/observational/appendix/forestplot_main_adolescents.pdf), [young adults](https://github.com/mattlee821/000_thesis/blob/master/index/data/observational/appendix/forestplot_main_young_adults.pdf), and [adults](https://github.com/mattlee821/000_thesis/blob/master/index/data/observational/appendix/forestplot_main_adults.pdf)). Across the majority of metabolites, CIs overlapped for all exposures within age groups (Appendix \@ref(appendix-observational-figures-forestplots)). Of the `r nlevels(metabolites)` metabolites measured in all age groups, directional consistency was supported by strong evidence of correlation across models within exposures and age groups (Appendix \@ref(appendix-observational-correlations)). \par

(ref:observational-figure-directional-consistency-cap) **Directional consistency between effect estimates across three linear models within exposures and age groups**. A positive effect reflects all model betas being in the positive direction; a negative effect reflects all model betas being in a negative direction; opposite effect reflects different directions for the model betas. BMI = body mass index; WHR = waist hip ratio; BF = body fat percentage.

(ref:observational-figure-directional-consistency-scap) Directional consistency between effect estimates across three linear models within exposures and age groups

```{r observational-figure-directional-consistency, eval=TRUE, include=TRUE, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE, out.width='100%', fig.cap='(ref:observational-figure-directional-consistency-cap)', fig.scap='(ref:observational-figure-directional-consistency-scap)'}
knitr::include_graphics("data/observational/figures/plot_directional_consistency_across_models_within_exposures_groups.pdf")
```

#### Directional consistency between effect estimates for linear model 2 across exposures within age groups
```{r echo=FALSE, warning=FALSE, error=FALSE}
# packages
library(tidyr)
library(dplyr)
library(ggplot2)
library(cowplot)
source("data/index/colour_palette.R")
# data ==== 
data <- read.table("data/observational/data/analysis/results/combined/combined.txt", header = T, sep ="\t")
data <- data[,c(1,3,14,15,16)]
data <- subset(data, model == "model2")
children <- subset(data, group == "children")
adolescents <- subset(data, group == "adolescents")
young_adults <- subset(data, group == "young_adults")
adults <- subset(data, group == "adults")

## convert to wide based on model ====
children <- spread(children, exposure, b)
adolescents <- spread(adolescents, exposure, b)
young_adults <- spread(young_adults, exposure, b)
adults <- spread(adults, exposure, b)

## assign values for directions consistent across non-clumped and clumped data ====
### children ====
data <- children[,c(4:ncol(children))]
data$direction <- sapply(1:nrow(data), function(x) ifelse(all(sign(data[x,]) == 1), 1, ifelse(all(sign(data[x,]) == -1), 2, 0)))
data <- data %>%
  mutate(direction_group = case_when(direction == 1 ~ "Positive effect",
                               direction == 2 ~ "Negative effect",
                               direction == 0 ~ "Opposite effect"))
data$direction_group <- factor(data$direction_group,
                            levels = c("Positive effect", "Negative effect", "Opposite effect"),ordered = TRUE)
children <- data
children$exposure <- "children"
# ### children inverse
# data <- children[,c(4:ncol(children))]
# data <- data[complete.cases(data), ]
# data$direction = sapply( 1:nrow(data), function(i){
# 	## set data vector
# 	x = sign( data[i, ] )
# 	##
# 	if( x[1] == -1 & sum(x[2:3]) == 2 ){
# 		direct = 1
# 	} else{
# 		if( x[1] == 1 & sum(x[2:3]) == -2 ){
# 		direct = 2
# 		} else {
# 			direct = 3
# 		}
# 	}
# 	}
# )
# data <- data %>%
#   mutate(direction_group = case_when(direction == 1 ~ "Positive effect",
#                                      direction == 2 ~ "Negative effect",
#                                      direction == 3 ~ "Opposite effect"))
# data$direction_group <- factor(data$direction_group,
#                                levels = c("Positive effect", "Negative effect", "Opposite effect"),ordered = TRUE)
# children_inverse <- data
# children_inverse$exposure <- "children_inverse"

### adolescents ====
data <- adolescents[,c(4:ncol(adolescents))]
data$direction <- sapply(1:nrow(data), function(x) ifelse(all(sign(data[x,]) == 1), 1, ifelse(all(sign(data[x,]) == -1), 2, 0)))
data <- data %>%
  mutate(direction_group = case_when(direction == 1 ~ "Positive effect",
                               direction == 2 ~ "Negative effect",
                               direction == 0 ~ "Opposite effect"))
data$direction_group <- factor(data$direction_group,
                            levels = c("Positive effect", "Negative effect", "Opposite effect"),ordered = TRUE)
adolescents <- data
adolescents$exposure <- "adolescents"
### young_adults ====
data <- young_adults[,c(4:ncol(young_adults))]
data$direction <- sapply(1:nrow(data), function(x) ifelse(all(sign(data[x,]) == 1), 1, ifelse(all(sign(data[x,]) == -1), 2, 0)))
data <- data %>%
  mutate(direction_group = case_when(direction == 1 ~ "Positive effect",
                               direction == 2 ~ "Negative effect",
                               direction == 0 ~ "Opposite effect"))
data$direction_group <- factor(data$direction_group,
                            levels = c("Positive effect", "Negative effect", "Opposite effect"),ordered = TRUE)
young_adults <- data
young_adults$exposure <- "young_adults"
### adults ====
data <- adults[,c(4:ncol(adults))]
data$direction <- sapply(1:nrow(data), function(x) ifelse(all(sign(data[x,]) == 1), 1, ifelse(all(sign(data[x,]) == -1), 2, 0)))
data <- data %>%
  mutate(direction_group = case_when(direction == 1 ~ "Positive effect",
                               direction == 2 ~ "Negative effect",
                               direction == 0 ~ "Opposite effect"))
data$direction_group <- factor(data$direction_group,
                            levels = c("Positive effect", "Negative effect", "Opposite effect"),ordered = TRUE)
adults <- data
adults$exposure <- "adults"
## join data frames ====
data <- bind_rows(children,adolescents,young_adults,adults)
data$exposure <- as.factor(data$exposure)
data$exposure <- factor(data$exposure ,levels(data$exposure)[c(3,1,4,2)])
```

Given the broad agreement between models, including the high correlation observed for effects between model 2 and model 3 (Spearmans rho = 0.81--0.995; Appendix \@ref(appendix-observational-correlations)), and the fact that model 3 was not run for children as data on physical activity were not available, results here on are presented for model 2 (all data are available on [GitHub](https://github.com/mattlee821/000_thesis/tree/master/index/data/observational)). \par

Across the three exposures (BMI, WHR, and BF) within each age group, effects showed mostly consistent directions of effect, the majority of which were positive (i.e., adiposity was associated with an increase in metabolites; Figure \@ref(fig:observational-figure-directional-consistency-2)). Where there were the most inconsistent directions across exposures was for children, with `r round(table(children$direction_group)[[3]]/(nrow(children))*100)`% of metabolites showed inconsistent directions of effect. For adolescents, young adults, and adults `r round(table(adolescents$direction_group)[[3]]/(nrow(adolescents))*100)`%, `r round(table(young_adults$direction_group)[[3]]/(nrow(young_adults))*100)`%, and `r round(table(adults$direction_group)[[3]]/(nrow(adults))*100)`% of metabolites showed inconsistent directions of effect across measures of adiposity, respectively. A large proportion of metabolites with evidence for a consistent direction of effect across all three exposures also reached the multiple testing threshold for all three adiposity measures for model 2: children = `r children_consistent_significant` out of 230 metabolites, adolescents = `r adolescents_consistent_significant` out of 230 metabolites, young adults = `r young_adults_consistent_significant` out of 224 metabolites, and adults = `r adults_consistent_significant` out of 228 metabolites. \par

Of the `r nlevels(metabolites)` metabolites measured across all age groups, directional consistency was supported by strong evidence of correlation across exposures within age groups (Spearmans rho = 0.61--0.97) and within exposures across age groups (Spearmans rho = 0.48--0.82; Appendix \@ref(appendix-observational-correlations)). \par

(ref:observational-figure-directional-consistency-2-cap) **Directional consistency for linear model 2 across exposures within age groups**. A positive effect reflects all model effect estimates being in the positive direction; a negative effect reflects all model effect estimates being in a negative direction; opposite effect reflects different directions for the model effect estimates. BMI = body mass index; WHR = waist hip ratio; BF = body fat percentage.

(ref:observational-figure-directional-consistency-2-scap) Directional consistency for linear model 2 across exposures within age groups

```{r observational-figure-directional-consistency-2, eval=TRUE, include=TRUE, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE, out.width='100%', fig.cap='(ref:observational-figure-directional-consistency-2-cap)', fig.scap='(ref:observational-figure-directional-consistency-2-scap)'}
knitr::include_graphics("data/observational/figures/plot_directional_consistency_model2_within_exposures_groups.pdf")
```

#### Global metabolic profile
To aid visual comparison between metabolite subclasses, derived measures were visualised separately to non-derived measures, and are presented in the Appendix (\@ref(appendix-observational-figures-circosplots)). Derived metabolites are metabolites that are not directly measured by the metabolomic array and are instead derived during the processing of the raw NMR spectra. Derived measures include ratios such as cholesterol esters in very large very low density lipoprotein (VLDL) to total lipids in very large VLDL ratio. \par

Overall, the global pattern of association was very similar for all measures of adiposity for children (Figure \@ref(fig:observational-figure-circosplot-main-children)), adolescents (Figure \@ref(fig:observational-figure-circosplot-main-adolescents)), young adults (Figure \@ref(fig:observational-figure-circosplot-main-young-adults)), and adults (Figure \@ref(fig:observational-figure-circosplot-main-adults)) across all directly measured metabolites (larger figures are available on GitHub, for [children](https://github.com/mattlee821/000_thesis/blob/master/index/data/observational/figures/circosplot_main_children.pdf), [adolescents](https://github.com/mattlee821/000_thesis/blob/master/index/data/observational/figures/circosplot_main_adolescents.pdf), [young adults](https://github.com/mattlee821/000_thesis/blob/master/index/data/observational/figures/circosplot_main_young_adults.pdf), and [adults](https://github.com/mattlee821/000_thesis/blob/master/index/data/observational/figures/circosplot_main_adults.pdf)). Across all age groups, the largest effect sizes for directly measured metabolites were found for the fatty acids subclass; the total fatty acids metabolite showed the largest effect size across all exposures for each age group. Metabolites in small VLDL, medium VLDL, large VLDL, and very large VLDL subclasses were the only metabolites to reach the specified multiple testing thresholds across all exposures and age groups. On the whole, effect sizes were lowest in children and increased with age. The largest effect sizes in adults, total fatty acids (change in metabolite (mmol/l) per SD change in the exposure = 0.51), was twice that observed in children (change in metabolite (mmol/l) per SD change in the exposure = 0.21). \par

For derived measures, the global pattern of association was very similar within age groups across exposures and across age groups within exposures (Appendix \@ref(appendix-observational-figures-circosplots)). There was considerable variation within subclasses for derived measures. For adults (Figure \@ref(appendix-observational-figure-circos-adults)), unlike the other age groups, extreme effect sizes were found for a number of metabolites (summary of effect size across all exposures and derived metabolites: minimum = -1.18 x 10^-15^; median = -0.11; mean = -2.09 x 10^-13^; maximum = 1.95 x 10^11^). The majority of effect sizes observed for adults were around the median value. \par

(ref:observational-figure-circosplot-main-children-cap) **Effect estimates from linear regression of adiposity measures on metabolites in children**. The Circos plot shows each track as one of the measures of adiposity; the outer track is body mass index (BMI), the middle track is waist hip ratio (WHR), the inner track is body fat percentage (BF). Effect estimates are given as a change in the raw metabolite unit per standard deviation change in the exposure; 95% confidence intervals are shown and may be hidden by the point estimate if very tight. Solid points indicate a multiple testing threshold has been reached (0.05/`r children_independent`).

(ref:observational-figure-circosplot-main-children-scap) Effect estimates from linear regression of adiposity measures on metabolites in children

```{r observational-figure-circosplot-main-children, echo=FALSE, out.width='100%', fig.cap='(ref:observational-figure-circosplot-main-children-cap)', fig.scap='(ref:observational-figure-circosplot-main-children-scap)'}
knitr::include_graphics("data/observational/figures/circosplot_metabolites_children.pdf")
```

\newpage

(ref:observational-figure-circosplot-main-adolescents-cap) **Effect estimates from linear regression of adiposity measures on metabolites in adolescents**. The Circos plot shows each track as one of the measures of adiposity; the outer track is body mass index (BMI), the inner track is body fat percentage (BF). Effect estimates are given as a change in the raw metabolite unit per standard deviation change in the exposure; 95% confidence intervals are shown and may be hidden by the point estimate if very tight. Solid points indicate a multiple testing threshold has been reached (0.05/`r adolescents_independent`).

(ref:observational-figure-circosplot-main-adolescents-scap) Effect estimates from linear regression of adiposity measures on metabolites in adolescents

```{r observational-figure-circosplot-main-adolescents, echo=FALSE, out.width='100%', fig.cap='(ref:observational-figure-circosplot-main-adolescents-cap)', fig.scap='(ref:observational-figure-circosplot-main-adolescents-scap)'}
knitr::include_graphics("data/observational/figures/circosplot_metabolites_adolescents.pdf")
```

\newpage

(ref:observational-figure-circosplot-main-young-adults-cap) **Effect estimates from linear regression of adiposity measures on metabolites in young adults**. The Circos plot shows each track as one of the measures of adiposity; the outer track is body mass index (BMI), the middle track is waist hip ratio (WHR), the inner track is body fat percentage (BF). Effect estimates are given as a change in the raw metabolite unit per standard deviation change in the exposure; 95% confidence intervals are shown and may be hidden by the point estimate if very tight. Solid points indicate a multiple testing threshold has been reached (0.05/`r young_adults_independent`).

(ref:observational-figure-circosplot-main-young-adults-scap) Effect estimates from linear regression of adiposity measures on metabolites in young adults

```{r observational-figure-circosplot-main-young-adults, echo=FALSE, out.width='100%', fig.cap='(ref:observational-figure-circosplot-main-young-adults-cap)', fig.scap='(ref:observational-figure-circosplot-main-young-adults-scap)'}
knitr::include_graphics("data/observational/figures/circosplot_metabolites_young_adults.pdf")
```

\newpage

(ref:observational-figure-circosplot-main-adults-cap) **Effect estimates from linear regression of adiposity measures on metabolites in adults**. The Circos plot shows each track as one of the measures of adiposity; the outer track is body mass index (BMI), the middle track is waist hip ratio (WHR), the inner track is body fat percentage (BF). Effect estimates are given as a change in the raw metabolite unit per standard deviation change in the exposure; 95% confidence intervals are shown and may be hidden by the point estimate if very tight. Solid points indicate a multiple testing threshold has been reached (0.05/`r adults_independent`).

(ref:observational-figure-circosplot-main-adults-scap) Effect estimates from linear regression of adiposity measures on metabolites in adults

```{r observational-figure-circosplot-main-adults, echo=FALSE, out.width='100%', fig.cap='(ref:observational-figure-circosplot-main-adults-cap)', fig.scap='(ref:observational-figure-circosplot-main-adults-scap)'}
knitr::include_graphics("data/observational/figures/circosplot_metabolites_adults.pdf")
```

\newpage

#### Subclass results
When looking at directly measured metabolite subclasses, there were associations between measures of adiposity and all subclasses except for large low density lipoprotein (LDL) in children, where CIs for all large LDL metabolites crossed the null. For the derived measure subclasses, associations were observed across measures of adiposity for all subclasses except extremely large VLDL ratios in young adults and adults. \par

Across all age groups and exposures, associations with every metabolite in a particular subclass was observed for small VLDL, medium VLDL, large VLDL, very large VLDL, and extremely large VLDL. As age increased, the number of associations within subclasses tended to increase across all measures of adiposity. For example, across small LDL, medium LDL, and large LDL, few associations were observed across measures of adiposity in children, however, in adolescents and young adults, a majority of metabolites showed evidence of association; while in adults, all metabolites showed evidence of association. Effect sizes tended to increase with age as well, for example, the largest effects seen for small, medium, and large LDL metabolites were in adults and young adults (Figure \@ref(fig:observational-figure-forestplot-subclass-LDL)). \par

(ref:observational-figure-forestplot-subclass-LDL-cap) **Effect estimates from linear regression using model 2 of adiposity measures on metabolites in all age groups for small, medium, and large LDL subclasses**. Effect estimates are given as a change in the raw metabolite unit per standard deviation change in the exposure. 95% confidence intervals are shown. BMI = body mass index; WHR = waist hip ratio; BF = body fat percentage.

(ref:observational-figure-forestplot-subclass-LDL-scap) Effect estimates from linear regression using model 2 of adiposity measures on metabolites in all age groups for small, medium, and large LDL subclasses 

```{r observational-figure-forestplot-subclass-LDL, echo=FALSE, out.width='100%',  fig.cap='(ref:observational-figure-forestplot-subclass-LDL-cap)', fig.scap='(ref:observational-figure-forestplot-subclass-LDL-scap)'}
knitr::include_graphics("data/observational/figures/forestplot_subclass_LDL.pdf")
```

To summarise the results for directly measured metabolites from model 2 across exposures and across age groups, the largest mean positive effect estimate were observed for metabolites in the lipoprotein particle size, fatty acids, and inflammation subclasses. The largest mean negative effect estimate were observed for metabolites in the medium, large, and very large HDL, and ketone bodies subclasses. The largest median effect estimates however were observed for the fatty acids, inflammation, and cholesterol subclasses, while the largest negative median effect estimates were observed for medium, large, and very large HDL, and ketone bodies and lipoprotein particle size subclasses. A summary (minimum, maximum, mean, and median values) for each age group across subclasses for model 2 are available on [GitHub](https://github.com/mattlee821/000_thesis/tree/master/index/data/observational/tables/effect_size_summary). \par

The above summary does not take account of the variation within subclasses. That is, metabolites within a subclass do not all have the same direction of effect. Focussing on results for adults given the directional consistency observed across age groups, all metabolites in the very small, small, medium, large, very large, and extremely large VLDL, and IDL, small, medium, and large LDL, and small HDL, fatty acids, inflammation, and branched-chain and aromatic amino acids subclasses had positive directions of effect (Figure on [GitHub](https://github.com/mattlee821/000_thesis/blob/master/index/data/observational/figures/adults_model2_effect_summary.pdf)). A majority of metabolites within the cholesterol, apolipoproteins, glycolysis related metabolites, and glycerides and phospholipids subclasses had positive directions of effect. All metabolites within the ketone bodies, and large and very large HDL subclasses had negative directions of effect. A majority of metabolites within the medium HDL, lipoprotein particle size, and amino acids subclasses had negative directions of effect. \par

Directions of effect were broadly consistent across exposures. The exceptions were for the negative effect of BMI on phosphatidylcholine and other cholines (change in metabolite (mmol/l) per SD increase in BMI = -0.0024; 95% CI = -0.014 -- 0.0089; p-value = 0.0057) compared to the positive effects of WHR and BF (neither of which reached the multiple testing threshold), and the positive effect of BF on concentration of medium HDL particles (change in metabolite (mmol/l) per SD increase in BF = 9.81 x 10^-9^; 95% CI = 1.73 x 10^-10^ -- 1.95 x 10^8^; p-value = 0.046), phospholipids in medium HDL (change in metabolite (mmol/l) per SD increase in BF = 0.0015; 95% CI = -0.0007 -- 0.00379; p-value = 0.19), and total lipids in medium HDL (change in metabolite (mmol/l) per SD increase in BF = 0.0023; 95% CI = -0.0025 -- 0.0073; p-value = 0.35) compared to the negative effects of BMI and WHR (neither of which reached the multiple testing threshold). \par

The apolipoproteins apolipoprotein A-1 and apolipoprotein B are markers of LDL and HDL particles respectively. There was a positive association between all three adiposity measures and apolipoprotein A-1, all of which reached the multiple testing threshold (e.g., change in metabolite (g/l) per SD increase in BMI = -0.034; 95% CI = -0.04 -- -0.03; p-value = 3.64 x 10^-34^). There was also a negative association between all three adiposity measures and apolipoprotein B all of which reached the multiple testing threshold (e.g., change in metabolite (g/l) per SD increase in BMI = 0.046; 95% CI = 0.039 -- 0.052; p-value = 2.93 x 10^-44^). In addition, there was a positive association across all adiposity measures with the ratio of apolipoprotein B to apolipoprotein A-1, all of which reached the multiple testing threshold (e.g., change in ratio of apolipoprotein B to apolipoprotein A-1 per SD increase in BMI = 0.037; 95% CI = 0.035 -- 0.043; p-value = 2.43 x 10^-78^). \par

### Comparison with previous work {#chapter4-wurtz-comparison}
A total of 82 metabolites were measured using the same platform by Wurtz et al. (2014)[@Wurtz2014]. Of these, 42 were also measured across children, adolescents, young adults, and adults in this analysis using ALSAPC. As Wurtz et al., transformed metabolite values into SD units, which was not the case here, directional consistency was investigated here instead. Across all analyses conducted here for model 2, a majority of metabolites showed a consistent direction of effect with the study by Wurtz et al.[@Wurtz2014] (Figure \@ref(fig:observational-figure-wurtz-comparison)). However, there were some metabolites where the effects from Wurtz et al., were opposite to the effects found here, such as fatty acid chain length and albumin. \par

The majority of effect estimates were in the positive direction, i.e. an increase in adiposity was associated with an increase in metabolites. Broadly, adiposity had an increasing effect on 30 metabolites and a decreasing effect on 12 metabolites, i.e. more than half of effect estimates across all exposures and age groups for a single metabolite were positive or negative. A total of 18 metabolites were positively associated with adiposity in analyses by Wurtz et al., and those conducted in all age groups and exposures here, this included phenylalanine, tyrosine, and apolipoprotein B. Seven metabolites were negatively associated with adiposity across analyses by Wurtz et al., and those conducted in all age groups and exposures here, including citrate and apolipoprotein A1. Looking at the overall effects for each subclass, there were primarily positive effects of adiposity for subclasses aromatic and branched chain amino acids, cholesterol, fatty acids, glycerides and phospholipids, and glycolysis related metabolites. Primarily negative effects were found for lipoprotein particle size and fatty acids ratios. The remaining subclasses were evenly split between positive and negative associations between adiposity measures and metabolites across analyses (amino acids and fluid balance) or composed of one metabolite (intermediate density lipoprotein (IDL), inflammation, and ketone bodies). In the latter case, total cholesterol in IDL (IDL) and glycoprotein acetlys (inflammation) were primarily positively associated with adiposity, while acetate (ketone bodies) was primarily negatively associated with adiposity. \par

(ref:observational-figure-wurtz-comparison-cap) **Comparison of the directions of effect estimates from linear regression of adiposity measures on metabolites across all age groups and analysis by Wurtz et al. (2014)[@Wurtz2014].** 

(ref:observational-figure-wurtz-comparison-scap) Comparison of the directions of effect estimates from linear regression of adiposity measures on metabolites across all age groups and analysis by Wurtz et al. (2014)[@Wurtz2014]

```{r observational-figure-wurtz-comparison, echo=FALSE, fig.align='center', out.width='100%', fig.cap='(ref:observational-figure-wurtz-comparison-cap)', fig.scap='(ref:observational-figure-wurtz-comparison-scap)'}
knitr::include_graphics("data/observational/figures/wurtz_comparison.pdf")
```

\vspace*{-\baselineskip}
\noindent The tile plot shows the direction of effect estimate for 42 metabolites as positive (blue) or negative (orange). If all analyses show the same direction of effect, then that row will be the same colour, e.g., all analyses showing a negative association between adiposity measures and apolipoprotein A-1 and a positive association between adiposity and apolipoprotein B BMI = body mass index; WHR = waist hip ratio; BF = body fat percentage; IDL = intermediate density lipoproteins. White space indicates no analysis was undertaken due to missing data. \par

## Discussion {#observational-discussion} 
The association between adiposity and NMR derived metabolites is global, with effects seen across all subclasses of metabolites. These effects are broadly consistent between metabolites within each subclass. In this chapter, the influence of adiposity on the metabolic profile is demonstrated in an observational framework. These effects persist, not only when measured at different ages, but also when adjusting for covarialbes such as smoking, physical activity, and diet. Effects are similar across multiple measures of adiposity and to those previously reported[@Wurtz2014]. \par

There were similar associations seen in individuals when looking at each exposure across age groups (e.g., the association between BMI on metabolites in children, adolescents, young adults and adults). As age increased, the number of associations within subclasses tended to increase across all measures of adiposity. Fewer associations were observed in children than in adolescents and young adults. A similar metabolic profile was apparent in adults, though effect sizes appeared slightly larger. These results may reflect (i) an effect of prolonged adiposity exposure, (ii) increased variation in metabolites with age, or (iii) be a result of the different SDs of the adiposity measures at different ages. Given that longitudinal work has shown that BMI tracks over time[@Singh2008; @Buscot2018], there may be a dose-response relationship here whereby adiposity has a compounding effect on metabolites over time. Alternatively, metabolite concentrations have been shown to increase in older populations over time[@Darst2019] and there was evidence that metabolite variation tended to increase as age increased. However, it is likely that the larger effect sizes observed as age increased are a result of the similar increase in SD of the adiposity measures with age. \par

Within each age group, there was directional consistency across the three exposures. Effect sizes across exposures were similar within age groups, with overlapping CIs. Effect estimates for the association of BMI on metabolites was generally larger, across age groups, than effect sizes for WHR or BF. Effects for BF appeared to be closer to, and crossed the null more often than those for BMI and WHR. This may suggest that overall body composition in addition to detrimental deposition, may be driving these effects. That is, the compounding effect of increased overall body mass (primarily through increased fat mass; as estimated by BMI) and the predilection to store this additional mass viscerally (as estimated by WHR), may be more powerful than either of these effects alone on the effect of metabolites. \par

Results were consistent across models, with effect estimates showing the same direction of association for the majority of tests. Effect sizes were broadly similar in the case of models 1 and 2. Effect sizes in model 3 were smaller compared with models 1 and 2. However, CIs overlapped across the majority of tests even if they crossed the null in some instances for model 3 - model 3 had a smaller sample size than model 1 and 2. Broadly speaking, effect estimates attenuated slightly for model 3 compared with model 1 and 2. There was also high consistency in the directions of effect estimates across models within exposures and age groups. These results suggest there could be some confounding but that this was appropriately accounted for given the included covarialbes. Whether these results represent a true causal effect or are subject to reverse causation or residual confounding however requires further investigation. \par

Previous work by Wurtz et al. (2014)[@Wurtz2014] identified numerous associations between BMI and metabolites in a large population. Results here show a broadly similar pattern of association with those by Wurtz et al., with the majority of the 42 comparable metabolites showing consistent directions of effect. This includes phenylalanine and tyrosine, both of which are known to be increased by adiposity and age[@Haufe2016; @Fattuoni2018; @Zheng2016a; @Ho2016; @Newgard2009; @Yu2018; @Xie2014; @Kim2010], as well as consistent negative associations with seven metabolites, including apolipoprotein A1 which is the major component of HDL particles and enables uptake of lipids by HDL particles and the subsequent recycle and excretion of lipids[@Feingold2000; @Marz2017]. Differences in effect estimate direction was minimal and split relatively evenly across age groups. Only one metabolite from the Wurtz et al., study, fatty acid chain length, had a direction of effect (negative) that was inconsistent across all age groups here. However, in follow-up Mendelian randomization analysis by Wurtz et al., BMI was positively associated with fatty acid chain length as was found in results in this chapter. This difference may be a result of residual confounding in the observational analysis conducted by Wurtz et al., as they did not adjust for all of the covarialbes used in analyses in this chapter. \par

Of particular note with the analyses performed here, is the variation observed for metabolite values within age groups. Variation was much larger for derived compared to directly measured metabolites. Pre-analysis processing of the metabolomic data, applied here using `metaboprep`, removed samples (individuals) and metabolites, it does not remove or change individual values. Instead, outlying values are flagged for follow-up. However, deciding what is and is not a real biological value, as opposed to a measurement or technical error, is challenging. This is made more difficult with large samples where the range of metabolite values can be large as it becomes possible to measure individuals at the extremes. In analyses here all raw metabolite values were retained for analysis. In adults, results for a number of derived metabolites (across exposures) were considerably large. When looking post-analysis at the raw values for these metabolites in adults, a number of individuals had metabolite values that appeared to be outliers. In post-hoc sensitivity analysis (data not shown), which aimed to investigate whether these individual values were influencing results, the raw metabolite values in the top and bottom 1% of metabolite values were changed to the next closest metabolite value below that 1% threshold. This Winsorization process was performed for all metabolites across age groups. All analyses were repeated and results were highly consistent for the main analysis and this sensitivity analysis for children, adolescents, and young adults. For adults, results were highly consistent for directly measured metabolites and a majority of derived metabolites. However, results for the metabolites that appeared to be outliers in the main analysis were considerably different in sensitivity analysis, with effect sizes much closer to the null and much tighter confidence intervals. This would suggest that the strange results in the main analysis for those derived metabolites where likely a result of the outlying metabolite values for a handful of individuals. Given the difficulty in identifying non-biological values, future work should perform sensitivity analyses such as that described here, to investigate whether outlying data has impacted results. This recommendation, as well as methods to control for outlying data, are being added to `metaboprep` to aid future work. \par

### Strengths and limitations
The key strengths of this work are the use of (i) rich and large-scale data resource that is ALSPAC, (ii) BMI, WHR and BF as complimentary adiposity measures to comprehensively examine the association between adiposity and metabolites and (iii) the longitudinal nature of the ALSPAC data allowing for this association to be assessed across the lifecourse. ALSPAC is a prospective cohort study with a general population base and here, analyses were performed on up to 4,450 individuals. The breadth of data measured across multiple generations enabled linear models to be adjusted for covarialbes that have been known to affect adiposity and metabolite concentrations[@Yu2012; @Brennan2020; @Brennan2020; @Zajacova2018; @Bonevski2014; @Afshin2019; @ODonoghue2018]. These analyses were performed across two generations of individuals, G0 (the original mothers recruited to the study) and G1 (the children from those pregnancies). Data on G1 individuals were available from multiple time points and, in combination with data from G0 individuals, enabled the investigation of the association between adiposity and metabolites across the lifecourse from childhood to adulthood. The use of multiple measures of adiposity, as has been recommended previously[@WorldHealthOrganisation2008; @Collaboration2009] but which has not been explored in relation to metabolic effects, may indicate that overall body composition and deposition is more important in metabolic changes than either component alone. \par

Analyses here used individuals from two generations, G0 (mothers and fathers) and G1 (offspring) individuals. Although G0 and G1 individuals are independent, there is familial overlap between them. There is also overlap for the children, adolescents, and young adults age groups whereby the same individual may be present in all three age groups. As such, any effects specific to a G1 individual will be propagated through all G1 clinics that individual attends. In addition there may be familial overlap within the G1 age groups given siblings were included (unlike twins, triplets, and quadruplets) in the metabolomic data collection. The number of sibling pairs in each age group is likely small. However, given the shared genetics and environment these individuals (siblings and parent and offspring pairs) will have, effects observed in one individual will likely be seen in their relations, and so results will not be truly independent within the G1 age groups and across the G1 and G0 age groups. Familial overlap will likely bias results towards the confounded effect estimate given familial overlap is similar to confounding by population structure. Future work could remove related individuals prior to analysis. However, where data is limited and exclusions may reduce power, the inclusion of genetic principal components to account for relatedness among individuals is also possible. \par

Observational analyses are limited due to issues of confounding and reverse causation. Prospective studies, such as ALSPAC, are able to provide some temporal separation between the measurement of an exposure and outcome, and thus mitigate the potential effects of reverse causation. However, as data on adiposity measures and metabolites were collected at the same time, reverse causation remains a limitation. It was however possible to account for the potential effects of confounding by adjusting linear models for measured confounders. Although many measured confounders were included across the models, it is likely these analyses will not have fully accounted for confounding, either due to measurement error or unmeasured confounding. For example, adults were asked 'do you take part in physical activity (e.g. running, swimming, dancing, golf, tennis, squash, jogging, and bowls)?', with possible answers of 'no', 'occasionally (less than monthly)' and 'frequently (once a month or more)'. Broad categories such as these are unlikely to capture the full impact of physical activity on adiposity and metabolites given that 'frequently' will encompass individuals who exercise once a month as well as every day. In addition, although it was possible to investigate the association between adiposity and metabolites across the life course, it is likely that the measurements used during adolescence and young adulthood may reflect a time where the G1 individuals were going through puberty. Puberty is likely to have an effect on adiposity and covarialbes such as physical activity, as well as the metabolome, and future work should therefore include pubertal timing as a covariable. \par

There are two key limitations in regards to metabolomic data. The first is that these data are limited in their breadth. Although a relatively large number of metabolites were investigated, they were predominantly lipid-based. This leaves a broad array of metabolites that have not been investigated. As such, the metabolites investigated here are broadly reflective of the lipidome and not a global metabolic profile. The Nightingale NMR platform does however provides many derived measures, such as ratios and number of bonds. However, there was considerable variation in the concentration of these derived measures compared to directly measured metabolites. Given the relationship between adiposity and lipids, it is unsurprising there were a large number of associations identified. There is thus a clear need to expand beyond lipid-based platforms in investigating the effects of adiposity, as the associations presented here are likely not reflective of a global metabolic profile. Mass spectrometry (MS) platforms, such as those used by Metabolon, are a potential next step. Though there have been a number of studies investigating the association between adiposity and MS-derived metabolites[@Moore2018; @Cirulli2019; @Perng2020], these have not used a complimentary assessment of body composition to assess adiposity and have not been able to investigate this relationship across the life-course. \par

There is no standardised approach, nor a gold standard, for performing metabolomic quality control. Here, pre-analysis processing of metabolomic data, including outlier detection and removal, was performed using the `metaboprep` `R` package. The default settings for exclusions based on metabolite missingness ($\geq$ 20%), sample missingness ($\geq$ 20%), total sum abundance ($\geq$ 5 SD), and principal components ($\geq$ 5 SD; PCs) were used. Most samples were removed for having missingness > 20% compared to total peak area. These thresholds were arbitrarily defined, or taken from other studies, such as 20% sample missingness which was previously used, and arbitrarily defined, in Lotta et al., (2021)[@Lotta2021]. More stringent thresholds (e.g. $\geq$ 5%) may have resulted in additional sample exclusions. Metadata, such as batch and runday, was not available and could not be included in pre-analysis processing, though this is likely more an issue with MS data. `metaboprep` calculated the number of independent metabolites using a clustering dendrogram and a tree cut height based on a Spearman's Rho of 0.5. This is one of many clustering approaches that could be used. For example, adjusting the tree cut height to any other value would have resulted in a different number of independent metabolites being identified, which subsequently would have impacted on multiple testing thresholds used. In addition, there is no rule as to what clustering method (i.e., linkage) should be used, and all methods would have produced different results. The complete-linkage method used here is potentially more appropriate for metabolite data compared to single-linkage, where it is possible for clusters to be combined due to a single element of each cluster being linked, even though all other elements of the clusters are not[@Everitt2011]. Although most metabolites were shared across age groups, differences in the number of independent metabolites were found. There were also differences in the number of clusters and truly independent metabolites. Similarly, inclusion of derived metabolites resulted in differing numbers of independent metabolites. Metabolites did not undergo transformation to standardise their concentrations in order to preserve clinical utility and improve interpretation of results. \par

Metabolomic data used here were from EDTA stored plasma and serum samples. There is evidence that the manner in which samples are stored (e.g., EDTA) can impact the readings from NMR platforms[@Paglia2018; @Sotelo-Orozco2021]. Specifically, evidence shows there is a measurable difference between metabolites from samples that have not-undergone storage and samples that have. For example, the study by Sotelo-Orozco et al., (2021)[@Sotelo-Orozco2021] using an NMR platform identified evidence for a difference between EDTA stored plasma and serum samples that had not been stored for 5 metabolites (out of 50), including lower levels of arginine and taurine, and higher levels of pyruvate, acetate, and formate in EDTA compared to non-EDTA samples; only pyruvate and acetate were measured in analyses in this chapter. Although more than the number of differences observed for heparin stored plasma (3 metabolites out of 50), this was fewer than the number of differences observed for fluoride stored plasma (11 out of 50 metabolites), citrate stored plasma (24 out of 50 metabolites), and acid citrate dextrose (ACD) stored plasma (29 out of 50 metabolites). Sotelo-Orozco et al., however only assessed 50 metabolites and did not assess lipids, which are a large component of the metabolomic data available in ALSPAC. Paglia et al., (2018)[@Paglia2018], using a MS platform, found similar changes for amino acids. They also found some evidence that lipid metabolite values are different in EDTA stored plasma samples compared to non-EDTA stored serum samples. They suggested that citrate stored plasma may be more appropriate for lipidomics as values were closer to those of the serum samples compared to the EDTA stored plasma samples. Both studies however found that a majority of metabolites did not differ considerably between EDTA and serum samples. Although results for some metabolites such as pyruvate need to be taken with consideration to the use of EDTA to store samples in ALSPAC, it is likely the majority of metabolite values will not be considerably different to samples which were not stored using EDTA. \par

Metabolomic data were measured using the same NMR platform for all individuals. Briefly, high-throughput proton (^1^H) NMR assays were performed on ethylenediaminetetraacetic acid (EDTA) plasma/serum samples. Samples were fasted. Measurements were taken at three molecular windows (lipoprotein lipids, low molecular-weight metabolites, and lipid extracts) enabling broad quantification of metabolomic measures. These measures also included derived measures, lipoprotein particle sizes, and fatty acid ratios, inclusion of which has shown to increase overall power in statistical analyses[@Altmaier2008; @Gieger2008; @Suhre2010]. Metabolite values for each individual were provided by the NMR platform in the originally measured units (e.g., mmol/l). Derived metabolite values are as a %. Full details on the NMR methodology has previously been described[@Soininen2009; @Inouye2010; @Kettunen2012; @Soininen2015] and is available from the [ALSPAC data dictionary](http://www.bristol.ac.uk/alspac/researchers/access/) (data dictionary identifiers: children = D5704, mothers = D5705, fathers = D5700). The Before Breakfast study does not have a documentation file and is described elsewhere[@Ong2004]; fasted metabolomic data, not the post glucose challenge metabolomic data, were taken from The Before Breakfast Study. Descriptions of metabolites are available on [GitHub](https://github.com/mattlee821/000_thesis/blob/master/index/data/observational/tables/metabolites.txt). \par

Metabolomic measures were taken at specific time points in ALSPAC children and at ~50 years of age in adults. Though measures of adiposity were available at the same time points, data on confounders were not. Smoking status for example was available for adult males at the metabolomic clinic but the closest available measure for adult females was a number of years earlier, in which time they may have changed smoking status. In addition, the smoking variable used was 'ever smoked a cigarette', this will have meant a lot of non-smokers were categorised as smokers. Although data were obtained where available from the closest time point, the mismatch in timings may have led to suboptimal adjustment of confounding factors, increasing the number of false positives. \par

In all age groups, the availability of data was limited. Absence of physical activity and diet data meant model 3 was not performed for children and models 2 and 3 for young adults and adults were not adjusted for diet. However, adjusting for diet, where available (children and adolescents), had little impact on results (Table \@ref(tab:observational-table-significant-results)) and is therefore not likely to have changed findings in young adults and adults. That being said, given children and adolescents will primarily eat what their parents want them to eat, the effect of diet on the association between adiposity and metabolites may get stronger with age. \par

In children, BF was not available. A raw impedance measure was available and derivation of BF using Equation \@ref(eq:BF) showed a positive correlation with BMI and weight in children and BF measures in adolescents. However, this derived measure included numerous negative values of BF. The equation performed well in adolescents, correlating highly with DXA-derived BF (Figure \@ref(fig:appendix-observational-figure-BF-validation-correlation-2-adolescents)). The negative values of BF are likely a result of Equation \@ref(eq:BF) being derived in an adult population. Brief investigation showed negative values of BF remained when using adolescent age with child height and weight (data not shown). Given that in single frequency devices, impedance is based on the volume of an individual, it is probable that the values used in the equation do not accurately reflect the proportions of children pre-puberty. Child-specific equations were not available and the manufacturer was unwilling to share the equation used by their impedance devices. However, as the estimates derived in linear models are in reference to the per unit increase in an exposure (rather than the range), the absolute value of the exposure does not need to positive. Thus, the negative values of BF will not impact the inference of a linear regression between BF and any metabolite. \par

The distribution of BMI at each age group was very similar across sexes. However, the distribution of WHR and BF differed among sexes in adolescents, young adults and adults; males had on average a higher WHR, while females had a higher BF. Though Z-scores were used and sex was included as a covarialbe, the differences in WHR and BF distributions may highlight an underlying difference which, if associated with metabolites, may have confounded the relationship between adiposity and metabolites. For example, hormonal contraceptive use has shown to influence the metabolome[@Rauschert2017] and may be used less often by individuals with a high BMI[@Mody2014]. Future work should therefore look at sex-combined and sex-specific analyses. \par

There was little difference in results from the three models in regards to direction of effect estimates. However, effect estimates for model 3 compared with models 1 and 2 attenuated slightly and had wider confidence intervals. In addition, the number of associations reaching a multiple testing threshold for model 3 was lower than for models 1 and 2. This is likely a result of the reduced sample size used in model 3 compared with models 1 and 2. However, given the consistency in the direction of effects across models and the highly similar effects across model 1 and 2, there is likely little effect of confounding. That being said, the possibility of unmeasured confounding can not be ruled. \par

### Conclusion
The large number of associations identified using multiple exposures adds weight to the evidence, shown previously for BMI, of the global association between adiposity and metabolites. Work in this chapter used a deeply phenotyped, general population based cohort with longitudinal data that enabled examination of the effect of adiposity on metabolites across multiple ages. As with previous work, there were associations between adiposity and many metabolites, including amino acids such as phenylalanine and tyrosine, and lipids, such as total fatty acids and HDL components. This work suggests the broad and consistent effects of adiposity on a wide array of lipid metabolites likely persist over time. Of particular note is the increasing effect size and number of associations as a result of adiposity across the metabolic profile with age. Though this may be a result of the different SDs for adiposity measures at different ages, given that many adiposity-associated diseases occur later in life, exposure to an altered metabolic profile over time may be important in disease development. This is especially true as weight loss in overweight and obese individuals is associated with a normalizing of metabolite changes[@Rangel-Huerta2019]. There was weak evidence of confounding, however the possibility of unmeasured or residual confounding remains. Follow-up analysis in Chapter \@ref(MR), will utilise Mendelian randomization to overcome these limitations. \par

\newpage