---
output: 
    thesisdown::thesis_pdf: default
bibliography: bib/thesis.bib
csl: csl/nature.csl
space_betwee_paragraphs: true
fig_caption: true
always_allow_html: yes
link-citations: true
toc-depth: 2
lot: true
lof: true
indent: true
header-includes: # include other LaTeX packages here
    \usepackage{booktabs}
    \usepackage{longtable}
    \usepackage{siunitx}
    \usepackage[left]{lineno}
    \usepackage{float}
    \floatplacement{figure}{H}
    \linenumbers
    \pagestyle{plain}
    \raggedbottom 
    \usepackage{indentfirst}
editor_options: 
  chunk_output_type: console
---

```{r include=FALSE}
# word count = 
doc.type <- knitr::opts_knit$get('rmarkdown.pandoc.to') # this will tell the .Rmd file what output you are knitting to (word, pdf, html) so that you cna use if else statements when making tables - html/pdf output tables do not work well in word. For word we need to use the 'flextable' packaged to make tables.
## packages
library(dplyr)
library(kableExtra)
library(flextable)
library(wesanderson)
library(cowplot)
library(ggplot2)
options(tinytex.verbose = TRUE)

##### Palettes
d1 <- wes_palette("Royal1", type = "discrete")
d2 <- wes_palette("GrandBudapest2", type = "discrete")
d3 <- wes_palette("Cavalcanti1", type = "discrete")
d4 <- wes_palette("Darjeeling1", type = "discrete")
discrete_wes_pal <- c(d1, d2, d3, d4)
continuous_wes_pal <- wes_palette("Zissou1", 100, type = "continuous")
```

# Associations between adiposity-associated metabolites and endometrial cancer: Mendelian randomization analysis {#mediation} 

<style>
body {
text-align: justify}
</style>

## Chapter summary {-}
This Chapter pulls together information from previous chapters to investigate the effects of adiposity-associated metabolites on endometrial cancer. In Chapter \@ref(systematic-review), a systematic review and meta-analysis showed that adiposity was likely causally implicated in 13 diseases as well as other traits including systolic blood pressure and fasting glucose. Of these, endometrial cancer, with which body mass index (BMI) was strongly related (OR = 1.57; 95% CI = 1.11 -- 2.22; p-value = 0.01), was selected for further investigation in this Chapter. Here, two-sample Mendelian randomization (MR) was used to estimate the association between adiposity (BMI, waist-hip ratio (WHR), and body fat percentage (BF)) and endometrial cancer, extending analyses presented in Chapter \@ref(systematic-review) by including a more comprehensive set of adiposity measures and increasing the sample sizes of data sources. Two-sample univariable MR was performed to estimate the effect of adiposity-associated metabolites, characterised in analyses undertaken in chapters \@ref(observational) and \@ref(MR), and endometrial cancer. To understand whether metabolites, for which there was evidence for an association with endometrial cancer, may play an intermediary role in the relationship between adiposity and endometrial cancer, the largest causal analysis (including multivariable MR; MVMR) to date was performed. \par

\newpage

## Introduction {#mediation-introduction}

<!--
Many observational findings have been replicated using Mendelian randomization (MR; Chapter \@ref(systematic-review)), while others, such as c-reactive protein[@Timpson2005; @CCGC2011], have not. Investigations of many cancers have shown consistent estimates across observational and MR analyses, for example BMI is observationally (OR 1.10; 95% CI = 1.07--1.13)[@Bhaskaran2014] and causally (men: OR = 1.23; 95% CI = 1.08--1.38; women: OR = 1.09; 95% CI = 0.97--1.22)[@Bull2020] associated with colorectal cancer. 

Maintaining a *healthy* weight can thus be an effective form of cancer prevention. However, although there is evidence to show that weight loss interventions reduce the risk of cancer[@Ma2017; @Force2018; @Arnold2019; @Luo2017; @Clifton2018; @Brown2016; @Cercato2019], there are many barriers to behavioural weight loss interventions[@McVay2018] not least understanding who will benefit best from what. \par
-->

In chapters \@ref(introduction) and \@ref(systematic-review), I showed that the number of adiposity-associated diseases is high, consistent with adiposity being a global health concern[@WorldHealthOrganisation2018; @Ritchie2019; @Ng2014; @NCD-RisC2016; @Abarca-Gomez2017]. Many observational studies have highlighted associations with common diseases, such as cardiovascular disease[@Hubert1983; @Kenchaiah2002; @Yusuf2004; @Cuilin2008; @Gelber2008; @Litwin2008; @Dobbelsteyn2001; @Yusuf2005; @Paul2006; @Koliaki2019; @Eckel1998; @Cepeda-Lopez2010; @Blokhin2013; @Thomsen2014; @Klop2013; @McGill2002; @Ades2017; @Mathew2008; @Plourde2014; @Pathak2015; @Lavie2017; @Kurth2002; @Kernan2013; @Rost2001; @Romero-Corral2010; @Lee2018], cancers[@Bhaskaran2014; @Choi2020; @Kyrgiou2017; @Hu2014; @Hajiebrahimi2013; @Munsell2014; @Harvie2003; @Gaudet2014; @Barrington2015; @Heo2021; @Lauby-Secretan2016; @Barberio2019; @Suzuki2009; @Freisling2017; @Lee2018; @Parra-Soto2021], and common risk factors such as blood pressure[@Mathew2008; @Wurtz2014]. Many of these associations have been supported in MR studies as discussed in Chapter \@ref(systematic-review). \par

Previous work[@Wurtz2014; @Moore2014; @SantosFerreira2017; @Cirulli2019; @Lau2020; @Brachem2020; @Stevens2020; @OKeeffe2020; @Wulaningsih2019; @Frigerio2021; @Neeland2019; @Bachlechner2016; @Rangel-Huerta2019; @Bull2020; @Bell2021a], and work conducted in chapters \@ref(observational) and \@ref(MR), has highlighted the numerous metabolites associated, in observational and MR analyses, with measures of adiposity. Many of these adiposity-associated metabolites have been linked with adiposity-associated diseases and risk factors including type 2 diabetes[@Liu2017c; @Porcu2021], fasting glucose[@Liu2017c; @Porcu2021], colorectal cancer[@Bull2020], and coronary heart disease[@Burgess2016a], suggesting a potential intermediate role for metabolites. There is also evidence that metabolites can be used to distinguish cancers[@Kovesdy2017]. \par

To date, few studies have investigated the causal relationship between adiposity-associated metabolites and adiposity-associated diseases. As an exemplar outcome to test the potential intermediary role of adiposity-associated metabolites, identified in chapters \@ref(observational) and \@ref(MR), and adiposity-associated disease, identified in Chapter \@ref(systematic-review), endometrial cancer was chosen. This was because endometrial cancer met four key requirements: there was strong evidence in Chapter \@ref(systematic-review) for an effect of one or more adiposity measures; there is consistent evidence across observational[@WILD2020] and MR[@Gharahkhani2019; @Painter2016; @Yarmolinsky2019] analyses that adiposity is associated with an increased risk of endometrial cancer; there is a large and publicly available genome-wide association study (GWAS) on endometrial cancer, and the extent to which circulating metabolites may play a role in the relationship between adiposity and endometrial cancer has not been assessed in the literature previously. \par

The World Cancer Report (2020) has identified endometrial cancer as one cancer that would benefit greatly from reductions in adiposity, as approximately 30-40% of cases are a result of obesity[@WILD2020]. The endometrium is the inner lining of the uterus, consisting of an epithelial cell layer and mucous membrane. The epithelial cell layer consists of two parts, a basal layer and a functional layer which thickens and sheds during menstruation in response to oestrogen and progesterone[@Critchley2020]. Broadly, endometrial cancer is separated into two types: endometrioid (type 1) and non-endometrioid (type 2); the latter being the more aggressive form of the disease. \par 

Endometrioid cancer is the more common type of endometrial cancer and is considered hormone dependent. That is, most endometrioid cancers develop from endometrial hyperplasia which is associated with prolonged and unopposed oestrogen exposure[@Amant2005]. Non-endometrioid cancer is much less common (~10%) and has typically been considered hormone independent[@Amant2005]. However, this distinction is not clear cut and, as many of the risk factors are shared, non-endometrioid cancer may not be oestrogen independent as once thought, though it is likely to be less dependent upon oestrogen than endometrioid cancer[@Setiawan2013; @WILD2020]. Molecular differences are apparent between the two types, with *PTEN* mutations (associated with oestrogen exposure) found in many endometrioid cancers and *TP53* and *HER2* (*TP53* is a tumour suppressor gene and *HER2* is a proto-oncogene that promotes cell growth) mutations found in non-endometrioid cancers[@Amant2005]. \par

A leading risk factor of endometrioid cancer in pre- and post-menopausal women is adiposity[@WILD2020]: in pre-menopausal women, adiposity leads to insulin resistance, increased androgen, anovulation, and decreased progesterone; in post-menopausal women, adiposity leads to increased oestrogen via excess androgen conversion[@Amant2005]. In both instances, these changes result in an increase in oestrogen and subsequent endometrial hyperplasia[@Amant2005]. Additionally, there is evidence of oestrogen-independent activation of the oestrogen-receptor, primarily through insulin-like growth factor 1[@Lacey2004], which is also up-regulated as a result of adiposity. \par

In Chapter \@ref(systematic-review), meta-analysis of three MR studies[@Gharahkhani2019; @Painter2016; @Yarmolinsky2019] found the offs of endometrial cancer is 1.57 (95% CI = 1.11 -- 2.22) higher per SD unit increase in BMI. Previous work has shown different metabolite profiles between overall endometrial cancer cases and controls independent of obesity and other risk factors such as diabetes[@Gaudet2012; @Kliemann2021]. More recently, adiposity-associated metabolites, such as those identified in chapters \@ref(observational) and \@ref(MR), have been linked with endometrial cancer risk in a prospective cohort[@Dossus2021]. Whether there is a causal effect of metabolites on endometrial cancer development however is unclear. Here, I combine work undertaken in the previous chapters, which identified metabolites causally related to adiposity, and utilise two-sample univariable MR and two-sample multivariable MR (MVMR) to understand the possible intermediary role played by adiposity-related metabolites on endometrial, endometrioid, and non-endometrioid cancers using data from the largest publicly available metabolomics and endometrial cancer GWAS to date.

## Methods
### Overview
```{r eval=TRUE, include=TRUE, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
# associated metabolites
associated_metabolites <- read.table("../../006_interval_metabolites/analysis/associated_metabolites.txt", header = T, sep = "\t")
associated_metabolites <- unique(associated_metabolites$outcome)
```

This Chapter details hypothesis-driven MR analyses investigating associations between adiposity-related metabolites (here in referred to as metabolites) and endometrial cancer (Figure \@ref(fig:mediation-figure-analysis-flow)). Firstly, a two-sample univariable MR analysis of the effect of multiple measures of adiposity on endometrial cancer was performed to independently validate findings from the meta-analysis performed in Chapter \@ref(systematic-review). Secondly, a two-sample univariable MR analysis of the effect of multiple measures of adiposity on metabolites was performed to further validate the work conducted in Chapter \@ref(MR) and to obtain the best possible estimates of the effect of adiposity on metabolites for MR analyses with endometrial cancer. This analysis was performed using recently available data which included the largest currently available GWAS of circulating metabolites conducted in UK Biobank. Thirdly, metabolites with consistent directions of effect across results performed in Chapter \@ref(MR) and this Chapter, were used in a two-sample univariable MR analysis to investigate their effect on endometrial cancer. Finally, metabolites for which there was evidence of an effect on endometrial cancer were taken forward and used in a MVMR analysis with adiposity measures to estimate their possible intermediate effects on endometrial cancer. \par

All data manipulation and analyses were performed using `R`[@r2019] (version 3.5.3) and `bash`. All code used is available on [GitHub](https://github.com/mattlee821/007_metabolites_outcomes/tree/master/scripts). Two-sample univariable MR analyses were performed using the `TwoSampleMR`[@Hemani2018] (version 0.4.22) `R` package. MVMR analyses were performed using the `MVMR` (version 0.3) `R` package. Summary statistics for adiposity measures and metabolites were obtained from the original GWAS sources (described in the next sextion). Summary statistics for metabolites were from UK Biobank and were obtained from collaborators prior to publication of the GWAS (unpublished; Carolina Borges, University of Bristol). Summary statistics for endometrial cancer were available from MR-Base[@Hemani2018] (accessed 17/07/2021). A list of metabolites used in this Chapter is available in the Appendix (Table \@ref(tab:appendix-mediation-table-metabolites)) and on [GitHub](https://github.com/mattlee821/000_thesis/blob/master/index/data/mediation/tables/metabolites.txt). Results were visualised using the `ggforestplot` (version 0.1.0) `R` package. \par

\newpage
\thispagestyle{empty}
\vspace*{-3cm}

(ref:mediation-figure-overview-cap) **Analysis overview**. In Chapter \@ref(MR), `r length(associated_metabolites)` metabolites were identified as being associated with adiposity in a meta-analysis using data from Kettunen et a., (2016)[@Kettunen2016] and INTERVAL (unpublished and provided by Adam Butterworth, University of Cambridge). For body mass index (BMI) and waist hip ratio (WHR) these metabolites were identified through directional consistency and multiple testing thresholds across Mendelian randomization (MR) and observational analyses. For body fat percentage (BF) metabolites were identified through directional consistency and a multiple testing threshold in the observational analysis only. These `r length(associated_metabolites)` metabolites were taken forward and used in two-sample univariable MR and multivariable MR analyses to investigate intermediate effects on endometrial cancer. NMR = nuclear magnetic resonance.

(ref:mediation-figure-overview-scap) Analysis overview

```{r mediation-figure-analysis-flow, echo=FALSE, warning=FALSE, error=FALSE, out.width='98%',fig.cap='(ref:mediation-figure-overview-cap)', fig.scap='(ref:mediation-figure-overview-scap)'}
knitr::include_graphics("../index/data/mediation/figures/overview.pdf")
```

### Instrumentation {#mediation-methods-instrumentation}
As discussed in Chapter \@ref(MR) Section \@ref(MR-methods-instrumentation), instrumentation of exposures is primarily achieved using either single genetic variants or multiple genetic variants as instrumental variables. Using multiple genetic variants in an instrument which, collectively, explain a greater proportion of the trait variance than any one individual variant can mitigate weak instrument bias. Generally, instruments are obtained from the largest and most recent GWAS using a genome-wide significance threshold of p-value < 5 x 10^-8^. This was the most common approach used for instrumenting adiposity measures identified in the systematic review (Chapter \@ref(systematic-review)) and was implemented in Chapter \@ref(MR). Few of these studies reported on the independence of SNPs (e.g., a linkage disequilibrium (LD) r^2^ and distance threshold). Similar approaches have been used for studies investigating the association between metabolites and outcomes using an MR framework[@Lord2021; @Lotta2016; @Qian2021; @Liu2017; @Yang2021; @Bull2020; @Qin2020; @Thomas2021]. Many of these metabolite MR studies did report on the independence of SNPs, however approaches were varied (e.g., LD r^2^ thresholds of 0.1, 0.05, and 0.001) and thresholds appeared arbitrarily set. \par

Potential overlap in was considered when selecting instruments for adiposity measures and metabolites. In two-sample univariable MR analyses, overlap between datasets providing summary statistics for the exposure and outcome may bias estimates in the presence of weak instruments. This bias may be exacerbated with greater overlap in the two samples. However, given a strong enough instrument, bias as a result of overlap will be close to the un-biased estimate[@Burgess2016]. For continuous outcomes (e.g., the association between adiposity measures and metabolites), bias away from the null is a linear function of the sample overlap (e.g., sample overlap of 50% leads to a bias of 5%). For binary outcomes (e.g., the association between adiposity measures and metabolites with endometrial cancer), when the association between the SNP and outcome is estimated in all participants, bias is similar to that for a continuous outcome. Where the association between the SNP and outcome is estimated in controls only, unbiased estimates can be obtained[@Burgess2016]. \par

UK Biobank is large and deeply phenotyped prospective sudy which recruited 502,639 participants aged 37–70 years in 22 assessment centres across the United Kingdom[@Sudlow2015]. In 2021, NMR data[@Julkunen2021; @Bell2021a] and corresponding GWAS data became available for 118,466 UK Biobank individuals, and represents the largest single study of its kind. Sex-specific GWAS data were also made available. Given the large sample size, ability to use sex-specific estimates of the association between genetic variation and metabolites and, desire to obtain the best possible SNP-metabolite estimates for MR analyses, UK Biobank metabolomic data were used here. However, as almost all individuals from UK Biobank were included in the GWAS for BMI (Yengo et al., (2018)[@Yengo2018]) and WHR (Pulit et al., (2019)[@Pulit2019]) used in the main analysis of Chapter \@ref(MR), overlap with the UK Biobank data were highly likely. As such, summary statistics for BMI and WHR were obtained from GWAS which did not include UK Biobank. Data from these GWAS (discussed in detail in th next section) were used in additional analyses in Chapter \@ref(MR) and showed highly consistent associations with metabolites when compared to the BMI and WHR instruments obtained from Yengo et al., (2018) and Pulit et al., (2019), respectively. As such, there is no overlap between adiposity data and metabolite data used in analyses here. However, as the largest available endometrial cancer GWAS, with information on subtypes, included individuals from UK Biobank, there is potential overlap between metabolite data and overall endometrial cancer data. In analyses here, the potential overlap between metabolite GWAS data and endometrial cancer GWAS data is a maximum of `r round(636/12906*100)`%, which is expected to equate to ~0.5% increased false positive rate. These data sources are described in full in the next section. \par

In this chapter, genetic variants were identified from summary statistics of GWAS for adiposity measures using a genome-wide significance threshold of p-value < 5 x 10^-8^ and no LD r^2^ and distance threshold. An LD r^2^ and distance threshold was not set as results from additional analyses in Chapter \@ref(MR) of the association between adiposity measures and metabolites using clumped and non-clumped instruments, did not identify a difference in effect estimates as a result of clumping. For metabolites, as no investigation of different instrumentation approaches has been conducted and, given studies appear to arbitrarily set LD r^2^ and distance thresholds and there is likely a common genetic architecture across metabolites, genetic variants were identified using a genome-wide significance threshold of p-value < 5 x 10^-8^ and a conservative LD r^2^ (0.001) and distance threshold (10,000 bases). \par

<!--
#### Overlap {#mediation-methods-overlap}
In two-sample univariable MR analyses, overlap between datasets providing summary statistics for the exposure and outcome may bias estimates in the presence of weak instruments which may be exacerbated with greater overlap in the two samples (given a strong enough instrument, bias as a result of overlap will be close to the un-biased estimate)[@Burgess2016]. For continuous outcomes, bias away from the null is a linear function of the sample overlap (e.g., sample overlap of 50% leads to a bias of 5%). For binary outcomes, when the association between the SNP and outcome is estimated in all participants, bias is similar to that for a continuous outcome. Where the association between the SNP and outcome is estimated in controls only, unbiased estimates can be obtained[@Burgess2016]. In the analyses discussed in this Chapter, there is potential overlap between metabolite data (intermediate exposure) and overall endometrial cancer data (outcome) of a maximum of `r round(636/12906*100)`%. This `r round(636/12906*100)`% maximum overlap would equate to ~0.5% increased false positive rate. \par
-->


### Data
The following section details a number of GWAS and meta-analyses of previously published studies, which were used in this Chapter to perform MR analyses and are required as per STROBE-MR guidelines[@DaveySmith2019]. I was not not directly involved in these GWAS. The total sample size (N) and sex-specific sample size does not always tally, this is due to variation in the sample size for each SNP. Where this is the case, 'sample size up-to' is used. Due to the way in which studies report GWAS differently, information present in one study (such as sex-specific sample sizes) are not always present in another. \par

<!--Ancestries (as opposed to ancestry) is used throughout as multiple ancestries are encompassed within the 'European ancestry' classification (see: [https://twitter.com/ewanbirney/status/1148574135284633600?s=20](https://twitter.com/ewanbirney/status/1148574135284633600?s=20)). -->

#### Exposures: Adiposity
##### Body mass index
Detailed information is presented in Chapter \@ref(MR) Section \@ref(MR-exposures-adiposity). Briefly, summary statistics for the association between genetic variation and BMI were obtained from Locke et al. (2015)[@Locke2015], in which 322,154 individuals of European ancestries were included in a fixed effects inverse variance weighted meta-analysis. A total of 82 GWAS and 43 studies using the Metabochip array were included in the meta-analysis. Individual GWAS were adjusted for age, age squared, and study specific covariates with residuals inverse rank normally transformed. Imputation was performed using HapMap phase II Utah residents of Northern and Western European ancestries (CEU) reference panel. Each study used a linear regression model assuming an additive genetic model with quality control following procedures outlined previously[@Winkler2014]. A fixed effects inverse variance weighted meta-analysis was performed using METAL for the 82 GWAS and 43 studies using the Metabochip array separately. The final meta-analysis combined the single nucleotide polymorphisms (SNPs) found in both the meta-analyses of GWAS and Metabochip studies that underwent genomic control. A total of 77 loci reaching genome-wide significance (p-value  $\leq$ 5x10^-8^) and separated by at least 500 kilobases were identified. These SNPs explained ~2% of the variance in BMI. All 77 SNPs identified by Locke et al., (2015) were also identified in the Yengo et al., (2018) BMI GWAS used in Chapter \@ref(MR). \par

##### Waist hip ratio
Detailed information is presented in Chapter \@ref(MR) Section \@ref(MR-exposures-adiposity). Briefly, summary statistics for the association between genetic variation and WHR were obtained from Shungin et al. (2016)[@Shungin2015], in which 210,088 individuals of European ancestries were included in a fixed effects inverse variance weighted meta-analysis. A total of 57 GWAS and 44 studies using the Metabochip array were included in the meta-analysis. WHR was adjusted for age, age squared, study-specific covariates if necessary with residuals inverse rank normally transformed. Imputation was performed using HapMap phase II CEU reference panel. Each study used a linear regression model assuming an additive genetic model. The final meta-analysis combined the SNPs found in both the meta-analyses of GWAS and Metabochip studies that underwent genomic control. A total of 26 loci reaching genome-wide significance (p-value $\leq$ 5x10^-8^) and separated by at least 500 kilobases were identified. The variance explained for these SNPs was not reported, however it will be lower than the 3% variance explained by the 316 SNPs identified by Pulit et al., (2019)[@Pulit2019]. All 26 SNPs identified by Shungin et al., (2016) were also identified in the Pulit et al., (2019) WHR GWAS used in Chapter \@ref(MR). \par

##### Body fat percentage
Detailed information is presented in Chapter \@ref(MR) Section \@ref(MR-exposures-adiposity). Briefly, summary statistics for the association between genetic variation and BF were obtained from Lu et al., (2016)[@Lu2016], in which 89,300 individuals of European ancestries were included in a fixed effects inverse variance weighted meta-analysis. A total of 43 GWAS and 13 studies using the Metabochip array were included in the meta-analysis. Individual GWAS were adjusted for age, age squared, and study specific covariates with residuals inverse rank normally transformed. Imputation was performed using HapMap phase II European reference panel. Each study used a linear regression model assuming an additive genetic model. A fixed effects inverse variance weighted meta-analysis was performed using METAL for the 43 GWAS and 13 studies using the Metabochip array separately. The final meta-analysis combined the SNPs found in both the meta-analyses of GWAS and Metabochip studies that underwent genomic control. A total of 7 loci reaching genome-wide significance (p-value  $\leq$ 5x10^-8^) and separated by at least 1 mega base were identified. Estimation of the variance explained was not available in the European ancestries meta-analysis. In a meta-analysis of individuals of all ancestries, which included up to 11,419 additional individuals of non-European ancestries, these 7 SNPs explained 0.416% of the variance in BF. The additional 5 SNPs identified in this all ancestries meta-analysis explained 0.58% of the variance in BF. Two of the identified SNPs (rs6738627 and rs2943650) have previously been associated with favourable adiposity[@Yaghootkar2014; @Yaghootkar2016], which has been associated with increased BF, BMI, fat mass, and reduced fat free mass while simultaneously being associated with reduced risk of type-2 diabetes, hypertension, and heart disease, as well as more favourable blood pressure[@Yaghootkar2016]. In additional analyses conducted in Chapter \@ref(MR), comparison of the effect of BF on metabolites, instrumenting BF with the 7 SNPs identified by Lu et al. and a 5-SNP instrument that did not include the two favourable adiposity SNPs, resulted in tighter confidence intervals (CIs) for the 5-SNP instrument. Additionally, a number of effect estimates reversed in direction when instrumenting BF using the 5 SNPs. As these favourable adiposity SNPs are likely adding noise to the instrumentation of BF, the 5-SNP instrument was used herein. The variance explained by these 5 SNPs will be lower than the variance explained by the 7 SNPs in the all ancestries meta-analysis. \par

#### Intermediates: Metabolites
A total of 54 metabolites were identified as being associated with adiposity in Chapter \@ref(MR); two metabolites, associated with BMI and WHR (serum total triglycerides) and BF (estimated description of fatty acid chain length not actual carbon number) were not available in the UK Biobank GWAS. Female-specific summary statistics for each metabolite was obtained from UK Biobank (Borges 2021, unpublished), in which, 118,466 women of European ancestries were included in a linear mixed model GWAS. UK Biobank is a prospective cohort study of ~500,000 individuals from the United Kingdom aged 37--70 with a host of genetic and phenotypic data[@Sudlow2015; @Bycroft2018]. Un-fasted individuals were selected at random to undergo high-throughput nuclear magnetic resonance (NMR) metabolomic analysis[@Julkunen2021] using the Nightingale Health Ltd biomarker quantification[@Soininen2009; @Wurtz2017] (version 2020); the same platform for which metabolomic data were avaialble from the Avon Longitudinal Study of Parents and Children (ALSPAC) used in Chapter \@ref(observational) and datasets used in Chapter \@ref(MR). All metabolites were inverse rank normally transformed prior to genome-wide analysis. Genome-wide association analysis was performed using the Medical Research Council, Integrative Epidemiology Unit (MRC IEU) UK Biobank GWAS pipeline[@Mitchell2019]. A linear mixed model using BOLT-LMM, adjusting for genotype array and fasting time was fit for 118,466 individuals. Population structure was controlled for using 143,006 directly genotyped SNPs (minor allele frequency > 0.01; genotyping rate > 0.015; Hardy-Weinberg equilibrium p-value < 0.0001 and linkage disequilibrium (LD) pruning to an R^2^ threshold of 0.1 using PLINKv2.00). Borges et al., had not identified lead SNPs at the time of writing. Lead SNPs were thus identified as those reaching a genome-wide significance threshold (p-value $\leq$ 5x10^-8^) and which were retained after clumping using a LD r^2^ threshold of 0.001 for SNPs within a 10,000 base window of each other and using the 1000 Genomes version 3 reference panel. \par

#### Outcomes: Endometrial cancer
Female-specific summary statistics for endometrial cancer were available from O'Mara et al. (2018)[@OMara2018]. This data included summary statistics on overall endometrial cancer as well as endometrioid and non-endometrioid cancer. Briefly, 17 studies on endometrial cancer including 12,906 cases and 108,979 country-matched controls of European ancestries were included in a fixed effects inverse variance weighted meta-analysis. Genotypes were imputed using the 1000 Genomes Project v3 reference panel or a combined 1000 Genomes (V3) UK10K reference panel. In each study, univariate GWAS were conducted adjusting for principal components (PCs). The analysis was repeated for endometrioid only (cases = 8,758; controls = 108,979) and non-endometrioid only (case = 1,230; controls = 108,979) cancers. Overall endometrial cancer included the cases from the endometrioid and non-endometrioid cancers alongside a number of other unclassified cases. Of the 12,906 cases and 108,979 controls for overall endometrial cancer, 636 cases (`r round(636/12906*100)`%) and 62,853 controls (`r round(62853/108979*100)`%) were from UK Biobank. For endometrioid and non-endometrioid cancer, no cases and 62,853 controls (`r round(62853/108979*100)`%) were from UK Biobank. \par

### Two-sample univariable Mendelian randomization {#mediation-methods-2smr}
Summary statistics for the association between genetic variation and BMI[@Locke2015], WHR[@Shungin2015], and BF[@Lu2016] were obtained from published GWAS, as described above, all genetic instruments are available on [GitHub](https://github.com/mattlee821/000_thesis/blob/master/index/data/mediation/tables/adiposity_snps.txt). As summary statistics of the association between genetic variation on metabolites in UK Biobank were unpublished, all genetic instruments are available on [GitHub](https://github.com/mattlee821/000_thesis/blob/master/index/data/mediation/analysis/003_metabolite_endometrial/exposure_data.txt), alongside a list of metabolites [GitHub](https://github.com/mattlee821/000_thesis/blob/master/index/data/mediation/tables/metabolites.txt) which is also available in the Appendix (Table \@ref(tab:appendix-mediation-table-metabolites)). \par

For all exposures, the following summary-level data were obtained from the original GWAS publications for each exposure-related genetic variant: rsID, effect allele, other/non-effect allele, effect allele frequency, effect estimate, standard error of the effect estimate, p-value, sample size, and units. All exposure-related genetic variants were obtained for each outcome separately. Genetic variants were extracted from each outcome GWAS and, where these were not present, proxy SNPs were included if LD was $\geq$ 0.8. For proxy SNPs, the inclusion of SNPs where the reference strand was ambiguous (strand flips) was allowed and the reference strand was inferred using a MAF threshold. That is, the reference strand was inferred using a MAF, so long as that MAF was not $\geq$ 0.3, in which case it was excluded. Exposure and outcome summary statistics for each of the exposure-related SNPs were harmonised in reference to the exposure effect allele being on the increasing scale. For included alleles where the reference strand was ambiguous, the positive strand was inferred using effect allele frequency. That is, if the effect allele frequency of a SNP was not $\geq$ 0.3 or $\leq$ 0.7, the reference strand was inferred using the effect allele frequency to harmonise exposure and outcome data; otherwise, it was removed. \par

Instruments for adiposity measures were not clumped as the studies from which they were obtained stated they were independent or near-independent and analysis in Chapter \@ref(MR) indicated clumping did not considerably alter associations with metabolites. Instruments for metabolites were identified using a genome-wide significance threshold of p-value $\leq$ 5x10^-8^ and an LD r^2^ threshold of 0.001 for SNPs within a 10,000 base window of each other and using the 1000 Genomes reference panel using the `TwoSampleMR`[@Hemani2018] `R`. To assess the possibility of weak instrument bias, F-statistics were calculated for each exposure-related SNP and an average F-statsistic was calculated for each exposure. \par

An inverse variance weighted (IVW), multiplicative random effects (IVW-MRE) model was used to estimate the effect of each exposure on the outcome. The model assumes that the strength of the association of the genetic instruments with the exposure is not correlated with the magnitude of the pleiotropic effects and that the pleiotropic effects have an average value of zero[@Bowden2017]. Where the number of instruments was not sufficient for an IVW-MRE model, the Wald ratio was used. As analysis presented in this chapter, which aimed to assess the potentially intermediary role of metabolites in the relationship between adiposity and endometrial cancer, aimed to validate results presented in Chapters \@ref(systematic-review), \@ref(observational), and \@ref(MR), no multiple testing threshold was set. Both the adiposity measures and metabolites were inverse rank normally transformed prior to genome-wide analysis and represent standard deviation (SD) units. For MR analyses using these measures, effect estimates are given in SD units. Results testing for the causal effect of both adiposity and metabolites on endometrial cancer are given as odds ratios (OR) and represent the odds of developing endometrial cancer per SD unit increase in the exposure. \par

#### Sensitivity analysis
Where possible (i.e., where there were three or more instruments), the assumptions of no pleiotropy among genetic instruments and outcomes were explored using: MR-Egger[@Bowden2015], weighted median[@Burgess2017a], and weighted mode[@Hartwig2017] based estimators. These methods are sensitive to the effects of potential pleiotropy. No p-value threshold requirements were set for these methods, instead consistency between the IVW-MRE model and these methods was investigated. Briefly, MR-Egger provides an estimate of unbalanced or directional horizontal pleiotropy via the intercept of a linear regression of the SNP-exposure and SNP-outcome association. In the presence of pleiotropy, the intercept will be biased away from the origin. MR-Egger gives consistent estimates when 100% of genetic instruments are invalid[@Bowden2015]. The weighted median is complimentary to MR-Egger but does not rely on the "instrument strength independent of direct effect" (InSIDE) assumption. It calculates the median of an empirical distribution of the causal effect estimates weighted for precision. It provides consistent estimates when at least 50% of the weight comes from valid genetic instruments and as long as no one genetic instrument contributes > 50% of the weight[@Burgess2017a]. The weighted mode assumes the true causal effect is the most common effect and it is robust when the majority of effect estimates are derived from valid instruments[@Hartwig2017]. \par

### Two-sample multivariable Mendelian randomization
All adiposity measures were included in the MVMR analysis. Of the metabolites included in the two-sample univariable MR analysis estimating the causal effect of adiposity on metabolites and of metabolites on endometrial cancer, only those which showed a consistent direction of effect with both adiposity and endometrial cancer were included in the MVMR analyses. That is, if an adiposity measure increased a metabolite, that metabolite increased endometrial cancer, and taht adiposity measure increased endometrial cancer, then the metabolite was considered as a potential intermediate. Or, if an adiposity measure decreased a metabolite, that metabolite decreased endometrial cancer, and that adiposity measure increased endometrial cancer, then the metabolite was considered a potential intermediate. Otherwise, the metabolite was excluded from MVMR analyses. \par

In MVMR, SNPs associated with the exposure (adiposity) and proposed intermediate (metabolites) were combined to create a combined instrument of the exposure and intermediate. This combined SNP list was used to extract instruments from both GWASs of the exposure and the intermediate metabolite. The resulting instrument was then clumped to remove duplicate SNPs and, SNPs in LD with one another to avoid overestimation of effects using the same clumping thresholds as with the univariable analysis described above (LD r^2^ > 0.001 and 10,000 base window). \par

The resulting instrument contains SNPs associated with both the exposure and intermediate; extracting data from the GWAS of the exposure gives an instrument for the exposure adjusted for the intermediate, and extracting data from the GWAS of the intermediate gives an instrument for the intermediate adjusted for the exposure (Figure \@ref(fig:mediation-figure-mvmr-dag)). As such, instruments for BMI and instruments for metabolites will use the same SNPs but different estimates. For example, a BMI instrument adjusted for metabolite 1 will include the SNPs associated with BMI and the SNPs associated with metabolite 1, with all data extracted from the GWAS of BMI. An instrument for metabolite 1 adjusted for BMI will include the SNPs associated with BMI and the SNPs associated with metabolite 1, with all data extracted from the GWAS of metabolite 1. The following process was followed for obtaining instruments:\par

1. SNPs associated with each adiposity measure and each metabolite were identified in the same way as for the two-sample univariable MR analysis described previously (\@ref(mediation-methods-2smr)).
2. Identified SNPs for each adiposity and metabolite pair were combined to create a SNP list for each adiposity and metabolite pair.
3. These SNP lists were used to extract summary level data for instruments (rsID, effect allele, other/non-effect allele, effect allele frequency, effect estimate, standard error of the effect estimate, p-value, N, and units) from the GWASs of each adiposity measure and each metabolite individually.
4. Each instrument was clumped using the `TwoSampleMR`[@Hemani2018] `R` package setting an LD R^2^ threshold of 0.001 for SNPs within a 10,000 base window of each other and using the 1000 Genomes reference panel.
5. Data for each adiposity and metabolite pair were harmonised in reference to the adiposity-increasing effect allele in the adiposity instrument .
6. Instruments for each adiposity and metabolite pair were extracted from the endometrial cancer GWAS using MR-Base. Where SNPs were not present in the outcome GWAS, proxy SNPs were included if LD was $\geq$ 0.8. For proxy SNPs, the inclusion of SNPs where the reference strand was ambiguous (strand flips) was allowed and the reference strand was inferred using a MAF threshold of 0.3.
7. Exposure-related SNPs from summary level data of the outcome were harmonised in reference to the adiposity-increasing effect allele in the adiposity instrument instrument.

(ref:mediation-figure-mvmr-dag-cap) **Directed acyclic graph of the multivariable Mendelian randomization principle using two exposures**. The directed acyclic graph illustrates the principle of multivariable Mendelian randomization (MR) using an exposure (X1) and a mediator (X2) on a single outcome. Genetic instruments (G) associated with X1 and X2 are combined into an instrument. In this example, the instrument includes G1, G2, and any number of other SNPs (Gn). The instrument is extracted from summary statistics for each of X1 and X2. G = single nucleotide polymorphism; X1 = exposure, X2 = mediator; Y = outcome; U = unmeasured confounding

(ref:mediation-figure-mvmr-dag-scap) Directed acyclic graph of multivariable MR principle

```{r mediation-figure-mvmr-dag, echo=FALSE, warning=FALSE, error=FALSE, fig.align='center', out.width='75%',fig.cap='(ref:mediation-figure-mvmr-dag-cap)', fig.scap='(ref:mediation-figure-mvmr-dag-scap)', warning=FALSE, error=FALSE, message=FALSE}
library(ggdag)
# dags ====
theme_set(theme_dag())
## mr dag 3
# set coordinates
coords <- tibble::tribble(~name, ~x, ~y,
                          "G1", 0,0,
                          "G2", 0,-0.5,
                          "Gn", 0,-1,
                          "X1", 1,0,
                          "X2", 1,-1,
                          "Y", 2,0,
                          "U", 1.5,1)
# make dag ====
dag <- dagify(X1 ~ G1,
              X2 ~ G1,
              X1 ~ G2,
              X2 ~ G2,
              X1 ~ Gn,
              X2 ~ Gn,
              X2 ~ X1,
              Y ~ X1,
              Y ~ X2,
              X1 ~U, 
              X2 ~ U,
              Y ~ U,
              coords = coords)
ggdag(dag, text = TRUE, edge_type = "link_arc", stylized = FALSE)
```

For MVMR analysis, the SNP, effect estimate, and standard error for each exposure (adiposity and metabolite) and outcome are required. Instrument strength for each exposure was estimated using a generalized version of Cochran's Q[@Sanderson2019b; @Sanderson2021a], using the `strength_mvmr()` function in the `MVMR` `R` package assuming a pairwise covariance of 0. The assumption being that the SNP-exposure association and the SNP-intermediate association are estimated in independent samples (e.g., two different GWAS of independent populations). Horizontal pleiotropy was evaluated using a modified form of Cochran's Q using the `pleiotropy_mvmr()` function in the `MVMR` `R` package and assuming a pairwise covariance of 0. \par

An IVW MVMR model was used to obtain the direct causal effect of each adiposity measure adjusted for each metabolite and each metabolite adjusted for each adiposity measure on endometrial cancer risk. The aim of this analysis was to inform whether the effect of adiposity on endometrial cancer is partly explained by the effect of adiposity on metabolites and the effect of those metabolites on endometrial cancer or whether any effects of adiposity and metabolites on endometrial cancer are independent. Results are presented as OR and represent the odds of developing endometrial cancer per SD unit increase in the exposure. \par

## Results
For this section, results are presented first for the two-sample univariable MR analyses of the effect of adiposity on endometrial cancer, the effect of adiposity on metabolites in UK Biobank, and the effect of those metabolites for which there was evidence for a causal effect of adiposity (consistent across analyses conducted in Chapter \@ref(MR) and UK Biobank here) and endometrial cancer. Finally, results from the two-sample MVMR analyses unpicking the potential intermediary role of these metabolites in the relationship between adiposity and endometrial cancer are presented. \par

### Two-sample univariable Mendelian randomization: association between adiposity measures and endometrial cancer {#mediation-results-adiposity-cancer}
```{r eval=TRUE, include=TRUE, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
# data 
data <- read.table("../../007_metabolites_outcomes/analysis/001_adiposity_endometrial/001_MR_results.txt", header = T, sep = "\t")
data$group[data$exposure == "Locke BMI EU sex combined 77 SNPs clumped"] <- "BMI"
data$group[data$exposure == "Shungin WHR EU sex combined 26 SNPs"] <- "WHR"
data$group[data$exposure == "Lu BF EU sex combined 5 SNPs"] <- "BF"
data$outcome_label[data$outcome == "Endometrial cancer (endometrioid histology) || id:ebi-a-GCST006465"] <- "Endometrioid"
data$outcome_label[data$outcome == "Endometrial cancer (Non-endometrioid histology) || id:ebi-a-GCST006466"] <- "Non-endometrioid"
data$outcome_label[data$outcome == "Endometrial cancer || id:ebi-a-GCST006464"] <- "Endometrial cancer"
data$method[data$method == "Inverse variance weighted (multiplicative random effects)"] <- "IVW-MRE"

data <- subset(data, method == "IVW-MRE")
bmi <- subset(data, group == "BMI")
whr <- subset(data, group == "WHR")
bf <- subset(data, group == "BF")
```

Here, using the largest endometrial cancer GWAS to date, the MR analyses performed as part of this thesis provided strong evidence of a causal effect of BMI on endometrial cancer (OR per SD unit increase in BMI = `r round(bmi$OR[1],2)`; 95% CI = `r round(bmi$lower_ci[1],2)`--`r round(bmi$upper_ci[1],2)`; p-value = 1.74 x 10^-15^; (Table \@ref(tab:mediation-table-adiposity-cancer)). A similar effect (OR = `r round(bmi$OR[2],2)`; 95% CI = `r round(bmi$lower_ci[2],2)`--`r round(bmi$upper_ci[2],2)`; p-value = 1.59 x 10^-13^) was found for endometrioid cancer, while a smaller effect was observed for non-endometrioid cancer (OR = `r round(bmi$OR[3],2)`; 95% CI = `r round(bmi$lower_ci[3],2)`--`r round(bmi$upper_ci[3],2)`; p-value = `r round(bmi$pval[3],2)`). \par

The effect of WHR on overall endometrial cancer and endometrioid cancer were directionally consistent but lesser in magnitude than results for BMI (OR = 1.2; 95% CI = 0.98 - 1.53 and 1.25; 95% CI = 0.93 - 1.58, respectively). The effect of BF on endometrial cancer (OR = `r round(bf$OR[1],2)`; 95% CI = `r round(bf$lower_ci[1],2)`--`r round(bf$upper_ci[1],2)`; p-value = 1.02 x 10^-16^), endometrioid (OR = `r round(bf$OR[2],2)`; 95% CI = `r round(bf$lower_ci[2],2)`--`r round(bf$upper_ci[2],2)`; p-value = 9.93 x 10^-7^), and non-endometrioid (OR = `r round(bf$OR[3],2)`; 95% CI = `r round(bf$lower_ci[3],2)`--`r round(bf$upper_ci[3],2)`; p-value = `r round(bf$pval[3],2)`) were consistently greater than those observed for BMI, though with wider CIs. \par

Sensitivity analyses, which included weighted median, weighted mode, and MR-Egger models, were consistent with the IVW-MRE model for each exposure (Appendix Figure \@ref(fig:appendix-mediation-figure-forestplot-adiposity-endometrial)). All figures associated with sensitivity analyses (e.g. leave-one-out analyses) are available on [GitHub](https://github.com/mattlee821/000_thesis/tree/master/index/data/mediation/figures/001_adiposity_endometrial) with representative figures presented in the appendix (Appendix \@ref(appendix-mediation)). All results are available on [GitHub](https://github.com/mattlee821/000_thesis/tree/master/index/data/mediation/results/001_adiposity_endometrial). \par

```{r mediation-table-adiposity-cancer, eval=TRUE, include=TRUE, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
table <- data[,c("group", "outcome_label", "OR", "lower_ci", "upper_ci", "pval")]
table$pval <- signif(table$pval,3)
table$pval <- as.character(table$pval)
knitr::kable(table, longtable = T, booktabs = T, format = "latex", escape = F,
    caption = 'Two-sample univariable Mendelian randomization results of the association between adiposity measures and endometrial cancer', row.names = F, digits = 2,
    col.names = c("","", "OR", "Lower CI", "Upper CI", "p-value")) %>%
  kable_styling(latex_options = c("striped", "HOLD_position", "repeat_header"), position = "center") %>%
  kableExtra::footnote(general = "Results for inverse variance weighted multiplicative random effects model are presented. Odds ratios (OR) and associated 95% confidence intervals (CI) per standard deviation unit increase in body mass index (BMI), waist hip ratio (WHR), or body fat percentage (BF)", threeparttable = T, general_title = "", footnote_as_chunk = T)

```

### Two-sample univariable Mendelian randomization: association between adiposity measures and metabolites {#mediation-results-adiposity-metabolites}
A total of 54 metabolites were identified as being associated with BMI (N = 45), WHR (N = 48), and BF (N = 9) in Chapter \@ref(MR). Two of these metabolites, associated with BMI and WHR (serum total triglycerides) and BF (estimated description of fatty acid chain length not actual carbon number) were not available in the UK Biobank GWAS. As such, a total of 44, 47, and 8 metabolites, which were associated with BMI, WHR, and BF in Chapter \@ref(MR), respectively, were available for analysis in UK Biobank. \par

For BMI, of the 44 metabolites, four had directions of effect that were not consistent with the previous MR analysis (Chapter \@ref(MR)), meta-analysis (Chapter \@ref(MR)), and observational (Chapter \@ref(observational)) analyses: apolipoprotein B, free cholesterol in medium very large low density lipoprotein (VLDL), total lipids in medium VLDL, total lipids in very small VLDL. For WHR, of the 47 metabolites, two had directions of effect that were not consistent with the previous MR analysis (Chapter \@ref(MR)), meta-analysis (Chapter \@ref(MR)), and observational (Chapter \@ref(observational)) analyses: phenylalanine and tyrosine. For BF, all 8 of the analysed metabolites had consistent directions of effect with the previous MR analysis (Chapter \@ref(MR)), meta-analysis (Chapter \@ref(MR)), and observational (Chapter \@ref(observational)) analyses. Sensitivity analyses did not highlight metabolite analyses which violated the MR assumptions. As such, a total of 40, 45, and 8 metabolites (53 unique metabolites) were taken forward for MR analysis to estimate the effect of these metabolites on endometrial cancer outcomes. Seven of these metabolites were only found to be associated with WHR and 5 of these metabolites were uniquely associated with BF. Two metabolites were associated with both BMI and BF and 38 metabolites were associated with both BMI and WHR. Only 1 metabolite, valine, was associated with all three measures. \par

All results for the investigation of the effect of adiposity measures on metabolites measured in UK Biobank are presented in Figure \@ref(fig:mediation-figure-forestplot-adiposity-metabolites) and are available on [GitHub](https://github.com/mattlee821/000_thesis/tree/master/index/data/mediation/results/002_adiposity_metabolite). All sensitivity plots are available on [GitHub](https://github.com/mattlee821/000_thesis/tree/master/index/data/mediation/figures/002_adiposity_metabolite) and representative figures are presented in the appendix (Appendix \@ref(appendix-mediation)). \par

\newpage
\thispagestyle{empty}
\vspace*{-3cm}

(ref:mediation-figure-forestplot-adiposity-metabolites-cap) **Two-sample univariable Mendelian randomization analysis: association between adiposity measures and metabolites using UK Biobank data**. Effect estimates and associated 95% confidence intervals shown for an standard deviation (SD) change in metabolite per SD unit higher body mass index (BMI), waist hip ratio (WHR), or body fat percentage (BF). Solid points indicate a nominal p-value threshold of 0.05 has been reached. Available on [GitHub](https://github.com/mattlee821/000_thesis/blob/master/index/data/mediation/figures/002_adiposity_metabolite/forestplot.pdf).

(ref:mediation-figure-forestplot-adiposity-metabolites-scap) Two-sample univariable Mendelian randomization analysis: association between adiposity measures and metabolites using UK Biobank data

```{r mediation-figure-forestplot-adiposity-metabolites, echo=FALSE, fig.align='center', out.width='100%', fig.cap='(ref:mediation-figure-forestplot-adiposity-metabolites-cap)', fig.scap='(ref:mediation-figure-forestplot-adiposity-metabolites-scap)'}
knitr::include_graphics("data/mediation/figures/002_adiposity_metabolite/forestplot.pdf")
```

### Two-sample univariable Mendelian randomization: association between adiposity-associated metabolites and endometrial cancer {#mediation-results-metabolites-cancer}
```{r eval=TRUE, include=TRUE, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
data <- read.table("../../007_metabolites_outcomes/analysis/003_metabolite_endometrial/mr_results_formatted_associated_metabolites.txt", header = T, sep = "\t")
unique_metabolites <- unique(data$metabolite)
n_metabolites <- read.table("../../007_metabolites_outcomes/analysis/003_metabolite_endometrial/total_associated_metabolites.txt", header = T, sep = "\t")
metabs <- n_metabolites[,1]
SNPS <- read.table("../../007_metabolites_outcomes/data/metabolite_instruments.txt", header = T, sep = "\t")
SNPS <- SNPS[SNPS$exposure %in% metabs,]
n_SNPS <- SNPS[!duplicated(SNPS$SNP), ]

freq_SNP <- SNPS %>% 
  group_by(SNP) %>% 
  mutate(count=n())
freq_SNP <- freq_SNP[!duplicated(freq_SNP$SNP), ]
freq_one_SNP <- subset(freq_SNP, count == 1)


# venndiagram of metab overlap
# data <- read.table("../007_metabolites_outcomes/analysis/002_adiposity_metabolite/001_mr_results.txt", header = T, sep = "\t")
# bmi <- subset(data, exposure == "bmi")
# bmi <- bmi$metabolit
# bmi <- bmi[order(bmi)]
# 
# whr <- subset(data, exposure == "whr")
# whr <- whr$metabolit
# whr <- whr[order(whr)]
# 
# bf <- subset(data, exposure == "bf")
# bf <- bf$metabolit
# bf <- bf[order(bf)]
# 
# x <- list(bmi=bmi,whr=whr,bf=bf)
# v0 <- venn.diagram(x, filename=NULL, 
#                     fill = c("red", "blue", "green"),
#                     alpha = 0.50,
#                     col = "transparent")



```

Two-sample univariable MR analysis was used to investigate the causal effect of `r nrow(n_metabolites)` metabolites (i.e., 40, 45, and 8 metabolites driven by BMI, WHR, and BF, respectively) on endometrial cancer.  Instruments for metabolites were obtained from an unpublished GWAS of 249 circulating metabolites performed in UK Biobank (Borges 2021, unpublished). Using a genome-wide p-value threshold of 5x10^-8^ and an LD r^2^ clumping threshold (0.001) for SNPs within a 10,000 base window, to ensure independance of SNPs, `r nrow(SNPS)` total associations with 53 metabolites were identified (unique number of SNPs = `r nrow(n_SNPS)`; minimum number of SNPs associated with a metabolite = `r min(table(SNPS$exposure))`, maximum number of SNPs associated with a metabolite = `r max(table(SNPS$exposure))`; Figure \@ref(fig:mediation-figure-metabolites-snp-distribution)). The greater the number of SNPs associated with a trait, the larger the variance explained by those SNPs and thus the power afforded by that instrument in an MR analysis. Of the `r nrow(n_SNPS)` unique SNPs identified, `r round(nrow(freq_one_SNP)/nrow(freq_SNP)*100,0)`% of SNPs were associated with just 1 metabolite (Figure \@ref(fig:mediation-figure-metabolites-snp-numbers)). The remaining `r 100-round(nrow(freq_one_SNP)/nrow(freq_SNP)*100,0)`% of SNPs were associated with two or more metabolites, with 1 SNP (rs12154627) associated with `r max(freq_SNP$count)` metabolites. Metabolites were inverse rank normally transformed prior to genome-wide analysis and represent SD units. \par

\newpage
\thispagestyle{empty}
\vspace*{-1cm}

(ref:mediation-figure-metabolites-snp-distribution-cap) **Distribution of the number of SNPs with which metabolites were associated**. A total of `r nrow(n_SNPS)` unique single nucleotide polymorphisms (SNPs) were identified across 53 metabolites. The number of SNPs associated with each metabolite varied between `r min(table(SNPS$exposure))` and `r max(table(SNPS$exposure))` with a mean of `r round(mean(table(SNPS$exposure)))`. Bin size = 10.

(ref:mediation-figure-metabolites-snp-distribution-scap) Distribution of the number of SNPs with which metabolites were associated.

```{r mediation-figure-metabolites-snp-distribution, echo=FALSE, out.width='80%', fig.align='center',  fig.cap='(ref:mediation-figure-metabolites-snp-distribution-cap)', fig.scap='(ref:mediation-figure-metabolites-snp-distribution-scap)'}
data <- SNPS %>% 
  group_by(exposure) %>% 
  mutate(count=n())
data <- data[!duplicated(data$exposure), ]

ggplot(data, 
       aes(x = count)) + 
  geom_histogram(binwidth = 10, alpha = 1,
                 color = discrete_wes_pal[1],
                 fill = discrete_wes_pal[1]) + 
  xlab("Total number of SNPs a metabolites is associated with") + ylab("Frequency") +
  theme_cowplot()
```

(ref:mediation-figure-metabolites-snp-numbers-cap) **Distribution of the number of metabolites with which individual SNPs were associated**. The majority of the `r nrow(n_SNPS)` identified single nucleotide polymorphisms (SNPs) were associated with one metabolite. One SNP was associated with 33 out of a total 53 metabolites. Bin size = 1.

(ref:mediation-figure-metabolites-snp-numbers-scap) Distribution of the number of metabolites with which individual SNPs were associated

```{r mediation-figure-metabolites-snp-numbers, echo=FALSE, out.width='80%', fig.align='center', fig.cap='(ref:mediation-figure-metabolites-snp-numbers-cap)', fig.scap='(ref:mediation-figure-metabolites-snp-numbers-scap)'}
ggplot(freq_SNP, 
       aes(x = count)) + 
  geom_histogram(binwidth = 1, alpha = 1,
                 color = discrete_wes_pal[1],
                 fill = discrete_wes_pal[1]) + 
  xlab("Total number of metabolites an individual SNP is associated with") + ylab("Frequency") +
  theme_cowplot()
```

\newpage

```{r eval=TRUE, include=TRUE, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
data <- read.table("../../007_metabolites_outcomes/analysis/003_metabolite_endometrial/mr_results_formatted_associated_metabolites.txt", header = T, sep = "\t")
data <- select(data, c("metabolite", "ng_anno_raw.label", "outcome", "nsnp", "b", "se", "pval", "method"))
data$OR <- exp(data$b)
data$lower_ci <- exp(data$b - (1.96 * data$se))
data$upper_ci <- exp(data$b + (1.96 * data$se))
data <- subset(data, method == "Inverse variance weighted (multiplicative random effects)")
valine <- subset(data, metabolite == "val")
```

A total of `r length(unique_metabolites)` metabolites showed evidence of association for at least one outcome (Figure \@ref(fig:mediation-figure-forestplot-metabolites-cancer)). Valine was the only metabolite found to be negatively associated with the risk of endometrial cancer (OR of endometrial cancer per SD unit increase in valine = `r round(valine$OR[1],2)`; 95% CI = `r round(valine$lower_ci[1],2)`--`r round(valine$upper_ci[1],2)`; p-value = `r round(valine$pval[1],2)`), which was directionally consistent with both endometrioid and non-endometrioid cancers. The largest effect was found for triglycerides in small VLDL, which increased the odds of non-endometrioid cancer by `r round(data$OR[3],2)` (95% CI = `r round(data$lower_ci[3],2)`--`r round(data$upper_ci[3],2)`; p-value = 0.001), which was directionally consistent with results for overall endometrial cancer and endometrioid cancer. The largest effect on endometrioid cancer was for phospholipids in very large high density lipoprotein (HDL; OR = `r round(data$OR[11],2)`; 95% CI = `r round(data$lower_ci[11],2)`--`r round(data$upper_ci[11],2)`; p-value = 0.04), results of which were consistent for overall endometrial cancer but opposite to those found in non-endometrioid cancer (though, CIs overlapped all analysis results and the null). The largest effect on overall endometrial cancer was for triglycerides in very large HDL (OR = `r round(data$OR[7],2)`; 95% CI = `r round(data$lower_ci[7],2)`--`r round(data$upper_ci[7],2)`;2; p-value = 0.06), which was directionally consistent with both endometrioid and non-endometrioid cancers. For triglycerides in very small VLDL, effects were largest for non-endometrioid cancer (OR = `r round(data$OR[6],2)`; 95% CI = `r round(data$lower_ci[6],2)`--`r round(data$upper_ci[6],2)`;; p-value = 0.008), which was directionally consistent with results for overall endometrial cancer and endometrioid cancer. \par

Of the 5 metabolites with evidence of an association with endometrial cancer, endometrioid cancer, and/or non-endometrioid cancer, only triglycerides in very large HDL was associated with BF in the previous analysis conducted in this chapter and in observational (Chapter \@ref(observational)) and MR (Chapter \@ref(MR)) analyses. The remaining four metabolites were not associated with BF in any analysis, but were all associated with BMI and WHR in the previous analysis here and in observational (Chapter \@ref(observational)) and MR (Chapter \@ref(MR)) analyses. As all 5 metabolites showed evidence of association with a measure of adiposity and endometrial cancer, all were considered for MVMR analysis. All results are available on [GitHub](https://github.com/mattlee821/000_thesis/tree/master/index/data/mediation/results/003_metabolite_endometrial) and presented in Appendix Figure \@ref(fig:appendix-mediation-figure-forestplot-metabolites-cancer) which is also available on [GitHub](https://github.com/mattlee821/000_thesis/blob/master/index/data/mediation/figures/003_metabolite_endometrial/forestplot_all.pdf). Figures for sensitivity analyses are also on [GitHub](https://github.com/mattlee821/000_thesis/tree/master/index/data/mediation/figures/003_metabolite_endometrial). \par

(ref:mediation-figure-forestplot-metabolites-cancer-cap) **Two-sample univariable Mendelian randomization analysis: association between metabolites and endometrial cancer using UK Biobank data**. Forest plot shows odds ratio (OR) and 95% confidence interval for metabolites associated with body mass index, waist hip ratio, and/or body fat percentage with endometrial cancer, endometrioid cancer, and non-endometrioid cancer. Data are presented for endometrial-associated metabolites. The main analysis (inverse variance weighted multiplicative random effects (IVW-MRE)) is presented alongside sensitivity analyses (weighted median, weighted mode, MR-Egger). Available on [GitHub](https://github.com/mattlee821/000_thesis/blob/master/index/data/mediation/figures/003_metabolite_endometrial/forestplot.pdf).

(ref:mediation-figure-forestplot-metabolites-cancer-scap) Two-sample univariable Mendelian randomization analysis: association between metabolites and endometrial cancer using UK Biobank data

```{r mediation-figure-forestplot-metabolites-cancer, echo=FALSE, out.width='100%', fig.cap='(ref:mediation-figure-forestplot-metabolites-cancer-cap)', fig.scap='(ref:mediation-figure-forestplot-metabolites-cancer-scap)'}
knitr::include_graphics("../../007_metabolites_outcomes/analysis/003_metabolite_endometrial/figures/forestplot.pdf")
```

### Two-sample multivariable Mendelian randomization: intermediate effects of adiposity-associated metabolites in the association between adiposity measures and endometrial cancer
```{r eval=TRUE, include=TRUE, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
# data 
adiposity_metab <- read.table("../../007_metabolites_outcomes/analysis/002_adiposity_metabolite/001_mr_results.txt", header = T, sep = "\t")
adiposity_metab <- select(adiposity_metab, c("exposure", "ng_anno_raw.label", "metabolite", "UKB_b", "UKB_se", "UKB_p"))
colnames(adiposity_metab) <- c("exposure", "ng_anno_raw.label", "metabolite", "b", "se", "pval")
adiposity_metab_svldltg <- subset(adiposity_metab, metabolite == "svldltg")
adiposity_metab_xsvldltg <- subset(adiposity_metab, metabolite == "xsvldltg")

bf_metab <- read.table("../../007_metabolites_outcomes/analysis/002_adiposity_metabolite/bf/mr_results_formatted.txt", header = T, sep = "\t")
bf_metab <- subset(bf_metab, method == "Inverse variance weighted (multiplicative random effects)")
bf_metab <- select(bf_metab, c("exposure", "ng_anno_raw.label", "metabolite", "b", "se", "pval"))
bf_metab$exposure <- "bf"
bf_metab_svldltg <- subset(bf_metab, metabolite == "svldltg")
bf_metab_xsvldltg <- subset(bf_metab, metabolite == "xsvldltg")

adiposity_metab_svldltg <- rbind(adiposity_metab_svldltg,bf_metab_svldltg)
adiposity_metab_svldltg$lower_ci <- adiposity_metab_svldltg$b - (1.96 * adiposity_metab_svldltg$se)
adiposity_metab_svldltg$upper_ci <- adiposity_metab_svldltg$b + (1.96 * adiposity_metab_svldltg$se)

adiposity_metab_xsvldltg <- rbind(adiposity_metab_xsvldltg,bf_metab_xsvldltg)
adiposity_metab_xsvldltg$lower_ci <- adiposity_metab_xsvldltg$b - (1.96 * adiposity_metab_xsvldltg$se)
adiposity_metab_xsvldltg$upper_ci <- adiposity_metab_xsvldltg$b + (1.96 * adiposity_metab_xsvldltg$se)

```

Of the 5 metabolites for which there was causal evidence of an effect of at least one measure of adiposity and an effect on endometrial cancer, two had directions of effect that were consistent with a potential intermediate role: triglycerides in small VLDL and triglycerides in very small VLDL. Of the remaining three metabolites, all three adiposity measures were found to decrease phospholipids in very large HDL and triglycerides in very large HDL, while an increase in these metabolites was associated with an increase in all three endometrial cancer outcomes. The remaining metabolite, valine, was found to be increased by all three adiposity measures, however increased valine was associated with a decreased risk of all three endometrial cancer outcomes. As such, triglycerides in small VLDL and triglycerides in very small VLDL were taken forward for MVMR analysis. Using individual level data from ALSPAC adults used in Chapter \@ref(observational) (N = 3,305; data not shown), the Spearman's Rho correlation between the two metabolites was 0.9. \par

Results from the two-sample univariable MR analysis of the association between adiposity measures and triglycerides in small and very small VLDL (Section \@ref(mediation-results-adiposity-metabolites)) are presented in Table \@ref(tab:mediation-table-mvmr-adiposity-metabolites). In these analyses, the strongest effect between adiposity measures and triglycerides in small VLDL was found for WHR (SD unit increase in metabolite per SD increase in adiposity measure = `r round(max(adiposity_metab_svldltg$b),2)`; 95% CI = `r round(adiposity_metab_svldltg$lower_ci[2],2)`--`r round(adiposity_metab_svldltg$upper_ci[2],2)`; p-value = `r signif(adiposity_metab_svldltg$pval[2],2)`). The strongest effect between adiposity measures and triglycerides in very small VLDL was also found for WHR (beta = `r round(max(adiposity_metab_xsvldltg$b),2)`; 95% CI = `r round(adiposity_metab_xsvldltg$lower_ci[2],2)`--`r round(adiposity_metab_xsvldltg$upper_ci[2],2)`; p-value = `r signif(adiposity_metab_xsvldltg$pval[2],2)`). The effects of BMI (effect on small VLDL = `r round(adiposity_metab_svldltg$b[1],2)`; effect on very small VLDL = `r round(adiposity_metab_xsvldltg$b[1],2)`) and BF (effect on small VLDL = `r round(adiposity_metab_svldltg$b[3],2)`; effect on very small VLDL = `r round(adiposity_metab_xsvldltg$b[3],2)`) on both metabolites were smaller with CIs that overlapped the null. \par

\newpage

```{r mediation-table-mvmr-adiposity-metabolites, eval=TRUE, include=TRUE, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
table <- read.table("data/mediation/tables/mvmr_adiposity_metabolites.txt", header = T, sep = "\t", colClasses = "character")
table <- table[,c(1,2,4,7,8,6)]
knitr::kable(table, longtable = T, booktabs = T, format = "latex", escape = F,
    caption = 'Two-sample univariable Mendelian randomization: association between adiposity measures and triglycerides in small and very small VLDL', row.names = F,
    col.names = c("Exposure", "Metabolite (units)", "Effect", "Lower CI", "Upper CI", "p-value")) %>%
  kable_styling(latex_options = c("striped", "HOLD_position"), position = "center") %>%
  kableExtra::footnote(general = "Results are given for the inverse variance weighted multiplicative random effects model. Effect estimates are given as SD unit increase in metabolite per SD unit increase in adiposity measure. Metabolite labels are presented with their originally measured units. BMI = body mass index; WHR = waist hip ratio; BF = body fat percentage; VLDL = very large low density lipoprotein.", threeparttable = T, general_title = "", footnote_as_chunk = T)
```

```{r eval=TRUE, include=TRUE, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
data <- read.table("../../007_metabolites_outcomes/analysis/004_mvmr/mvmr_results_formatted.txt", header = T, sep = "\t")
data <- subset(data, method == "Two Sample MR")
data <- data[c(10:15),]
table <- data[,c("exposure", "outcome", "OR", "lower_ci", "upper_ci", "pval")]
table$exposure[table$exposure == "SVLDLTG"] <- "Triglycerides in small VLDL (mmol/l)"
table$exposure[table$exposure == "XSVLDLTG"] <- "Triglycerides in very small VLDL (mmol/l)"

table$OR <- signif(table$OR,2)
table$lower_ci <- signif(table$lower_ci,2)
table$upper_ci <- signif(table$upper_ci,2)
table$pval <- signif(table$pval,3)
```

Results from the two-sample univariable MR analysis of the association between triglycerides in small and very small VLDL with endometrial cancer (Section \@ref(mediation-results-metabolites-cancer)) are presented in Table \@ref(tab:mediation-table-mvmr-metabolites-endometrial). In these analyses, the strongest effect for both triglycerides in small (OR per SD unit increase in metabolite = `r table$OR[3]`; 95% CI = `r table$lower_ci[3]`--`r table$upper_ci[3]`; p-value = `r table$pval[3]`) and very small VLDL (OR = `r table$OR[6]`; 95% CI = `r table$lower_ci[6]`--`r table$upper_ci[6]`; p-value = `r table$pval[6]`) was for non-endometrioid cancer. Both metabolites had positive effects on endometrioid and overall endometrial cancer with highly similar effect sizes. However, effects were smaller than for non-endometrioid cancer and CIs overlapped the null. These results are presented alongside the MVMR results in Figure \@ref(fig:mediation-figure-forestplot-adiposity-metabolites-cancer). \par

\newpage

```{r mediation-table-mvmr-metabolites-endometrial, eval=TRUE, include=TRUE, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
table <- read.table("data/mediation/tables/mvmr_metabolites_cancer.txt", header = T, sep = "\t", colClasses = "character")
knitr::kable(table, longtable = T, booktabs = T, format = "latex", escape = F,
    caption = 'Two-sample univariable Mendelian randomization: association between triglycerides in small and very small VLDL and endometrial cancer', row.names = F, 
    col.names = c("Exposure", "Outcome", "OR", "Lower CI", "Upper CI", "p-value")) %>%
  kable_styling(latex_options = c("striped", "HOLD_position"), position = "center", font_size = 10) %>%
  kableExtra::footnote(general = "Results are given for the inverse variance weighted multiplicative random effects model. Effect estimates are given as the odds (OR) of endometrial cancer per SD unit increase in metabolite. Metabolite labels are presented with their originally measured units", threeparttable = T, general_title = "", footnote_as_chunk = T)
```

```{r eval=TRUE, include=TRUE, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
data <- read.table("../../007_metabolites_outcomes/analysis/004_mvmr/mvmr_results_formatted.txt", header = T, sep = "\t")
data$OR <- round(data$OR,2)
data$lower_ci <- round(data$lower_ci,2)
data$upper_ci <- round(data$upper_ci,2)

mr <- subset(data, method == "Two Sample MR")
mvmr <- subset(data, method == "Multivariable MR")

mvmr_bmi <- subset(mvmr, exposure == "BMI")
mvmr_bf <- subset(mvmr, exposure == "BF")
mvmr_whr <- subset(mvmr, exposure == "WHR")

mvmr_metab_bf <- subset(mvmr, adjusted == "BF")

mr_bmi <- subset(mr, exposure == "BMI")
mr_bf <- subset(mr, exposure == "BF")
mr_whr <- subset(mr, exposure == "WHR")
mr_svldltg <- subset(mr, exposure == "SVLDLTG")
mr_xsvldltg <- subset(mr, exposure == "XSVLDLTG")

endometrial <- subset(data, outcome == "Endometrial cancer")
endometrial_svldltg <- subset(endometrial, exposure == "SVLDLTG")
endometrioid <- subset(data, outcome == "Endometroid")
non_endometrioid <- subset(data, outcome == "Non-endometroid")
```

All results relevant to the MVMR analysis (this includes univariable two-sample MR results) are presented in Figure \@ref(fig:mediation-figure-forestplot-adiposity-metabolites-cancer) and Appendix Tables \@ref(tab:appendix-mediation-table-mvmr-bmi), \@ref(tab:appendix-mediation-table-mvmr-whr), and \@ref(tab:appendix-mediation-table-mvmr-bf) for BMI, WHR, and BF, respectively. A summary of the MVMR analysis is given herein. \par

In MVMR analysis, the effect of BMI on overall endometrial cancer was similar after adjustment for either metabolite: OR per SD unit increase in BMI adjusted for triglycerides in small VLDL = `r mvmr_bmi$OR[1]`; 95% CI = `r mvmr_bmi$lower_ci[1]`--`r mvmr_bmi$upper_ci[1]`; p-value = 5.18 x 10^-9^; OR per SD unit increase in BMI adjusted for triglycerides very small VLDL = `r mvmr_bmi$OR[4]`; 95% CI = `r mvmr_bmi$lower_ci[4]`--`r mvmr_bmi$upper_ci[4]`; p-value = 7.26 x 10^-8^. These results were similar to the unadjusted (i.e., two-sample univariable MR) effect of BMI on endometrial cancer (OR per SD unit increase in BMI = `r mr_bmi$OR[1]`; 95% CI = `r mr_bmi$lower_ci[1]`--`r mr_bmi$upper_ci[1]`; p-value = 7.47 x 10^-15^). A similar pattern of association was found across endometrioid and non-endometrioid cancer, with similar effect sizes and CIs after adjustment for each metabolite and when comparing effect size and CIs with the unadjusted effects. The effect for non-endometrioid cancer appeared to attenuate more, but CIs overlapped the two-sample univariable MR results. When looking at the effect of metabolites on all endometrial cancer outcomes after adjustment for BMI, there was little difference with effect size in comparison to the unadjusted effect, however CIs were generally wider in the adjusted estimates. For non-endometrioid cancer, the adjusted effect size was larger than the unadjusted effect size for both metabolites (though, CIs overlapped). Given the similarity between unadjusted and adjusted effects, there was little evidence for an intermediate role of either metabolite on the association between BMI and all endometrial cancer outcomes. \par

\blandscape
\thispagestyle{empty}

(ref:mediation-figure-forestplot-adiposity-metabolites-cancer-cap) **Multivariable Mendelian randomization: intermediate effects of adiposity-associated metabolites in the association between adiposity measures and endometrial cancer**. Forest plot shows odds ratio and 95% confidence interval for: the two-sample univariable Mendelian randomization (MR) analyses presented in this chapter (Section \@ref(mediation-results-adiposity-cancer) and \@ref(mediation-results-metabolites-cancer)) and the two-sample multivariable MR results of the intermediary effect of adiposity-associated metabolites and endometrial endometrial (left), endometrioid (middle) and non-endometrioid cancer (right). Points and confidence intervals are coloured for the variable they represent in the two-sample univariable MR and for the adjusted variable in the two-sample multivariable MR, e.g. body mass index (BMI) is coloured orange and any metabolite effect adjusted for BMI is coloured orange also. An inverse variance weighted model was used. BF = body fat percentage; WHR = waist hip ratio; SVLDLTG = triglycerides in small VLDL; XSVLDLTG = triglycerides in very small VLDL. Available on [GitHub](https://github.com/mattlee821/000_thesis/blob/master/index/data/mediation/figures/004_mvmr/forestplot.pdf).

(ref:mediation-figure-forestplot-adiposity-metabolites-cancer-scap) Multivariable Mendelian randomization: intermediate effects of adiposity-associated metabolites in the association between adiposity measures and endometrial cancer

```{r mediation-figure-forestplot-adiposity-metabolites-cancer, echo=FALSE, out.width='95%', fig.align='center', fig.cap='(ref:mediation-figure-forestplot-adiposity-metabolites-cancer-cap)', fig.scap='(ref:mediation-figure-forestplot-adiposity-metabolites-cancer-scap)'}
knitr::include_graphics("../../007_metabolites_outcomes/analysis/004_mvmr/figures/forestplot.pdf")
```

\elandscape

For the effect of BF and WHR on endometrial and endometrioid cancer, a similar pattern of association to that of BMI was seen, with effect sizes and CIs highly similar after adjustment with either metabolite. For example, the effect of WHR on endometrial cancer adjusted for triglycerides in small VLDL was `r mvmr_whr$OR[1]` (95% CI = `r mvmr_whr$lower_ci[1]`--`r mvmr_whr$upper_ci[1]`; p-value = `r round(mvmr_whr$pval[1],2)`) whereas after adjustment for triglycerides in very small VLDL the OR was `r mvmr_whr$OR[4]` (95% CI = `r mvmr_whr$lower_ci[4]`--`r mvmr_whr$upper_ci[4]`; p-value = `r round(mvmr_whr$pval[4],2)`). For both BF and WHR, the adjusted effects had much wider CIs than the unadjusted effects; for BF, adjustment for either metabolite resulted in CIs which crossed the null. For WHR, CIs crossed the null in unadjusted and adjusted analyses. When looking at the metabolites effects on endometrial and endometrioid cancer after adjustment for BF or WHR, there is a similar pattern of association to when adjusting for BMI, with similar sized effect sizes and CIs which overlapped the null. The exception was for the effect of triglycerides in small VLDL on overall endometrial cancer, where, when adjusted for BF (OR = `r mvmr_metab_bf$OR[1]`; 95% CI = `r mvmr_metab_bf$lower_ci[1]`--`r mvmr_metab_bf$upper_ci[1]`; p-value = `r round(mvmr_bf$pval[1],2)`) the CIs did not cross the null unlike in the unadjusted analysis (OR = `r mr_svldltg$OR[1]`; 95% CI = `r mr_svldltg$lower_ci[1]`--`r mr_svldltg$upper_ci[1]`; p-value = `r round(mr_svldltg$pval[1],2)`), but CIs overlapped between both analyses. \par

For non-endometrioid cancer, the pattern of association for BF and WHR was different to that of BMI. For BF, adjustment for triglycerides in small VLDL resulted in directions of effect that were opposite to the unadjusted (univariable) MR estimate (BF-adjusted OR = `r mvmr_bf$OR[3]`; 95% CI = `r mvmr_bf$lower_ci[3]`--`r mvmr_bf$upper_ci[3]`; p-value = `r round(mvmr_bf$pval[3],2)`; BF-unadjusted OR = `r mr_bf$OR[3]`; 95% CI = `r mr_bf$lower_ci[3]`--`r mr_bf$upper_ci[3]`; p-value = `r round(mr_bf$pval[3],2)`). The same was true for WHR (WHR-adjusted OR = `r mvmr_whr$OR[3]`; 95% CI = `r mvmr_whr$lower_ci[3]`--`r mvmr_whr$upper_ci[3]`; p-value = `r round(mvmr_whr$pval[3],2)`; WHR-unadjusted OR = `r mr_whr$OR[3]`; 95% CI = `r mr_whr$lower_ci[3]`--`r mr_whr$upper_ci[3]`; p-value = `r round(mr_whr$pval[3],2)`). In regards to triglycerides in very small VLDL, a similar picture was found for BF, where, the effect of BF adjusted for triglycerides in very small VLDL resulted in an effect estimate that was opposite to the unadjusted effect estimate (BF-adjusted OR = `r mvmr_bf$OR[6]`; 95% CI = `r mvmr_bf$lower_ci[6]`--`r mvmr_bf$upper_ci[6]`; p-value = `r round(mvmr_bf$pval[6],2)`). For WHR, adjustment for triglycerides in very small VLDL resulted in an attenuated effect estimate that was close to the null (WHR-adjusted OR = `r mvmr_whr$OR[6]`; 95% CI = `r mvmr_whr$lower_ci[6]`--`r mvmr_whr$upper_ci[6]`; p-value = `r round(mvmr_whr$pval[6],2)`). For all BF and WHR analyses, CIs overlapped for the adjusted and unadjusted effect estimates. When looking at the effect of triglycerides in small and very small VLDL on non-endometrioid cancer, adjustment for BF and WHR led to a greater increased risk of non-endometrioid cancer than was found in the unadjusted analysis. CIs for triglycerides in small and very small VLDL were wider in adjusted analyses compared to unadjusted analyses and overlapped. \par

Overall, the estimates of the effect of BMI, WHR, and BF adjusted for either metabolite led to an increased risk of overall endometrial and endometrioid cancer. While BMI adjusted for either metabolite increased the risk of non-endometrioid cancer, BF and WHR reduced the risk of non-endometrioid cancer when adjusted for triglycerides in small VLDL. The effect of WHR on non-endometrioid cancer after adjusting for triglycerides in very small VLDL, although positive, was attenuated compared to the unadjusted effect. CIs for all adjusted analyses overlapped with CIs for respective unadjusted analyses. For triglycerides in small VLDL, effect sizes were broadly consistent across adjusted and unadjusted analyses with overlapping CIs; the exception being for the effect on non-endometroid cancer, where effect sizes increased compared to the unadjusted effect size. A similar picture is apparent for triglycerides in very small VLDL. \par

#### Sensitivity analysis
Instrument strength, estimated using the F-statistic, can give an estimate of weak instrument bias in MR analyses. In general, an F-statistic of $\geq$ 10 is considered sufficiently strong that there is not substantial weak instrument bias[@Burgess2011; @Burgess2015b]. In a single sample setting, the bias is towards the observational estimate, whereas in a two-sample setting, the bias is towards the null. Weak instrument bias was estimated using a generalized version of Cochran's Q[@Sanderson2019b]. F-statistics were $\geq$ 10 for all but triglycerides in small VLDL adjusted for BMI (F = 9) and WHR (F = 7) (Table \@ref(tab:mediation-table-fstatistics)). F-statistics in MVMR analysis were lower than those associated with the instruments used in the two-sample univariable MR analysis, for example BMI in two-sample univariable MR had an F-statistic of 66 while after adjustment for both metabolites this dropped to 54. \par

\newpage

```{r mediation-table-fstatistics, eval=TRUE, include=TRUE, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
data <- read.table("../../007_metabolites_outcomes/analysis/004_mvmr/mvmr_results.txt", header = T, sep = "\t")
data <- data[!duplicated(data$f_stat.exposure1),]

metab <- read.table("../../007_metabolites_outcomes/analysis/003_metabolite_endometrial/exposure_data.txt", header = T, sep = "\t")
svldltg <- subset(metab, exposure == "S_VLDL_TG_int_imputed.txt_snps.txt")
svldltg$f_stats <- (svldltg$b / svldltg$se)^2 
svldltg$mean_fstat <- mean(svldltg$f_stats)
xsvldltg <- subset(metab, exposure == "XS_VLDL_TG_int_imputed.txt_snps.txt")
xsvldltg$f_stats <- (xsvldltg$b / xsvldltg$se)^2 
xsvldltg$mean_fstat <- mean(xsvldltg$f_stats)

adiposity <- read.table("../../007_metabolites_outcomes/analysis/001_adiposity_endometrial/exposure_data.txt", header = T, sep = "\t")
bmi <- subset(adiposity, exposure == "Locke BMI EU sex combined 77 SNPs clumped")
whr <- subset(adiposity, exposure == "Shungin WHR EU sex combined 26 SNPs")
bf <- subset(adiposity, exposure == "Lu BF EU sex combined 5 SNPs")

table <- data.frame(exposure = c("SVLDLTG", "XSVLDLTG", "BMI", "WHR", "BF"),
                    BMI = c(round(data$f_stat.exposure1[1],0), round(data$f_stat.exposure1[2],0),"","",""),
                    
                    WHR = c(round(data$f_stat.exposure1[3],0), round(data$f_stat.exposure1[4],0),"","",""),
                    
                    BF = c(round(data$f_stat.exposure1[5],0), round(data$f_stat.exposure1[6],0),"","",""),
                    
                    SVLDLTG = c("","",round(data$f_stat.exposure2[1],0),round(data$f_stat.exposure2[2],0),round(data$f_stat.exposure2[3],0)),
                    
                    XSVLDLTG = c("","",round(data$f_stat.exposure2[4],0),round(data$f_stat.exposure2[5],0),round(data$f_stat.exposure2[6],0)))

knitr::kable(table, longtable = T, booktabs = T, format = "latex", escape = F,
    caption = 'F-statistics for instruments used in univariable and multivariable Mendelian randomization analyses', row.names = F,
    col.names = c("Adjusted variable", "BMI (66)", "WHR (49)", "BF (47)", "SVLDLTG (43)", "XSVLDLTG (53)")) %>%
    add_header_above(c(" " = 1, "Exposures" = 5)) %>%
  kable_styling(latex_options = c("striped", "HOLD_position"), position = "center") %>%
  kableExtra::footnote(general = "F-statistics are presented for each exposure (column) and the adjusted (row) variable, e.g., body mass index (BMI) adjusted for triglycerieds in small very low density lipoprotein (SVLDLTG) (column 1 row 1) = 54. F-statistics for each exposure used in the univariable two-sample Mendelian rnadomization analyses are given in (). WHR = waist hip ratio; BF = body fat percentage; SVLDLTG = Triglycerides in small VLDL; XSVLDLTG = Triglycerides in very small VLDL.", threeparttable = T, general_title = "", footnote_as_chunk = T)
```

```{r eval=TRUE, include=TRUE, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
data <- read.table("../../007_metabolites_outcomes/analysis/004_mvmr/mvmr_sensitivity_results_formatted.txt", header = T, sep = "\t")
data <- data[!duplicated(data$Qstat),]
sig <- subset(data, Qpval <= 0.05)
```

Horizontal pleiotropy, whereby the effect of the instrument on the outcome is not exclusively through the exposure, was estimated using a modified form of Cochran's Q[@Sanderson2019b]. In total, `r nrow(data)` instruments (combined adiposity-associated SNP and metabolite-associated SNP) were used in MVMR analyses. Of these, `r nrow(sig)` showed evidence of horizontal pleiotropy (p-value < 0.05). That is, for a majority of tests, the association between the instrument and the outcome was not solely via the association between the instrument and the exposure and mediator. All 6 analyses investigating associations with non-endometrioid cancer, and one analysis with endometrial cancer showed weak evidence of pleiotropy (p-value > 0.05). On the whole, evidence of pleiotropy was strongest when instrumenting triglycerides in very small VLDL (Table \@ref(tab:mediation-table-qstatistics)). These results highlight the potential violation of the exclusion restriction assumption for analyses investigating adjusted effects on overall endometrial and endometrioid cancer. \par

\newpage

```{r mediation-table-qstatistics, eval=TRUE, include=TRUE, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
data <- read.table("../../007_metabolites_outcomes/analysis/004_mvmr/mvmr_sensitivity_results_formatted.txt", header = T, sep = "\t")
data <- data[!duplicated(data$Qstat),]
data <- data[,c("group", "outcome", "Qstat", "Qpval")]
data <- data[order(data$Qpval),]
data$Qstat <- round(data$Qstat, 0)
data$Qpval <- round(data$Qpval, 4)
data$group[data$group == "bf_svldltg"] <- "BF & SVLDLTG"
data$group[data$group == "bf_xsvldltg"] <- "BF & XSVLDLTG"
data$group[data$group == "bmi_svldltg"] <- "BMI & SVLDLTG"
data$group[data$group == "bmi_xsvldltg"] <- "BMI & XSVLDLTG"
data$group[data$group == "whr_svldltg"] <- "WHR & SVLDLTG"
data$group[data$group == "whr_xsvldltg"] <- "WHR & XSVLDLTG"
data <- droplevels(data)
data <- data[order(data$outcome, data$group),]
knitr::kable(data, longtable = T, booktabs = T, format = "latex", escape = T,
             caption = 'Multivariable Mendelian rnadomization Q statistics', row.names = F,
             col.names = c("Instrument", "Outcome", "Q", "P-value")) %>%
  kable_styling(latex_options = c("striped", "HOLD_position"), position = "center") %>%
  kableExtra::footnote(general = "Table is ordered by outcome and then by instrument. BMI = body mass index; WHR = waist hip ratio; BF = body fat percentage; SVLDLTG = Triglycerides in small VLDL; XSVLDLTG = Triglycerides in very small VLDL; Q = Cochran's Q.", threeparttable = T, general_title = "", footnote_as_chunk = T)
```

## Discussion
<!-- 
The majority of MR analyses identified in Chapter \@ref(systematic-review) investigated the effects of adiposity on health outcomes using sex-combined exposure and outcome data. Where the outcome was sex-specific, the majority of analyses used sex-specific exposure data. In the analyses detailed here in, exposure and outcome data were sex-combined. However, as many studies identified in Chapter \@ref(systematic-review) used sex-combined exposure data to investigate sex-specific outcomes such as ovarian[@Dixon2016a], breast[@Shu2019], and endometrial cancer[@Nead2015a; @Yarmolinsky2019], as well as cardiovascular disease[@Hagg2015], the use of sex-combined exposure data does not mean that results cannot be used to inform sex-specific estimates. \par

Of particular note is the effect of WHR adjusted for either metabolite and either metabolite adjusted for WHR on non-endometrioid cancer; WHR and either metabolite were associated with non-endometrioid cancer in the unadjusted analysis, however, after adjusting WHR for either metabolite, the effect of WHR on non-endometrioid cancer attenuated to the null while the effect of the metabolites adjusted for WHR remained consistent with the unadjusted analysis. A similar picture is seen for the effect of BF adjusted for either metabolite and either metabolite adjusted for BF on non-endometrioid cancer; the effect of BMI adjusted for either metabolite shows a smaller attenuation to the null. These results would suggest the effect of WHR (and to a lesser extent BF) on non-endometrioid cancer may be mediated by triglycerides in small and very small VLDL or there is considerable pleiotropy. A table of the full results is available on [GitHub](https://github.com/mattlee821/000_thesis/blob/master/index/data/mediation/results/004_mvmr/mvmr_results_formatted.txt). \par


-->
```{r eval=TRUE, include=TRUE, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
# data 
data <- read.table("../../007_metabolites_outcomes/analysis/001_adiposity_endometrial/001_MR_results.txt", header = T, sep = "\t")
data$group[data$exposure == "Locke BMI EU sex combined 77 SNPs clumped"] <- "BMI"
data$group[data$exposure == "Shungin WHR EU sex combined 26 SNPs"] <- "WHR"
data$group[data$exposure == "Lu BF EU sex combined 5 SNPs"] <- "BF"
data$outcome_label[data$outcome == "Endometrial cancer (endometrioid histology) || id:ebi-a-GCST006465"] <- "Endometrioid"
data$outcome_label[data$outcome == "Endometrial cancer (Non-endometrioid histology) || id:ebi-a-GCST006466"] <- "Non-endometrioid"
data$outcome_label[data$outcome == "Endometrial cancer || id:ebi-a-GCST006464"] <- "Endometrial cancer"
data$method[data$method == "Inverse variance weighted (multiplicative random effects)"] <- "IVW-MRE"

data <- subset(data, method == "IVW-MRE")
bmi <- subset(data, group == "BMI")
whr <- subset(data, group == "WHR")
bf <- subset(data, group == "BF")
```

In this chapter, the role of adiposity on endometrial cancer and the potential intermediary role of adiposity-related metabolites in this association was investigated. Firstly, results add to the growing evidence for a causal effect of BMI on endometrial cancer (OR per normalized SD unit increase in BMI = `r round(bmi$OR[1],2)`; 95% CI = `r round(bmi$lower_ci[1],2)`--`r round(bmi$upper_ci[1],2)`; p-value = 1.74 x 10^-15^); Table \@ref(tab:mediation-table-adiposity-cancer). This was consistent with previous MR results[@Gharahkhani2019; @Painter2016; @Yarmolinsky2019] presented as part of Chapter \@ref(systematic-review) (OR = 1.57; 95% CI = 1.11 – 2.22; p-value = 0.01) and observational results reported in the literature by Bhaskaran et al., (2014; OR = 1.62; 95% CI = 1.56 – 1.69; p-value = 1 x 10^-4^)[@Bhaskaran2014]. In addition, this work highlighted the causal role for BMI on sub-types of endometrial cancer (specifically, endometrioid and non-endometrioid cancer) and for both WHR and BF on these endometrial cancer outcomes. Two metabolites, triglycerides in small and in very small VLDL, for which there was evidence for a causal relationship with adiposity measures and endometrial cancer, were included in MVMR analyses. There was some evidence for an intermediate effect of triglycerides in small VLDL and triglycerides in very small VLDL on the effect of WHR on non-endometrioid cancer and BF on non-endometrioid cancer. However, whether the effect of WHR (and BF) on non-endometrioid cancer was mediated by triglycerides in small and very small VLDL was unclear; sensitivity analysis did not indicate that effects were due to horizontal pleiotropy, however weak instruments may have biased the results for these effects on non-endometrioid cancer towards the null. There was weak evidence for an intermediate effect of either metabolite on endometrioid and overall endometrial cancer across adiposity measures. There appeared some differentiation in the association between adiposity measures and endometrial cancer subtype. There was also stronger evidence for an association between both metabolites and non-endometrioid cancer than for overall endometrial and endometrioid cancer. These differences could be the result of heterogeneity and differences in the underlying biological mechanisms. different biases effecting the different subtypes, or that the GWAS (from which summary statistics were obtained) could have been influenced by issues such as differential loss to follow up, menopause status, and medication status. In regards to the latter, while it is difficult to predict their impact on MR analyses, it is possible they could act differently across cancer sub-types. \par

Previous observational work has found increased triglycerides to be associated with an increased risk of endometrial cancer[@Lindemann2009]. The effect increased in a dose response manner and persisted after adjusting for BMI and other potential confounders. Similar results have been found elsewhere[@Ulmer2009; @Trabert2015], including associations between increasing triglycerides and advancing tumour stage[@Luo2019]. Results elsewhere have also shown weak evidence for an association between triglycerides and endometrial cancer after adjusting for BMI[@Cust2007]. There is some evidence to suggest individual triglyceride metabolites, such as myristic acid (a component of the triglyceride trimyristin[@NationalCenterforBiotechnologyInformation2021]), are associated with endometrial cancer development[@Lindemann2009]. However, there are few studies focussing solely on the effects of specific triglycerides on endometrial cancer, as those discussed focussed primarily on triglycerides as a class rather than specific lipoprotein compositions. Though there is benefit to looking at metabolite classes (i.e., overall picture of metabolite effects), this does not allow for differentiation in metabolite effects[@Corbin2020], e.g., VLDL may be associated with a decreased risk of endometrial cancer, but triglycerides in small VLDL is associated with an increase in endometrial cancer risk. Additionally, it is reasonable to expect that the properties of specific metabolites are exhibited differently in different tissues and cells around the body[@VanderHeiden2009; @DeBerardinis2012; @Zhu2019; @Pavlova2016]. It is therefore hard to compare results here with those from previous analyses. Results here suggest there is an increasing effect of triglycerides on endometrial cancer risk, but that this is likely to be specific to non-endometrioid cancer, and specific to the alignment of triglycerides in small and very small VLDL, which has not been reported in the literature. Instead, associations have focussed on overall endometrial cancer risk. \par

Alongside triglycerides in small and very small VLDL, three other metabolites showed evidence of association with endometrial cancer. Increased phospholipids in very large HDL and triglycerides in very large HDL were associated with an increased risk of endometrial, endometrioid, and non-endometrioid cancer. There is some evidence for an increasing effect of phospholipids and triglycerides on endometrial cancer risk[@Raffone2020] but whether this is specific to very large HDL composition is unclear as studies have not focussed on individual metabolites. Valine was negatively associated with endometrial, endometrioid, and non-endometrioid cancer risk. Previous studies have reported increased valine to be positively associated with endometrial cancer risk[@Raffone2020]. In two-sample univariable MR analysis here, BMI and WHR were associated with an increase in valine, which has been proposed as a potential intermediate in the association between adiposity and endometrial cancer[@Kliemann2021]. However, in two-sample univariable MR analysis here, valine was associated with a decreased risk of endometrial, endometrioid, and non-endometrioid cancer. This protective effect could be due to metabolic reprogramming of cancer cells which can utilise branched-chain amino acids, such as valine, as alternative fuel sources[@Lieu2020; @Green2016]. \par

When comparing two-sample univariable MR (unadjusted) and two-sample MVMR (adjusted) effect estimates for the association of adiposity measures, there was little change in effect size and CIs overlapped for effects on overall and endometrioid cancer risk. For non-endometrioid cancer risk, the effect of all three adiposity measures, for which CIs did not cross the null in the unadjusted analysis (except for BF), crossed the null in adjusted analyses. This may be reflective of reduced power given the smaller number of cases for non-endometrioid cancer (N = 1,230) as opposed to endometrioid (N = 8,758) and overall endometrial cancer (N = 12,906). For metabolites, when comparing the unadjusted and adjusted effect estimates, effect sizes and CIs were generally larger in the adjusted analysis for all endometrial cancer outcomes. Although in some instances, MVMR can increase power[@Burgess2015b], MVMR is generally considered to be less powered than two-sample univariable MR. The observed decreased precision and larger effect sizes in adjusted analyses may be a result of weak instrument bias, especially for triglycerides in small VLDL adjusted for BMI and WHR. But it may also reflect a true causal association or a highly pleiotropic environment, especially for analyses of the effect on endometrioid and endometrial cancer, where evidence of horizontal pleiotropy was likely high. \par

Previous observational and MR work has found similar effects to those found here. In one study, observational and MR results indicated an association between BMI and endometrial cancer but not WHR and endometrial cancer[@Painter2016]. Additional studies using MR have found consistent effects[@Gharahkhani2019; @Yarmolinsky2019]. There is further observational evidence for associations between BMI and endometrial cancer, as well as WHR, waist circumference, and waist to height ratio[@Barberio2019]. Observational investigations have highlighted links between BMI and both endometrioid and non-endometrioid[@Gates2006; @Bjørge2007; @Crosbie2010; @Yang2013; @Setiawan2013; @Aune2015a; @Gao2018], but there has been a lack of genetic data on both subtypes in large, well-powered studies to enable MR investigations. Additionally, studies have predominantly focussed on BMI with some focus on measures of deposition such as WC and WHR, but as yet very little work using a more comprehensive measure of adiposity such as BF. Here, investigation of both subtypes and multiple measures of adiposity was possible. MR results were consistent with those found in the aforementioned observational studies. This consistency was similar in the strength of effects, with a stronger effect observed for endometrioid cancer compared to non-endometrioid (though CIs overlapped). Of note is the association found for WHR and non-endometrioid cancer, which has not previously been reported. \par

<!--
An interesting question that arises from this work is the relationship between adiposity measures, metabolites, and subtypes of endometrial cancer. Findings here suggested that metabolites may play an intermediary role in the association between WHR and BF and non-endometrioid cancer, but not an association with BMI, and not an association with endometrioid cancer. This raises two key question: (i) what is different about the association between BMI and non-endometrioid cancer and, the association between WHR and BF and non-endometrioid cancer and, (ii) what is different about the association between adiposity and endometrioid cancer and adiposity and non-endometrioid cancer. As discussed in Chapter \@ref(MR), endometrial cancer can broadly be categorised as hormone dependent (endometrioid cancer) and hormone-independent (non-endometrioid cancer), though this distinction is not clear cut; there is some evidence to suggest that non-endometrioid cancers may also have hormone dependent characteristics[@WILD2020]. 
-->

### Instrumentation
Instrumentation, described briefly for measures of adiposity in Chapter \@ref(MR), is primarily achieved by using either single genetic variants or multiple genetic variants as instrumental variables. The choice of instruments for measures of adiposity is generally to use the largest available GWAS and a genome-wide p-value threshold of p-value $\leq$ 5 x 10^-8^. This approach was utilised in Chapter \@ref(MR), however instruments for BMI and WHR were obtained from GWAS which included UK Biobank. UK Biobank is a unique resource with wide application, one of which is as a component of the most recent and largest metabolomics GWAS (unpublished) and endometrial cancer GWAS[@OMara2018], both used here. Overlap between the exposure and outcome in a two-sample MR can bias estimates towards the observational estimate in a manner proportional to the size of the overlap when there is weak instrument bias[@Burgess2016] (see limitations section below). As such, instruments for adiposity measures were obtained from GWAS which did not include UK Biobank. In Chapter \@ref(MR), BMI[@Locke2015; @Yengo2018] and WHR[@Shungin2015; @Pulit2019] were instrumented using the largest available GWAS (which included UK Biobank) and the second largest GWAS which was used in analyses in this chapter, where results were highly correlated. As endometrial cancer is a sex-specific condition, this work is limited by the use of sex-combined instruments for adiposity, though this does have utility in identifying the general effects of adiposity, especially for metabolites. Instrumentation practices for metabolites are less well established, but have largely followed a similar principle to that described above. Here, a simple instrumentation approach was used for metabolites (p-value $\leq$ 5 x 10^-8^ and LD R^2^ $\leq$ 0.0001 thresholds). This approach could have been strengthened by collapsing metabolites into broader categories[@Bull2020] or by using composite instruments[@Sulc2021]. \par

### Strengths and limitations
A prominent strength of this work was the breadth of metabolomics data used here and in previous chapters. Metabolites used in MVMR analyses here, were selected based on evidence from prior observational (Chapter \@ref(observational)) and MR analyses (Chapter \@ref(MR)) for causal relationships with measures of adiposity. In Chapter \@ref(observational), the metabolic profile of BMI, WHR, and BF was investigated using the Avon Longitudinal Study of Parents and Children (ALSPAC). The association between adiposity measures and numerous metabolites was robust after accounting for a number of potential confounders such as smoking status, diet, and physical activity level. In Chapter \@ref(MR), the metabolic profile of BMI, WHR, and BF was investigated in a causal framework. Effects identified in the previous observational analysis, as well as in previously published studies, was replicated in two independent studies. Finally, in this chapter, replication of the causal effects of adiposity on the metabolic profile was performed using the largest independent data set available, UK Biobank. Additionally, replication of the causal effect of BMI on endometrial cancer, as obtained from a meta-analysis in Chapter \@ref(systematic-review), was performed. \par

An additional complication here is the bias introduced through sample overlap. As discussed in the instrumentation section above, sample overlap can bias estimates towards the observational confounded estimate in a manner proportional to the overlap in the presence of weak instruments. In order to minimise overlap, and knowing that summary statistics for both the metabolites and the outcomes included individuals from UK Biobank, exposure-related instruments were sought from GWAS which did not include UK Biobank. Overlap was therefore limited to the metabolite data endometrial cancer data, specifically the overall endometrial cancer data. For binary outcomes, bias due to overlap is specific to the overlap in individuals in the exposure GWAS and those who are cases in the outcome GWAS. In this regard, the number of UK Biobank participants included in the case arm is at most 5% of the total cases. As such, the relative bias in analyses using metabolite and overall endometrial cancer data is at most 0.5%. \par

In MR, the three main assumptions (i.e., that SNPs are associated with the exposure, SNPs share no common cause with the outcome, and SNPs do not affect the outcome except via the exposure) need to be satisfied in order to obtain robust causal estimates. Adiposity instruments used here have been used widely and are robustly associated with their respective phenotypic measures[@Locke2015; @Shungin2015; @Lu2016]. For metabolites, previous work has used a standard approach of identifying SNPs reaching a genome-wide p-value threshold (p-value $\leq$ 5 x 10^-8^), which was employed here alongside a stringent LD clumping threshold (R^2^ = 0.001 and 10,000 base window). However, the strength of these instruments in two-sample MVMR was considerably weaker than in two-sample univariable MR and, for triglycerides in small VLDL, this was below the nominal threshold of 10. It is likely then that weak instrument bias[@Burgess2011] is a present feature of the MVMR results and is likely to in part explain the decreased precision in CIs. This reduced precision may for example explain why triglycerides in very small VLDL adjusted for WHR was no longer associated with non-endometroid cancer in MVMR versus two-sample univariable MR. In addition to this, instruments reflect a life-time change in the exposure and do not take into account change due to lifestyle for instance. \par

It is possible that SNPs associated with metabolites do not conform to the third IV assumption. That is, SNPs associated with the exposure have an effect on the outcome which is not directly via their association with the exposure. This is of particular concern with metabolites given they are highly inter-correlated and have a shared genetic architecture. In this sense, the effect of a metabolite on an outcome could be through a highly correlated second metabolite. Of the 56 and 50 SNPs associated with triglycerides in small and very small VLDL, respectively (90 unique SNPs), 16 were shared between them (18%). It is likely both metabolites share SNPs with many other metabolites given that just over half of the 934 identified SNPs for 53 metabolites were associated with just one metabolite and the average number of SNPs associated with a metabolite was 49. There is evidence to show that, not only are there many shared SNPs across many metabolites, but also that the variance explained for any metabolite instrument is more often than not higher for an alternative metabolite[@Guida2021]. Non-specificity of metabolite instruments will either lead to horizontal pleiotropy or vertical pleiotropy. Distinguishing one from the other without understanding the functions those SNPs have is challenging[@Swerdlow2016]. However, recently developed tools that use colocalisation methods[@Foley2021] may help to improve the specificity of instruments going forward as they are able to identify shared genetic signals across related traits. These shared genetic signals could then be removed from an instrument, or be used as an instrument for the collection of traits. \par

In two-sample univariable MR, sensitivity methods (MR-Egger, weighted median, and weighted mode) showed almost entirely consistent directions of effect for both metabolites with the IVW-MRE method, though the majority of CIs crossed the null. The exception to the consistent direction was triglycerides in small VLDL on endometrioid cancer in MR-Egger analysis which, although the CIs crossed the null, was opposite to the directions of effect of other methods. In MVMR, sensitivity methods such as these have not been developed. Instead, unbalanced horizontal pleiotropy was evaluated using a modified form of Cochran's Q[@Sanderson2019], which found evidence (p-value < 0.05) of unbalanced horizontal pleiotropy for a majority of tests and may therefore bias estimates. That is, there was evidence that the effect of the instrument on the outcome was not solely via the association between the instrument and the exposure. This could be a result of the inter-correlated nature of metabolites and their shared genetic architecture. For example, it is possible that an association between triglycerides in small VLDL and endometrial cancer is biased by the fact that triglycerides in small VLDL share a genetic architecture with triglycerides in very small VLDL, which could be the true causal pathway. The sensitivity analyses performed here only highlight the potential presence of pleiotropy, further work, for example alternative instrumentation, is required to understand the potential pleiotropic pathways. \par

A key limitation of this work is the use of non-sex-specific instruments for measures of adiposity. A major driver of endometrial cancer, specifically endometrioid cancer, is oestrogen[@WILD2020], which, as a sex hormone, is important in endometrial cell growth. Oestrogen is also involved in the distribution and accumulation of adipose tissue[@Wells2007]. Importantly, adipose tissue is a key source of oestrogen synthesis, particularly during post-menopause[@Simpson2003]. There is sexual dimorphism in the accumulation and distribution of adipose tissue that has been highlighted in genetic studies, which found sex-specific SNPs associated with adiposity[@Locke2015; @Shungin2015; @Pulit2019; @Yengo2018; @Hubel2019]. There is also evidence of sex differences in associations between adiposity and metabolites[@OKeeffe2020]. The use of non-sex-specific adiposity instruments may therefore have biased estimates. This is because, in a two-sample MR, the assumption is that the underlying population of the two samples is the same. Violation of this assumption and therefore of the second IV assumption (that instruments do not share a common cause with the outcome) can lead to biased estimates for example from genetic confounding and population structure. \par

In this chapter, NMR-derived metabolites were used, as with work in chapters \@ref(observational) and \@ref(MR). Although a relatively large number of metabolites were investigated, they are predominantly lipid-based. This leaves a broad array of metabolites that have not been investigated. Mass spectrometry (MS) platforms, such as those used by Metabolon, which measure many hundreds of metabolites[@Alseekh2021], are a potential next step. Though there have been a number of studies investigating the association between adiposity and MS-derived metabolites[@Moore2018; @Cirulli2019; @Perng2020], these have not used a complimentary assessment of body composition to assess adiposity and have not been able to investigate this relationship across the life-course. For example, a study by Ho et al. (2016)[@Ho2016] focussed solely on the metabolic profile of adiposity in 2,383 adults and found 69 out of 217 metabolites, including triglycerides and a number of other metabolites (e.g., amino acids) identified here, to be associated with an increased BMI. A number of studies have investigated associations between MS-derived metabolites and endometrial cancer. However, these studies have included fewer than 100 cases and have focussed on women who are post-menopausal[@Audet-Delage2018] or have obesity[@Njoku2021] only. In the latter case, phospholipids were identified as a biomarker of significance. A study by Audet-Delage et al. (2018)[@Audet-Delage2018], which focussed on women who are post-menopausal, identified lipid pathways to be most affected by case status. Although numerous SNPs have been associated with MS-derived metabolites[@Shin2014; @Cheng2021; @Lotta2021], there have been no studies that have employed MR to investigate metabolic associations with adiposity measures or endometrial cancer. \par

### Conclusion
There were few metabolites that showed evidence of a causal relationship with adiposity measures and increased the risk of endometrial cancer. Of the few metabolites with evidence for a causal relationship with adiposity and endometrial cancer, there was evidence that triglycerides in small and very small VLDL may play an intermediary role in the effect of both WHR and BF on non-endometrioid cancer. There was some suggestive evidence for a similar intermediary role between BMI and BF and non-endometrioid cancer. Non-endometrioid cancer is the more aggressive of the two common subtypes of endometrial cancer, and is typically considered oestrogen independent but recent work has shown this is not necessarily the case. There was weak evidence of horizontal pleiotropy in MVMR analyses investigating effects on non-endometrioid cancer compared to endometrioid and overall endometrial cancer. However, this does not rule out the possibility of pleiotropic effects, especially as metabolites are highly intercorrelated and have a shared genetic architecture. Future work to elucidate the intermediary relationship of triglycerides in small and very small VLDL should have three focusses: (i) specific metabolites within the VLDL triglyceride subclass, (ii) comprehensive consideration of instrumentation, and (iii) wet lab studies, which, when informed by MR analyses, have been fruitful in identifying molecular mechanisms[@Nounu2021]. \par

\newpage





<!--
```{r mediation-table-N-SNP, eval=TRUE, include=TRUE, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
# adiposity SNPs
adiposity <- read.table("../../007_metabolites_outcomes/analysis/001_adiposity_endometrial/exposure_data.txt", header = T, sep = "\t")
bmi <- subset(adiposity, exposure == "Locke BMI EU sex combined 77 SNPs clumped")
whr <- subset(adiposity, exposure == "Shungin WHR EU sex combined 26 SNPs")
bf <- subset(adiposity, exposure == "Lu BF EU sex combined 5 SNPs")

# metabolite SNPs
metab <- read.table("../../007_metabolites_outcomes/analysis/003_metabolite_endometrial/exposure_data.txt", header = T, sep = "\t")
svldltg <- subset(metab, exposure == "S_VLDL_TG_int_imputed.txt_snps.txt")
svldltg$f_stats <- (svldltg$b / svldltg$se)^2 
svldltg$mean_fstat <- mean(svldltg$f_stats)
xsvldltg <- subset(metab, exposure == "XS_VLDL_TG_int_imputed.txt_snps.txt")
xsvldltg$f_stats <- (xsvldltg$b / xsvldltg$se)^2 
xsvldltg$mean_fstat <- mean(xsvldltg$f_stats)

# mvmr adiposity SNPs
mvmr_bmi_svldltg <- read.table("../../007_metabolites_outcomes/analysis/004_mvmr/adiposity/bmi_svldltg_mvmr_instruments.txt", header = T, sep = "\t")
mvmr_bmi_xsvldltg <- read.table("../../007_metabolites_outcomes/analysis/004_mvmr/adiposity/bmi_xsvldltg_mvmr_instruments.txt", header = T, sep = "\t")
mvmr_whr_svldltg <- read.table("../../007_metabolites_outcomes/analysis/004_mvmr/adiposity/whr_svldltg_mvmr_instruments.txt", header = T, sep = "\t")
mvmr_whr_xsvldltg <- read.table("../../007_metabolites_outcomes/analysis/004_mvmr/adiposity/whr_xsvldltg_mvmr_instruments.txt", header = T, sep = "\t")
mvmr_bf_svldltg <- read.table("../../007_metabolites_outcomes/analysis/004_mvmr/adiposity/bf_svldltg_mvmr_instruments.txt", header = T, sep = "\t")
mvmr_bf_xsvldltg <- read.table("../../007_metabolites_outcomes/analysis/004_mvmr/adiposity/bf_xsvldltg_mvmr_instruments.txt", header = T, sep = "\t")

# mvmr metabolite SNPs
mvmr_svldltg_bmi <- read.table("../../007_metabolites_outcomes/analysis/004_mvmr/metabolites/bmi_svldltg_mvmr_instruments.txt", header = T, sep = "\t")
mvmr_svldltg_whr <- read.table("../../007_metabolites_outcomes/analysis/004_mvmr/metabolites/whr_svldltg_mvmr_instruments.txt", header = T, sep = "\t")
mvmr_svldltg_bf <- read.table("../../007_metabolites_outcomes/analysis/004_mvmr/metabolites/bf_svldltg_mvmr_instruments.txt", header = T, sep = "\t")
mvmr_xsvldltg_bmi <- read.table("../../007_metabolites_outcomes/analysis/004_mvmr/metabolites/bmi_xsvldltg_mvmr_instruments.txt", header = T, sep = "\t")
mvmr_xsvldltg_whr <- read.table("../../007_metabolites_outcomes/analysis/004_mvmr/metabolites/whr_xsvldltg_mvmr_instruments.txt", header = T, sep = "\t")
mvmr_xsvldltg_bf <- read.table("../../007_metabolites_outcomes/analysis/004_mvmr/metabolites/bf_xsvldltg_mvmr_instruments.txt", header = T, sep = "\t")

# table 
table <- data.frame(a = "N",
                    BMI = nrow(bmi),
                    WHR = nrow(whr),
                    BF = nrow(bf),
                    SVLDLTG = nrow(svldltg),
                    XSVLDLTG = nrow(xsvldltg))

table1 <- data.frame(a = "N",
                    `BMI SVLDLTG` = nrow(mvmr_bmi_svldltg),
                    `BMI XSVLDLTG` = nrow(mvmr_bmi_xsvldltg),
                    `WHR SVLDLTG` = nrow(mvmr_whr_svldltg),
                    `WHR XSVLDLTG` = nrow(mvmr_whr_xsvldltg),
                    `BF SVLDLTG` = nrow(mvmr_bf_svldltg),
                    `BF XSVLDLTG` = nrow(mvmr_bf_xsvldltg),
                    
                    `SVLDLTG BMI` = nrow(mvmr_svldltg_bmi),
                    `SVLDLTG WHR` = nrow(mvmr_svldltg_whr),
                    `SVLDLTG BF` = nrow(mvmr_svldltg_bf),
                    `XSVLDLTG BMI` = nrow(mvmr_xsvldltg_bmi),
                    `XSVLDLTG WHR` = nrow(mvmr_xsvldltg_whr),
                    `XSVLDLTG BF` = nrow(mvmr_xsvldltg_bf))

knitr::kable(table, longtable = T, booktabs = T, format = "latex", escape = F,
    caption = 'Number of SNPs used in the two-sample MR analyses and multivaraibel MR analyses', row.names = F,
    col.names = c("", "BMI", "WHR", "BF", "SVLDLTG", "XSVLDLTG")) %>%
  kable_styling(latex_options = c("striped", "HOLD_position"), position = "center") %>%
  kableExtra::footnote(general = " BMI = body mass idnex; WHR = waist hip ratio; BF = body fat percentage; SVLDLTG = Triglycerides in small VLDL; XSVLDLTG = Triglycerides in very small VLDL.", threeparttable = T, general_title = "", footnote_as_chunk = T)

knitr::kable(table1, longtable = T, booktabs = T, format = "latex", escape = F,
    caption = 'F-statistics for adjusted exposures used in multivariable MR analyses', row.names = F,
    col.names = c("", "SVLDLTG", "XSVLDLTG", "SVLDLTG", "XSVLDLTG", "SVLDLTG", "XSVLDLTG", "BMI","WHR","BF","BMI","WHR","BF")) %>%
  add_header_above(c("", "BMI" = 2, "WHR" = 2, "BF" = 2, "SVLDLTG" = 3, "XSVLDLTG" = 3)) %>%
  kable_styling(latex_options = c("striped", "HOLD_position"), position = "center") %>%
  kableExtra::footnote(general = "BMI = body mass idnex; WHR = waist hip ratio; BF = body fat percentage; SVLDLTG = Triglycerides in small VLDL; XSVLDLTG = Triglycerides in very small VLDL.", threeparttable = T, general_title = "", footnote_as_chunk = T)


                    

```
-->

